# Political Affiliation

This chapter explores time trends in the political affiliation of the US population.

**Input**

-   DemographyGunOwners/data/clean/gs-gen-pop.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "gtsummary",
    "survey"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

##  Load data

```{r}
gun <-
  read_rds(
    here(
      path_od,
      "data",
      "clean",
      "gs-gun-owners.rds"
    )
  )
```

## Redefine political affiliation

```{r}
gun <- 
  gun %>% 
  mutate(
    altparty = 
      case_when(
        grepl(
          "Strong Democrat", 
          party7, 
          ignore.case = T
          ) ~ "Democrat",
        grepl(
          "Strong Republican", 
          party7, 
          ignore.case = T
          ) ~ "Republican",
        grepl(
          "Unknown", 
          party7
        ) ~ "Unknown",
        year == 2015 ~ "Question not asked",
        TRUE ~ "Independent"
      ) %>% 
      factor()
  )  %>% 
  filter(party3 != "Unknown") %>% 
  mutate(
    across(
      c(starts_with("party"),  "altparty"), ~ droplevels(.x)
    )
  ) %>% 
  set_variable_labels(
    party3 = "Leaners as Partisans",
    party5 = "Strong and not so Strong as Partisans",
    altparty = "Leaners as Independents",
    party7 = "Seperate Categories"
  )
```

## Demographic variables

```{r}
demovars <-
  c(
    "party3",
    "altparty",
    "party5",
    "party7"
  )
```

## Tabulate values

```{r}
walk(demovars, ~ print(tabyl(gun, .x)))
```


## Yearly Summary

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = gun,
    strata = ~ year,
    weights = ~ weight
    ) 
```

When you use the general population sample, numbers are slightly off what is in Barry et al. 2018 because they remove unknowns from their proportions. Maybe I should remove unknowns.

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  # add_ci() %>% 
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      
      tbl <- .x
      
      # compute unweighted yearly totals
      counts <- format(table(gun$year), big.mark = ",")
      
      # create row with obs
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(
          as.list(counts),
          paste0("stat_", seq_along(counts))
          )
        ) %>% 
        mutate(across(starts_with("stat"), trimws)
          )

      bind_rows(n_row, tbl) %>%
        filter(
          ! grepl("Ques", label),
          ) %>% 
        mutate(
           label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label),
            label),
          stat_2 = ifelse(
              row_type == "level",
               "-", stat_2
               )
           )
    }
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  bold_labels() %>% 
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = TRUE,
    label = "PoliticalComparison",
    linesep = "",
    caption = 
      "Political Affiliation of US Gun Owners by Survey Year and Definition, 2013-2025",
    align = "lccccccc"
  ) %>%
  kable_styling(
    latex_options =
      c("hold_position", "scale_down"),
   font_size = 9
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(
    2:8, 
    width = "1.3cm"
    ) %>% 
  row_spec(1, bold = TRUE) 

save_kable(
  year_tbl,
  file = here(path_ol, "tables/political-summary-year.tex")
)
```
