# Gender and Racial composition

This chapter focuses on exploring the gender and racial composition of US gun owners
using figures.


**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`
```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "paletteer",
    "fastDummies"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  ) %>% 
  mutate(
    educ = 
      case_when(
        grepl("Less", educ5) ~ "Less than high school",
        grepl("HS", educ5) ~ "High school graduate",
        grepl("Some|Voc", educ5) ~ "Some college, or associates degree",
        grepl("Post", educ5) ~ "Graduate study",
        TRUE ~ educ5
      ) %>% 
      factor(
        levels = 
          c(
            "Less than high school",
            "High school graduate",
            "Some college, or associates degree",
            "Bachelors degree",
            "Graduate study"
          )
      )
  )
```


## Race by gender

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

des <- update(
  des,
  rg = interaction(racethnicity, gender, sep = " x ")
)

rg_ci <- 
  svyby(
    ~rg,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))


long <-
  rg_ci %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    gender = 
      case_when(
        str_detect(group, "female") ~ 2,
        str_detect(group, "male") ~ 1,
        TRUE ~ 3) %>% 
      factor(
        levels = 1:3, 
        labels = c("Male", "Female", "Other")
      ),
    race = case_when(
      str_detect(group, "white") ~ 1,
      str_detect(group, "black") ~ 2,
      str_detect(group, "asian") ~ 3,
      str_detect(group, "other") ~ 4,
      str_detect(group, "races") ~ 5,
      str_detect(group, "rg_hispanic") ~ 6
    ) %>% 
      factor(
        levels = 1:6,
        labels = 
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Asian, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
            )
      ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  ) %>% 
  filter(!is.na(pct)) 
```


## Percentage of U.S. Gun Owners that are Women of Color

```{r}
cap <-
  paste0(
    "This figure plots non-White female-identifying U.S. gun-owner adults as a ",
    "percent of all U.S. gun-owners (N=",
     nrow(gun),") from seven waves of the National Survey of Gun Policy, ",
    "spanning 2013 to 2025.",
    " For reference, during the same period, ",
    "White non-Hispanic women composed roughly ",
    mean(filter(long, group == "rg_white_non_hispanic_x_female")$pct) %>% 
      round(0),
    " percent of U.S. gun owners, and White non-Hispanic women composed roughly ",
    mean(filter(long, group == "rg_white_non_hispanic_x_male")$pct) %>% 
      round(0),
    " percent of U.S. gun owners."
  )
```

```{r}
woc <-
  long %>% 
  filter(gender != "Other" & gender == "Female" & race != "White, non-Hispanic" ) %>% 
  mutate(
    race = 
      race %>% 
      factor(
        levels = 
         
      )
  )
```

```{r}
woc %>%  
  mutate(year = factor(year)) %>%  
  ggplot(
    aes(x = year, y = pct, group = race, color = race)
  ) +
  geom_line(linewidth = 1.25) +
  theme +
  theme(
    legend.position = c(0.2, .7),
     legend.background = element_rect(
      fill = "white",
      color = "black"
    ),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10)
  ) +
  scale_color_paletteer_d("nationalparkcolors::DeathValley") +
  scale_y_continuous(
    breaks = seq(0, 8, 2),
    limits = c(0, 8)
    ) +
  labs(
    x = "Survey Year",
    y = "Percent of U.S. gun owners",
    color = str_wrap("Racial ethnic group", 20)
  )

ggsave(
  height = 6,
  width = 8, 
  here(
    path_ol, 
    "figures/non-white-female.jpeg"
  )
)
```

### Table with confidence intervals

```{r}
tab <- 
  long %>% 
  filter(gender == "Female") %>% 
  mutate(
      across(pct:ci_u, ~ round(.x, 1) %>% as.character())  
    ) %>% 
  transmute(
    Year = factor(year),
    race = 
      race %>% 
      factor(
      labels = 
        c("White, non-Hispanic",
          "Black, non-Hispanic",
          "Hispanic",
          "Asian, non-Hispanic",
          "2+ Races, non-Hispanic",
          "Other, non-Hispanic"
          )
      ),
      pct = paste0(
        pct, " (", ci_l, ", ", ci_u, ")"
      )
    ) %>% 
  arrange(race) %>% 
  pivot_wider(names_from = "race", values_from = "pct")
```


```{r}
note <-
  paste0(
    "Note: This table shows female-identifying U.S. gun-owner adults stratified by ",
    "racial ethnic group as a ",
    "percent of all U.S. gun-owners (N=",
     nrow(gun),") from seven waves of the National Survey of Gun Policy, ",
    "spanning 2013 to 2025. Lower bounds of confidence intervals were truncated ",
    "at 0."
  )
```

```{r}
out <-
  tab %>% 
    kable(
      format = "latex",
      booktabs = TRUE,
      align = "lcccccc",
      caption = "Female-identifying U.S. gun-owner adults stratified by racial ethnic group  as percent (95 \\% CI) of all U.S. gun-owners, 2013-2025",
      label = "FemaleRace",
      escape = FALSE
    ) %>%
    kable_styling(
      latex_options = c("hold_position", "scale_down"),
      position = "center"
    ) %>% 
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, bold = TRUE) %>% 
  footnote(
    general = note,
    general_title = "",
    threeparttable = TRUE,
    escape = FALSE
    )
```

```{r}
save_kable(
  out,
  file = 
    here(
      path_ol,
      "tables/female-race.tex"
    )
)
```

## Educational status among black women

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

des <- update(
  des,
  rge = interaction(racethnicity, gender, educ, sep = " x ")
)

rge_ci <- 
  svyby(
    ~rge,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))


long <-
  rge_ci %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    gender = 
      case_when(
        str_detect(group, "female") ~ 2,
        str_detect(group, "male") ~ 1,
        TRUE ~ 3) %>% 
      factor(
        levels = 1:3, 
        labels = c("Male", "Female", "Other")
      ),
    race = case_when(
      str_detect(group, "white") ~ 1,
      str_detect(group, "black") ~ 2,
      str_detect(group, "asian") ~ 3,
      str_detect(group, "other") ~ 4,
      str_detect(group, "races") ~ 5,
      str_detect(group, "rge_hispanic") ~ 6
    ) %>% 
      factor(
        levels = 1:6,
        labels = 
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Asian, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
            )
      ),
    educ = 
      case_when(
        grepl("less", group) ~ 1,
        grepl("high_sch", group) ~ 2,
        grepl("som", group) ~ 3,
        grepl("post", group) ~ 5,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:5, 
        labels = 
          c(
            "Less than high school",
            "High school graduate",
            "Some college, or associates degree",
            "Bachelors degree",
            "Graduate study"
          )
      ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  ) %>% 
  filter(!is.na(pct)) 
```



## Look at black women

Calculate the percentage amoung black women

```{r}
bw <-
  long %>% 
  filter(gender == "Female" & grepl("Black", race))  %>% 
  group_by(year) %>% 
  mutate(tot = sum(pct)) %>% 
  ungroup() %>% 
  transmute(
    year,
    gender,
    race,
    educ,
    prop = (pct/tot)*100
    ) 
```

## Plot education level of black women gun owners

```{r}
label_data <- bw %>%
  filter(year == max(year)) %>%
  mutate(
    label = if_else(
      educ == "Some college, or associates degree",
      stringr::str_wrap(educ, width = 20),
      educ
    )
  )
```


```{r}
# ggplot(
#     bw ,
#     aes(
#       x = year,
#       y = prop,
#       group = educ, 
#       color = educ,
#       fill = educ
#       )
#     ) +
#     geom_col() +
#     geom_text(
#         data = label_data,
#         aes(
#             x = year,
#             y = prop,
#             label = label,
#             color = educ
#         ),
#         family = "Times New Roman",
#         fontface = "bold",
#         size = 3.5,
#         hjust = -0.05
#     ) +
#     scale_x_continuous(
#         breaks = unique(plot_data$year),
#         limits = c(min(plot_data$year), max(plot_data$year) + 1.2)
#     ) +
#     scale_y_continuous(
#         breaks = seq(0, 50, 5),
#         limits = c(0, 50)
#     ) +
#     scale_color_paletteer_d("MoMAColors::Althoff") +
#     theme +
#     theme(
#         legend.position = "none",
#         text = element_text(
#             family = "Times New Roman",
#             size = 12
#         ),
#         axis.title = element_text(size = 12),
#         axis.text = element_text(size = 12),
#         plot.margin = margin(10, 90, 10, 10),
#         panel.border = element_blank(),
#         panel.grid.major = element_line(color = "grey95"),
#         panel.grid.minor = element_line(color = "grey100") 
#     ) +
#     labs(
#         x = "Survey Year",
#         y = "Percent of Black, non-Hispanic Gun Owners"
#         ) + 
#   coord_cartesian(clip = "off") 
# 
# 
# ggsave(
#   height = 6,
#   width = 8, 
#   here(
#     path_ol, 
#     "figures/bw-educ.jpeg"
#   )
# )
```
