# Gender and Racial composition

This chapter focuses on exploring the gender and racial composition of US gun owners using figures.


**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`
```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "ggh4x",
    "vtable",
    "paletteer",
    "fastDummies"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  ) 
```


## Race by gender

```{r}
des <- svydesign(
  ids = ~1,       
  weights = ~weight,
  data = gun
)

des <- update(
  des,
  rg = interaction(racethnicity, gender, sep = " x ")
)

rg_ci <- 
  svyby(
    ~rg,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))
```

```{r}
long <-
  rg_ci %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    gender = 
      case_when(
        str_detect(group, "female") ~ 2,
        str_detect(group, "male") ~ 1,
        TRUE ~ 3) %>% 
      factor(
        levels = 1:3, 
        labels = c("Male", "Female", "Other")
      ),
    race = case_when(
      str_detect(group, "rg_hispanic") ~ 6,
      str_detect(group, "white") ~ 1,
      str_detect(group, "black") ~ 2,
      str_detect(group, "asian") ~ 3,
      str_detect(group, "other") ~ 4,
      str_detect(group, "races") ~ 5
    ) %>% 
      factor(
        levels = 1:6,
        labels = 
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Asian, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
            )
      ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l),
    # indicator for people of color
    poc = race !="White, non-Hispanic",
    year = factor(year)
  ) 
```

```{r}
long %>% tabyl(poc)
```

## Percentage of U.S. Gun Owners that are Women of Color


```{r}
genlist <- c("female", "male")
genrace <-
  map(
   genlist, ~ 
     filter(
       long, 
       poc & gender == str_to_title(.x)
       )
  )
```

## Define function to plot by gender

```{r}
p <-
  map(
    genrace, 
    ~ {
      
      .x %>%  
        ggplot(
          aes(
            x = year,
            y = pct,
            group = race,
            color = race
            )
        ) +
        geom_line(linewidth = 1.25) +
        theme +
        theme(
          legend.position = c(0.2, .7),
           legend.background = element_rect(
            fill = "white",
            color = "black"
          ),
          legend.text = element_text(size = 8),
          legend.title = element_text(size = 10)
        ) +
        scale_color_paletteer_d("nationalparkcolors::DeathValley") +
        labs(
          x = "Survey Year",
          y = "Percent of U.S. gun owners",
          color = str_wrap("Racial ethnic group", 20)
        )
      }
    )
```

```{r}
walk(
  seq_along(genlist), ~{
    
    p <- p[[.x]]
    
    if (.x == 1){
      
      p <-
        p +
        scale_y_continuous(
          breaks = seq(0, 8, 2),
          limits = c(0, 8)
          )
    }
    
    else (
      p <-
        p +
        scale_y_continuous(
          breaks = seq(0, 14, 2),
          limits = c(0, 14)
          ) + 
        theme(legend.position = c(0.2, .75))
    )
  
  print(p)
  
  ggsave(
    height = 6,
    width = 8, 
    here(
      path_ol, 
      "figures",
      paste0("non-white-", genlist[[.x]], ".jpeg")
      )
    )
  }
)
```


## Define caption

```{r}
capw <-
  paste0(
    "This figure plots non-White Female-identifying U.S. gun-owner adults as a ",
    "percent of all U.S. gun-owner adults (N=",
     nrow(filter(gun, !grepl("White", racethnicity) & gender == "Female")),
     ") from seven waves of the National Survey of Gun Policy, ",
    "spanning 2013 to 2025.",
    " For reference, during the same period, ",
    "White, non-Hispanic Female-identifying U.S. gun-owner adults ", 
    "composed roughly ",
    mean(filter(long, group == "rg_white_non_hispanic_x_female")$pct) %>% 
      round(0),
    " percent of U.S. gun owner adults. The survey weights in 2017 and 2019 ",
    "are general population weights and are not scaled to the gun-owner population."
  )

capm <-
  paste0(
    "This figure plots non-White Male-identifying U.S. gun-owner adults as a ",
    "percent of all U.S. gun-owner adults (N=",
     nrow(filter(gun, !grepl("White", racethnicity) & gender == "Male")),
     ") from seven waves of the National Survey of Gun Policy, ",
    "spanning 2013 to 2025.",
    " For reference, during the same period, ",
    "White, non-Hispanic Male-identifying U.S. gun-owner adults ", 
    "composed roughly ",
    mean(filter(long, group == "rg_white_non_hispanic_x_male")$pct) %>% 
      round(0),
    " percent of U.S. gun owner adults. The survey weights in 2017 and 2019 ",
    "are general population weights and are not scaled to the gun-owner population."
  )

```


### Table with confidence intervals

```{r}
notes <-
  map(
    genlist %>% str_to_title,
    ~  paste0(
    "Note: This table shows ", .x ," U.S. gun-owner adults stratified by ",
      "racial ethnic group as a ",
      "percent of all U.S. gun-owner adults (N=",
       nrow(filter(gun, gender ==.x)),
      ") from seven waves of the National Survey of Gun Policy, ",
      "spanning 2013 to 2025. Lower bounds of confidence intervals were truncated ",
      "at 0. The survey weights in 2017 and 2019 are general population weights ",
      "and are not scaled to the gun-owner population."
    )
  )

cap <-
  map(
    genlist %>% str_to_title,
    ~ paste0(
      .x,
      "-identifying U.S. gun-owner adults stratified by racial ethnic group  ",
      "as percent (95\\% CI) of all U.S. gun-owners, 2013-2025"
    )
  )
```

```{r}
out <-
  map(
    seq_along(genlist),
    function(x) {
      
      tab <- 
      filter(long, gender == str_to_title(genlist[[x]])) %>% 
        mutate(
            across(pct:ci_u, ~ round(.x, 1) %>% as.character())  
          ) %>% 
        transmute(
          Year = factor(year),
          race = 
            race %>% 
            factor(
            levels = 
              c("White, non-Hispanic",
                "Black, non-Hispanic",
                "Hispanic",
                "Asian, non-Hispanic",
                "2+ Races, non-Hispanic",
                "Other, non-Hispanic"
                )
            ),
            pct = paste0(
              pct, " (", ci_l, ", ", ci_u, ")"
            )
          ) %>% 
        arrange(race) %>% 
        pivot_wider(names_from = "race", values_from = "pct") 
      
    t <-
      tab %>% 
        kable(
          format = "latex",
          booktabs = TRUE,
          align = "lcccccc",
          caption = cap[[x]],
          label =  paste0(str_to_title(genlist[[x]]), "Race"), 
          escape = FALSE
        ) %>%
        kable_styling(
          latex_options = c("hold_position", "scale_down"),
          position = "center"
        ) %>% 
      row_spec(0, bold = TRUE) %>% 
      column_spec(1, bold = TRUE) %>% 
      footnote(
        general = str_wrap(notes[[x]], 100),
        general_title = "",
        threeparttable = TRUE,
        escape = FALSE
        )
    
    save_kable(
      t,
      file = 
        here(
          path_ol,
          "tables",
          paste0(genlist[[x]], "-race.tex")
        )
    )
    }
  )
  
```

## Age

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

des <- update(
  des,
  int = interaction(racethnicity, gender, age4, sep = " x ")
)

ci <- 
  svyby(
    ~int,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))
```

```{r}
long <-
  ci %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    gender = 
      case_when(
        str_detect(group, "female") ~ 2,
        str_detect(group, "male") ~ 1,
        TRUE ~ 3) %>% 
      factor(
        levels = 1:3, 
        labels = c("Male", "Female", "Other")
      ),
    race = case_when(
      str_detect(group, "white") ~ 1,
      str_detect(group, "black") ~ 2,
      str_detect(group, "asian") ~ 3,
      str_detect(group, "other") ~ 4,
      str_detect(group, "races") ~ 5,
      str_detect(group, "int_hispanic") ~ 6
    ) %>% 
      factor(
        levels = 1:6,
        labels = 
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Asian, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
            )
      ),
    age = 
      case_when(
        grepl("18_29", group) ~ 1,
        grepl("30_44", group) ~ 2,
        grepl("45_59", group) ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c(
            "18-29",
            "30-44",
            "45-59",
            "60+"
          )
      ),
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  )
```

## Figure among black men and women

Calculate the percentage amog black women

```{r}
raceage <-
  long %>% 
  group_by(year, gender, race) %>% 
  mutate(tot = sum(pct)) %>% 
  ungroup() %>% 
  transmute(
    year = factor(year),
    gender,
    race,
    age,
    prop = (pct/tot)*100
    ) 
```

## Age x race by year

```{r}
raceage %>% 
  filter(gender != "Other" & !grepl("Other|Asian", race))  %>% 
  ggplot(
    aes(
      x = year,
      y = prop,
      group = age,
      color = age,
      fill = age
      )
    ) +
  geom_col() +
  facet_grid2(
    rows = vars(race), 
    cols = vars(gender),
    axes = "all"
    ) +
  scale_color_paletteer_d("MoMAColors::Althoff") +
  scale_fill_paletteer_d("MoMAColors::Althoff") +
  guides(color = "none") +
  theme +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5),
     strip.background = element_rect(
      fill = "gray95",
      color = "gray90"
    )
    ) +
  labs(
    x = "Survey Year",
    y = NULL,
    fill = "Age (years)"
    )


ggsave(
  height = 8,
  width = 8,
  here(
    path_ol,
    "figures/race-age.jpeg"
  )
)
```


## Age x black women

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

des <- update(
  des,
  int = interaction(racethnicity, gender, age4, sep = " x ")
)

ci <- 
  svyby(
    ~int,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))


long <-
  ci %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    gender = 
      case_when(
        str_detect(group, "female") ~ 2,
        str_detect(group, "male") ~ 1,
        TRUE ~ 3) %>% 
      factor(
        levels = 1:3, 
        labels = c("Male", "Female", "Other")
      ),
    race = case_when(
      str_detect(group, "white") ~ 1,
      str_detect(group, "black") ~ 2,
      str_detect(group, "asian") ~ 3,
      str_detect(group, "other") ~ 4,
      str_detect(group, "races") ~ 5,
      str_detect(group, "int_hispanic") ~ 6
    ) %>% 
      factor(
        levels = 1:6,
        labels = 
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Asian, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
            )
      ),
    age = 
      case_when(
        grepl("18_29", group) ~ 1,
        grepl("30_44", group) ~ 2,
        grepl("45_59", group) ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c(
            "18-29",
            "30-44",
            "45-59",
            "60+"
          )
      ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  )
```

### Restrict at black women

Calculate the percentage among black women

```{r}
bw <-
  long %>% 
  filter(gender == "Female" & grepl("Black", race))  %>% 
  group_by(year) %>% 
  mutate(tot = sum(pct)) %>% 
  ungroup() %>% 
  transmute(
    year = factor(year),
    gender,
    race,
    age,
    prop = (pct/tot)*100
    ) 
```

### Plot age of black women gun owners

Weights in 2017 and 2019 are not gun owners specific and lead to erroneous inference.

```{r}
bw %>% 
  filter(!(year %in% as.character(2017:2019))) %>% 
  ggplot(
    aes(
      x = year,
      y = prop,
      group = age,
      color = age,
      fill = age
      )
    ) +
    geom_col() +
    scale_color_paletteer_d("MoMAColors::Althoff") +
    scale_fill_paletteer_d("MoMAColors::Althoff") +
    guides(color = "none") +
    theme +
    theme(
      legend.position = "top",
      legend.direction = "horizontal",
      legend.background = element_blank(),
      legend.key = element_blank(),
      legend.text = element_text(size = 8),
      legend.title = element_text(size = 10, hjust = .5)
      ) +
    labs(
      x = "Survey Year",
      y = "Percent of Black non-Hispanic Female Gun Owners",
      fill = "Age (years)"
      )


ggsave(
  height = 8,
  width = 8,
  here(
    path_ol,
    "figures/bw-age.jpeg"
  )
)
```
