# Type of Gun

This chapter focuses time trends in the type of gun owned


**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`
```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "paletteer",
    "fastDummies"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```


## Estimates with CIs

```{r}
walk(
  select(gun, starts_with("own")) %>% names(),
  ~  print(gun %>% tabyl(.x))
)
```

## Check years of available data

```{r}
type <-
  gun %>% 
  filter(
   if_any(starts_with("own"),
          ~ .x != "Question not asked"
          )
  )
```

```{r}
type %>% tabyl(year) %>% adorn_percentages()
```

## Arrange data

```{r}
clean <- 
  type %>% 
  select(
    year,
    starts_with("own"),
    state,
    gender, 
    educ5,
    income4,
    party3,
    employ,
    weight
  )
```




```{r}
long <- 
  clean %>% 
  pivot_longer(
    cols = starts_with("own"),
    names_to = "guntype",
    values_to = "owner",
    names_prefix = "own"
  ) %>% 
  mutate(
    owner = recode_values(
      owner,
      "Yes" ~ 1,
      "No" ~ 0,
      "Unknown" ~ 0
    ),
    year = as.factor(year),
    guntype = 
      str_to_title(guntype) %>% 
      ifelse(. == "Shot", "Shotgun", .) %>% 
      factor(
        levels = 
          c("Handgun",
            "Rifle",
            "Shotgun",
            "Other")
      )
  ) 
```

## Percentage by gun type

```{r}
main <-
  long %>% 
  group_by(guntype, year) %>% 
  summarise(
    pct = sum(owner*weight)/sum(weight),
    .groups = "drop"
    ) 
```

```{r}
des <- svydesign(
  ids = ~1,       
  weights = ~weight,
  data = clean
)

ci <- 
  svyby(
    ~rg,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))
```


### Bar chart over time

```{r}
main %>% 
  ggplot(
    aes(
      x = year,
      y = pct,
      group = guntype,
      fill = guntype,
      color = guntype
      )
  ) +
  geom_col(position = "dodge") +
  scale_fill_paletteer_d("MoMAColors::Althoff") +
  scale_color_paletteer_d("MoMAColors::Althoff") +
  theme  +
  guides(
    color = "none"
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5)
    ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, .9),
    breaks = seq(0, .9, .15)
  ) +
  labs(
    x = "Survey Year",
    y = "Percent of U.S. Gun Owners",
    fill = "Type of firearm"
    )


ggsave(
  height = 8,
  width = 8,
  here(
    path_ol,
    "figures/type.jpeg"
  )
)
```

### Caption

```{r}
note <-
  paste0(
    "This figure plots the percentage of",
    " U.S. gun owners by self-reported",
    " ownership of the following types of",
    " firearms: handgun, shotgun, rifle, and other.",
    " The data are from the 2013, 2023 and 2025 waves",
    " of the National Survey of Gun Policy (N=", 
    nrow(clean), ").",
    " The survey weights are scaled to the",
    " gun-owner population.",
    " Respondents can indicate ownership of",
    " multiple types of firearms so percentages ",
    "do not add up to 100%."
  )
```

## Percentage by gun type x gender

```{r}
gen <-
  long %>% 
  group_by(guntype, year, gender) %>% 
  summarise(
    pct = sum(owner*weight)/sum(weight),
    .groups = "drop"
    ) %>% 
  filter(gender != "Other")
```


### Bar chart over time

```{r}
gen %>% 
  ggplot(
    aes(
      x = year,
      y = pct,
      group = guntype,
      fill = guntype,
      color = guntype
      )
  ) +
  geom_col(position = "dodge") +
  scale_fill_paletteer_d("MoMAColors::Althoff") +
  scale_color_paletteer_d("MoMAColors::Althoff") +
  facet_wrap(~gender, ncol = 1) +
  theme  +
  guides(
    color = "none"
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5)
    ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1),
    breaks = seq(0, 1, .25)
  ) +
  labs(
    x = "Survey Year",
    y = "Percent of U.S. Gun Owners",
    fill = "Type of firearm"
    )


ggsave(
  height = 8,
  width = 8,
  here(
    path_ol,
    "figures/type-gen.jpeg"
  )
)
```


### Caption

```{r}
note <-
  paste0(
    "This figure plots the percentage of",
    " U.S. gun owners by self-reported",
    " ownership of the following types of",
    " firearms: handgun, shotgun, rifle, and other.",
    " The data are from the 2013, 2023 and 2025 waves",
    " of the National Survey of Gun Policy (N=", 
    nrow(filter(clean, gender != "Female")),
    ").",
    " The survey weights are scaled to the",
    " gun-owner population.",
    " Respondents can indicate ownership of",
    " multiple types of firearms so percentages ",
    "do not add up to 100%."
  )
```
