# Gum Ownership Over Time

This chapter creates a descriptive table and figure that focuses on how the "percentage of US adults that are gun owners" changes from 2013 to 2025. We use `weight` for this analysis, which is the General population weight.

**Input**

-   DemographyGunOwners/data/clean/gs-all.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
all <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-all.rds"
      )
  )
```


## Table

```{r}
demovars <-
  c(
    "gunowner",
    "gender",
    "age4",
    "race",
    "educ5",
    "employ",
    "income4",
    "hhsize4",
    "marital",
    "region4"
  )

summstats <-
  c("notNA(x)", "mean(x)")
summnames <-
  c("Observations", "Mean")
```

```{r, results = "asis", warning = F, message = F}
st(
  gun %>% filter(year > 2019 | year < 2017),
  vars = demovars,
  group.weights = "weight",
  group = "year",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```
