# Political affiliation

This chapter focuses time trends in political affiliation


**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`
```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "paletteer",
    "fastDummies",
    "MetBrewer"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```

## Manually compute shares as a sanity check

Everything aligns perfectly.
```{r}
gun %>% 
  group_by(year, party5) %>% 
  summarise(w = sum(weight), .groups = "drop_last") %>% 
  mutate(pct = w / sum(w)) %>% 
  ungroup()
```

## Estimates with CIs

```{r}
gun %>% tabyl(party5)
```

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

party <- 
  svyby(
    ~party5,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))


long <-
  party %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    party = 
      case_when(
        group == "party5republican" ~ 1,
        group == "party5lean_republican" ~ 2,
        group == "party5independent" ~ 3,
        group == "party5lean_democrat" ~ 4,
        group == "party5democrat" ~ 5,
        TRUE ~ 6) %>% 
      factor(
        levels = 1:6,
        labels = 
          c(
            "Republican",
            "Lean Republican",
            "Independent",
            "Lean Democrat",
            "Democrat", 
            "Missing"
            )
        ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  ) %>% 
  filter(
    !is.na(pct),
     year != 2015 & party != "Missing"
    ) %>% 
  mutate(
    year = factor(year),
    party = droplevels(party)
  )
```


## Gun ownership by political party

```{r}
cap <-
  paste0(
    "This figure plots percent of U.S. gun-owner adults (N=",
     nrow(filter(gun , year != 2015)),") stratified by political affiliation ",
     "from seven waves of the National ",
    "Survey of Gun Policy, spanning 2013 to 2025. 2015 is excluded because of missing data."
  )
```


```{r}
long %>% 
  ggplot(
    aes(
      x = year,
      y = pct,
      group = party,
      color = party,
      fill = party
      )
  ) +
  geom_col() +
  theme +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5)
  ) +
  guides(color = "none") +
  scale_color_paletteer_d("MetBrewer::Benedictus") +
  scale_fill_paletteer_d("MetBrewer::Benedictus") +
  scale_y_continuous(breaks = seq(0, 100, 20)) +
  labs(
    x = "Survey Year",
    y = "Percent of US gun owners",
    fill = "Political affiliation"
  ) 

ggsave(
  height = 6,
  width = 8, 
  here(
    path_ol, 
    "figures/political.jpeg"
  )
)
```

## Line plot

```{r}
long %>% 
  mutate(
    across(
      c("pct", "ci_l", "ci_u"), ~ .x/100
    )
  ) %>% 
  ggplot(
    aes(
      x = year,
      color = party,
      group = party,
      y = pct
      )
  ) +
  geom_line(
    linewidth = 1.2,
    linetype = "dashed"
  ) +
  geom_ribbon(
    aes(
      ymin = ci_l,
      ymax = ci_u,
      fill = party
      ),
    alpha = .7,
    color = NA
  ) +
  scale_color_paletteer_d("MetBrewer::Benedictus") +
  scale_fill_paletteer_d("MetBrewer::Benedictus") +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, .8),
    breaks = seq(0, .8, .1)
  ) +
  theme + 
  guides(
    color = "none",
    fill = 
      guide_legend(
        override.aes = 
          list(alpha = 1)
        )
  ) +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5)
  ) +
  labs(
    x = "Survey Year",
    y = "Percent of US gun owners",
    color = "Political affiliation",
    fill = "Political affiliation"
  ) 

ggsave(
  height = 6,
  width = 8, 
  here(
    path_ol, 
    "figures/political-line.jpeg"
  )
)
```


## 3 party 

### Estimates with CIs

```{r}
gun %>% tabyl(party3)
```

```{r}
des <- svydesign(
  ids = ~1,          # no clustering info
  weights = ~weight,
  data = gun
)

party <- 
  svyby(
    ~party3,
    ~year,
    des,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
    ) %>% 
  clean_names() %>% 
  mutate(across(-year, ~ .x * 100))


long <-
  party %>%
  pivot_longer(
    cols = -year,
    names_to = "name",
    values_to = "value"
  ) %>%
  mutate(
    stat = case_when(
      str_starts(name, "se_")   ~ "se",
      str_starts(name, "ci_l_") ~ "ci_l",
      str_starts(name, "ci_u_") ~ "ci_u",
      TRUE                      ~ "pct"
    ),
    group = name %>%
      str_remove("^se_") %>%
      str_remove("^ci_l_") %>%
      str_remove("^ci_u_")
  ) %>%
  select(year, group, stat, value) %>%
  pivot_wider(
    names_from = stat,
    values_from = value
  ) %>% 
   mutate(
    party = 
      case_when(
        group == "party3republican" ~ 1,
        group == "party3independent" ~ 2,
        group == "party3democrat" ~ 3,
        TRUE ~ 4) %>% 
      factor(
        levels = 1:4,
        labels = 
          c(
            "Republican",
            "Independent",
            "Democrat", 
            "Missing"
            )
        ),
    # lower CI at least 0
    ci_l = ifelse(ci_l < 0, 0, ci_l)
  ) %>% 
  filter(
    !is.na(pct),
     year != 2015 & party != "Missing"
    ) %>% 
  mutate(
    year = factor(year),
    party = droplevels(party)
  )
```


### Gun ownership by political party

```{r}
cap <-
  paste0(
    "This figure plots percent of U.S. gun-owner adults (N=",
     nrow(filter(gun , year != 2015)),") stratified by political affiliation ",
     "from seven waves of the National ",
    "Survey of Gun Policy, spanning 2013 to 2025. 2015 is excluded because of missing data."
  )
```


```{r}
long %>% 
  ggplot(
    aes(
      x = year,
      y = pct,
      group = party,
      color = party,
      fill = party
      )
  ) +
  geom_col() +
  theme +
  theme(
    legend.position = "top",
    legend.direction = "horizontal",
    legend.background = element_blank(),
    legend.key = element_blank(),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 10, hjust = .5)
  ) +
  guides(color = "none") +
  scale_color_paletteer_d("MetBrewer::Benedictus") +
  scale_fill_paletteer_d("MetBrewer::Benedictus") +
  scale_y_continuous(breaks = seq(0, 100, 20)) +
  labs(
    x = "Survey Year",
    y = "Percent of US gun owners",
    fill = "Political affiliation"
  ) 

ggsave(
  height = 6,
  width = 8, 
  here(
    path_ol, 
    "figures/political-3.jpeg"
  )
)
```
