# Basic Demographics of Gun Owners

This chapter creates a descriptive table that shows the demographics of US 
gun owners from 2013 to 2025.

**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```

## Demographic Table

```{r}
demovars <-
  c(
    "gender",
    "age4",
    "racethnicity",
    "party3",
    "educ",
    "employ",
    "income4",
    "marital",
    "region4"
  )

summstats <-
  c("notNA(x)", "mean(x)")
summnames <-
  c("Observations", "Mean")
```

Solid table, not perfect so I replace with the tables below

```{r, results = "asis", warning = F, message = F}
# st(
#   gun,
#   vars = demovars,
#   group = "year",
#   group.weights = "weight",
#   summ = summstats,
#   summ.names = summnames,
#   title = "Demographic characteristics of US gun owners between 2013 and 2025",
#   out = "latex",
#   factor.counts = TRUE,
#   numformat = "comma",
#   labels = T,
#   file =  here(path_ol, "tables/gun-own-chars"),
#   digits = 2,
#   anchor = "tab:GunOwnChars",
#   fit.page = '\\textwidth'
#   )
```


```{r}
nrow(gun)
```


## Full sample summary

```{r}
ds <- survey::svydesign(~ 1, data = gun, weights = ~ weight) 
```

```{r}
# overview <-
#   ds %>%
#   tbl_svysummary(
#     include = demovars,
#     statistic = list(all_categorical() ~ "{p}%")
#   )
# 
# overview %>% 
#   as_gt() %>%
#   tab_options(
#     table.layout = "fixed",
#     latex.tbl.pos = "h",
#     data_row.padding = gt::px(2),
#     row_group.padding = gt::px(3),
#     summary_row.padding = gt::px(2)
#   ) %>%
#   tab_caption(
#     "Demographic characteristics of US gun owners, 2013 and 2025"
#   ) %>% 
#   gtsave(filename = here(path_ol, "tables/summary.tex")
#   )

# Replace \fontsize{12.0pt}{14.0pt}\selectfont = \fontsize{10.5pt}{12.5pt}\selectfont with 
# And add \label{tab:Summary}
```

```{r}
overview <-
  ds %>%
  tbl_svysummary(
    include = demovars,
    statistic = list(all_categorical() ~ "{p}%")
  )
```

```{r}
overview_tbl <-
  overview %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "Overall"
  ) %>%
  modify_table_body(
    ~ .x %>%
      mutate(
        label = ifelse(
          row_type == "level",
          paste0("  ", label),
          label
        )
      )
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>% 
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "Summary"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  ) %>%
  column_spec(1, width = "6cm") %>%
  column_spec(2, width = "2cm") %>% 
  footnote(
    general = "Values are weighted percentages describing the demographic composition of U.S. gun owners between 2013 and 2025.",
    threeparttable = TRUE,
    escape = FALSE
  )

save_kable(
  overview_tbl,
  file = here(path_ol, "tables/summary.tex")
)
```


```{r}
overview <-
  ds %>%
  tbl_svysummary(
    include = demovars,
    statistic = list(all_categorical() ~ "{p}%")
  )

overview_tbl <-
  overview %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "Overall"
  ) %>%
  modify_table_body(
    ~ .x %>%
      mutate(
        label = ifelse(row_type == "level", paste0("  ", label), label)
      )
  ) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  kable_styling(
    latex_options = c("hold_position"),   # ← removed scale_down
    font_size = 9,
    position = "center"
  ) %>%
  column_spec(1, width = "0.65\\textwidth") %>%  # ← expand stub
  column_spec(2, width = "0.25\\textwidth") %>%  # ← expand data col
  footnote(
    general =
      "Values are weighted percentages describing the demographic composition of U.S. gun owners between 2013 and 2025.",
    threeparttable = TRUE,
    escape = FALSE
  )

save_kable(
  overview_tbl,
  file = here(path_ol, "tables/summary.tex")
)



```

## By year

```{r}
year <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  )


year %>% 
  as_gt() %>%
  tab_options(
    table.layout = "fixed",
    latex.tbl.pos = "h",
    data_row.padding = gt::px(2),
    row_group.padding = gt::px(3),
    summary_row.padding = gt::px(2)
  ) %>% 
  tab_caption("Demographic characteristics of US gun owners by year, 2013 and 2025") %>% 
  
  gtsave(filename = here(path_ol, "tables/summary-year.tex"))

# Replace \fontsize{12.0pt}{14.0pt}\selectfont = \fontsize{10.5pt}{12.5pt}\selectfont with 
# And add \label{tab:SummaryYear}
```


```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  modify_header(
    label ~ "Characteristic",
    stat_1 ~ "2013 (N = 947)",
    stat_2 ~ "2015 (N = 524)",
    stat_3 ~ "2017 (N = 530)",
    stat_4 ~ "2019 (N = 445)",
    stat_5 ~ "2021 (N = 807)",
    stat_6 ~ "2023 (N = 997)",
    stat_7 ~ "2025 (N = 1,002)"
  ) %>%
  modify_table_body(
    ~ .x %>%
      mutate(
        label = ifelse(
          row_type == "level",
          paste0("  ", label),
          label
        )
      )
  ) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "SummaryYear"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  ) %>%
  column_spec(1, width = "4cm") %>%
  column_spec(2:8, width = "1.3cm") %>% 
  footnote(
    general = "Values are weighted percentages describing the demographic composition of U.S. gun owners by survey year.",
    threeparttable = TRUE,
    escape = FALSE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/summary-year.tex")
)
```
