# Basic Demographics of Gun Owners

This chapter creates a descriptive table that shows the demographics of US 
gun owners from 2013 to 2025.

**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```

## Demographic variables

```{r}
demovars <-
  c(
    "gender",
    "age4",
    "racethnicity",
    "party5",
    "nranow",
    "educ",
    "employ",
    "income4",
    "marital",
    "region4"
  )
```

## Tabulate values

```{r}
walk(demovars, ~ print(tabyl(gun, .x)))
```

## Yearly Summary

This function gives the same results as `tbl_svysummary` but includes standard errors. 
```{r}
svyby(~age4, ~year, design = ds, FUN = svymean)
```


```{r}
note <-
  paste0(
    "Note: ``Other'' gender category was not ",
    "included until the 2025 wave. ",
    "``Some college or associate’s degree'' also includes respondents ",
    "reporting vocational or technical school. ",
    "Political affiliation was not collected in the 2015 wave. ",
    "Asian, non-Hispanic was not included as a separate category until ",
    "the 2019 wave. ",
    "For marital status, ``Living with partner'' was not included ",
    "in the 2023 and 2025 waves."
  )
```

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      
      tbl <- .x
      
      # compute unweighted yearly totals
      counts <- format(table(gun$year), big.mark = ",")
      
      # create row with obs
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(
          as.list(counts),
          paste0("stat_", seq_along(counts))
          )
        ) %>% 
        mutate(across(starts_with("stat"), trimws)
          )

      bind_rows(n_row, tbl) %>%
        filter(
          ! grepl("Ques", label),
          ! (grepl("No", label) & variable == "nranow")
          ) %>% 
        mutate(
           label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label),
            label),
          stat_2 = ifelse(
               variable == "party5" & row_type == "level",
               "-", stat_2
               ),
          across(
            # Asian american not a category until > 2017
            c(stat_1, stat_2, stat_3),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label),
              "-", .x
              )
            ),
           across(
             c(stat_6, stat_7),
             ~ ifelse(
               variable == "marital" & grepl("with", label),
               "-", .x
               )
             ),
           across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%",
              "-", .x
              )
          )
           )
    }
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "SummaryYear",
    linesep = "",
    caption = 
      "Demographic Composition of US Adult Gun Owners by Survey Year, 2013–2025",
    align = "lccccccc"
  ) %>%
  kable_styling(
    latex_options =
      c("hold_position", "scale_down"),
   font_size = 9
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(
    2:8, 
    width = "1.3cm"
    ) %>% 
  row_spec(1, bold = TRUE) %>% 
  footnote(
   number = note,
   threeparttable = TRUE,
   escape = FALSE,
   fixed_small_size = TRUE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/summary-year.tex")
)
```


## Significance test
Chi-squared p-values (survey-weighted, Rao-Scott)

```{r}
chisq_pvals <- map_dbl(demovars, function(v) {
  des_sub <- subset(ds, year %in% c(2013, 2025))
  
  # For gender, drop "Other" since it wasn't a category in 2013
  if (v == "gender") {
    des_sub <- subset(des_sub, gender != "Other")
    des_sub$variables$gender <- droplevels(des_sub$variables$gender)
  }
  
  if (v == "party5") {
    des_sub <- subset(des_sub, party5 != "Unknown")
    des_sub$variables$party5 <- droplevels(des_sub$variables$party5)
  }
  
  if (v == "marital") return(NA_real_)
  
  tryCatch(
    svychisq(
      as.formula(paste0("~", v, " + year")),
      design    = des_sub,
      statistic = "Chisq"
    )$p.value,
    error   = function(e) NA_real_,
    warning = function(w) NA_real_
  )

}) %>%
  set_names(demovars)

# Format p-value for display
format_pval <- function(p) {
  case_when(
    is.na(p)   ~ "",
    p < 0.001  ~ "<0.001",
    p < 0.01   ~ sprintf("%.3f", p),
    p < 0.05   ~ sprintf("%.3f", p),
    TRUE       ~ sprintf("%.2f", p)
  )
}
```

Compute weighted % for one variable in one year
```{r}
get_pct <- function(var, yr) {
  svyby(
    formula = as.formula(paste0("~", var)),
    by      = ~year,
    design  = subset(ds, year %in% c(yr)),
    FUN     = svymean,
    na.rm   = TRUE
  ) %>%
    select(-year, -starts_with("se.")) %>%
    unlist() %>%
    `*`(100) %>%
    round(1)
}

pct_2013 <- map(demovars, ~ get_pct(.x, 2013)) %>% set_names(demovars)
pct_2025 <- map(demovars, ~ get_pct(.x, 2025)) %>% set_names(demovars)

# Build a lookup: variable + level → difference string
diff_lookup <- map_dfr(demovars, function(v) {
  p13 <- pct_2013[[v]]
  p25 <- pct_2025[[v]]
  
  # svyby prefixes names with the variable name; strip it
  lvls <- names(p13) %>% str_remove(paste0("^", v))
  
  tibble(
    variable = v,
    level    = lvls,
    diff_str = sprintf("%+.1f", p25 - p13)
  )
})
```


```{r}
note <-
  paste0(
    "``Some college or associate's degree'' includes vocational or technical school. ",
    "``Other'' gender, Asian non-Hispanic, and ``Living with partner'' were not collected ",
    "in all waves (introduced in 2025, 2019, and discontinued after 2021, respectively). ",
    "Political affiliation was not collected in 2015. ",
    "\\\\textit{p}-values are from Rao-Scott chi-squared tests comparing 2013 and 2025 ",
    "weighted distributions; the gender test is restricted to Male/Female, the ",
    "political affiliation test excludes ``Unknown'' responses, and marital status ",
    "is not tested because response categories differed across waves."
  )
```


```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by        = year,
    include   = demovars,
    missing   = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      tbl <- .x

      # Unweighted totals row
      counts <- format(table(gun$year), big.mark = ",")
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(as.list(counts), paste0("stat_", seq_along(counts)))
      ) %>%
        mutate(across(starts_with("stat"), trimws))

      tbl <- bind_rows(n_row, tbl) %>%
        filter(
          !grepl("Ques", label),
          !(grepl("No", label) & variable == "nranow")
        ) %>%
        mutate(
          label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label), label
          ),
          stat_2 = ifelse(
            variable == "party5" & row_type == "level", "-", stat_2
          ),
          across(
            c(stat_1, stat_2, stat_3),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label), "-", .x
            )
          ),
          across(
            c(stat_6, stat_7),
            ~ ifelse(
              variable == "marital" & grepl("with", label), "-", .x
            )
          ),
          across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%", "-", .x
            )
          )
        )

      # Add difference and p-value columns
      tbl <- tbl %>%
        rowwise() %>%
        mutate(
          stat_diff = case_when(
            label == "Observations"  ~ "",
            row_type == "label" & !is.na(variable) ~ "",
            variable == "gender" & grepl("Other", label) ~ "-",
            variable == "marital" & grepl("partner", label, ignore.case = TRUE) ~ "-",
            row_type == "level" ~ {
              lv      <- str_trim(label)
              matched <- diff_lookup %>%
                filter(variable == .data$variable, level == lv)
              if (nrow(matched) == 1) matched$diff_str else "-"
            },
            TRUE ~ ""
          ),
          stat_pval = case_when(
            label == "Observations"                    ~ "",
            row_type == "label" & !is.na(variable)    ~ {
              format_pval(chisq_pvals[variable])
            },
            TRUE ~ ""
          )
        ) %>%
        ungroup()

      tbl
    }
  ) %>%
  modify_header(
    stat_diff ~  "2025--2013",
    stat_pval ~ "p-value"
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    label     = "SummaryYearTest",
    linesep   = "",
    caption   = "Demographic Composition of US Adult Gun Owners by Survey Year, 2013-2025",
    align     = "lccccccccc"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(2:8, width = "1.3cm") %>%
  column_spec(9, width = "1.3cm") %>%
  column_spec(10, width = "1.3cm") %>%
  row_spec(1, bold = TRUE) %>%
  footnote(
    number           = note,
    threeparttable   = TRUE,
    escape           = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/summary-year-2.tex")
)
```


## Full sample summary

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = gun, 
    strata = ~ year,
    weights = ~ weight
    ) 
```

```{r}
overview <-
  ds %>%
  tbl_svysummary(
    include = demovars,
    statistic = list(all_categorical() ~ "{p}%")
  )
```

```{r}
overview_tbl <-
  overview %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "Overall"
  ) %>%
  modify_table_body(
    ~ .x %>%
      filter(
        !(variable == "nranow" & label == "No"),
        ! grepl("Ques", label)
        ) %>%
      mutate(
        label = ifelse(
          row_type == "level",
          paste0("  ", label),
          label
        ),
         # replace 0% in 2025 to indicate question was not asked
         stat_0 = case_when(
           variable == "marital" & grepl("with", label) ~ ".",
           TRUE ~ stat_0
           )
      )
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "Demographic Characteristics of U.S. Gun Owners",
    linesep = "",
    align = "lc",
    label = "Summary"
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 8
  ) %>%
  column_spec(1, width = "6cm") %>%
  column_spec(2, width = "2cm") %>% 
  footnote(
    number = 
      paste(
        "Values are weighted percentages",
        "describing the demographic composition of U.S.",
        "gun owners between 2013 and 2025.",
        "Due to differences in the survey questions",
        "over time, ``Some college or associates degree''",
        "also includes responses for vocational or technical schools."
        ),
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = T
  )

save_kable(
  overview_tbl,
  file = here(path_ol, "tables/summary.tex")
)
```

