# Basic Demographics of Gun Owners

This chapter creates a descriptive table that shows the demographics of US 
gun owners from 2013 to 2025.

**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(packages, character.only = TRUE)
```

## Load data

```{r}
gun <-
  read_rds(
    here(
      path_od,
      "data",
      "clean",
      "gs-gun-owners.rds"
    )
  )

genpop <- read_rds(here(path_od, "data", "clean", "gs-gen-pop.rds"))
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = gun,
  strata  = ~year,
  weights = ~weight
)

ds_genpop <- svydesign(
  ~1,
  data    = genpop,
  strata  = ~year,
  weights = ~weight
)
```

## Demographic variables

```{r}
demovars <-
  c(
    "gender",
    "age4",
    "racethnicity",
    "party5",
    "nranow",
    "educ",
    "employ",
    "income4",
    "marital",
    "region4"
  )
```

## Tabulate values

```{r}
walk(demovars, ~ print(tabyl(gun, .x)))
```

## Gun ownership prevalence from general population

Gets the share of US gun ownership over time from the general population survey.

```{r}
years_ordered <- sort(unique(genpop$year))

gun_prev_raw <- svyby(
  formula = ~gun,
  by      = ~year,
  design  = ds_genpop,
  FUN     = svymean,
  na.rm   = TRUE
)

# gun is coded "Yes"/"No" in genpop
gun_prev_vec <- gun_prev_raw$gunYes

# Format as percentage strings (e.g. "22%")
gun_prev_strs <- paste0(round(gun_prev_vec * 100), "%")

# Diff: 2025 minus 2013 in percentage points
gun_diff_str <- sprintf(
  "%+.1f",
  (gun_prev_vec[length(gun_prev_vec)] - gun_prev_vec[1]) * 100
)

# P-value: Rao-Scott chi-squared comparing 2013 and 2025
genpop_ends  <- filter(genpop, year %in% range(years_ordered))
ds_ends      <- svydesign(~1, data = genpop_ends, weights = ~weight)
gun_pval_raw <- svychisq(~gun + year, ds_ends, statistic = "Chisq")$p.value
```

## Significance tests

Chi-squared p-values (survey-weighted, Rao-Scott) for gun owner demographics.

```{r}
chisq_pvals <- map_dbl(demovars, function(v) {
  des_sub <- subset(ds, year %in% c(2013, 2025))

  if (v == "gender") {
    des_sub <- subset(des_sub, gender != "Other")
    des_sub$variables$gender <- droplevels(des_sub$variables$gender)
  }

  if (v == "party5") {
    des_sub <- subset(des_sub, party5 != "Unknown")
    des_sub$variables$party5 <- droplevels(des_sub$variables$party5)
  }

  if (v == "nranow") {
    des_sub <- subset(des_sub, nranow != "Unknown")
    des_sub$variables$nranow <- droplevels(des_sub$variables$nranow)
  }

  if (v == "marital") return(NA_real_)

  tryCatch(
    svychisq(
      as.formula(paste0("~", v, " + year")),
      design    = des_sub,
      statistic = "Chisq"
    )$p.value,
    error   = function(e) NA_real_,
    warning = function(w) NA_real_
  )
}) %>%
  set_names(demovars)

# Add gun ownership prevalence p-value so it is picked up automatically
# by the label-row case_when in the table
chisq_pvals["gun_prev"] <- gun_pval_raw
```

## P-value formatter

```{r}
format_pval <- function(p) {
  case_when(
    is.na(p)  ~ "",
    p < 0.001 ~ "<0.001",
    p < 0.01  ~ sprintf("%.3f", p),
    p < 0.05  ~ sprintf("%.3f", p),
    TRUE      ~ sprintf("%.2f", p)
  )
}
```

## Difference lookup

Compute weighted % for one variable in one year, then build a lookup
table of variable + level -> difference string.

```{r}
get_pct <- function(var, yr) {
  svyby(
    formula = as.formula(paste0("~", var)),
    by      = ~year,
    design  = subset(ds, year %in% c(yr)),
    FUN     = svymean,
    na.rm   = TRUE
  ) %>%
    select(-year, -starts_with("se.")) %>%
    unlist() %>%
    `*`(100) %>%
    round(1)
}

pct_2013 <- map(demovars, ~ get_pct(.x, 2013)) %>% set_names(demovars)
pct_2025 <- map(demovars, ~ get_pct(.x, 2025)) %>% set_names(demovars)

diff_lookup <- map_dfr(demovars, function(v) {
  p13  <- pct_2013[[v]]
  p25  <- pct_2025[[v]]
  lvls <- names(p13) %>% str_remove(paste0("^", v))
  tibble(
    variable = v,
    level    = lvls,
    diff_str = sprintf("%+.1f", p25 - p13)
  )
})
```

## Table note

```{r}
note <-
  paste0(
    "Gun ownership prevalence is the weighted prevalence of gun ",
    "ownership in the full general population sample. ",
    "``Some college or associate's degree'' includes vocational or technical school. ",
    "``Other'' gender, Asian non-Hispanic, and ``Living with partner'' were not collected ",
    "in all waves (introduced in 2025, 2019, and discontinued after 2021, respectively). ",
    "Political affiliation was not collected in 2015. ",
    "p-values are from Rao-Scott chi-squared tests comparing 2013 and 2025 ",
    "weighted distributions; the gender test is restricted to Male/Female, the ",
    "political affiliation test excludes ``Unknown'' responses, and marital status ",
    "is not tested because response categories differed across waves."
  )
```

## Yearly summary table

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by        = year,
    include   = demovars,
    missing   = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      tbl <- .x

      # Observations row (unweighted counts)
      counts <- format(table(gun$year), big.mark = ",")
      n_row <- tibble(
        variable  = NA_character_,
        var_label = NA_character_,
        row_type  = "label",
        label     = "Observations",
        !!!setNames(as.list(counts), paste0("stat_", seq_along(counts)))
      ) %>%
        mutate(across(starts_with("stat"), trimws))

      # Gun ownership prevalence row (weighted, from genpop)
      gun_prev_row <- tibble(
        variable  = "gun_prev",
        var_label = "Gun ownership prevalence",
        row_type  = "label",
        label     = "Gun ownership prevalence",
        !!!setNames(
          as.list(gun_prev_strs),
          paste0("stat_", seq_along(gun_prev_strs))
        )
      )

      # Bind: Observations | Gun prevalence | everything else
      tbl <- bind_rows(n_row, gun_prev_row, tbl) %>%
        filter(
          !grepl("Ques", label),
          !(grepl("No", label) & variable == "nranow")
        ) %>%
        mutate(
          label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label), label
          ),
          stat_2 = ifelse(
            variable == "party5" & row_type == "level", "-", stat_2
          ),
          across(
            c(stat_1, stat_2, stat_3),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label), "-", .x
            )
          ),
          across(
            c(stat_6, stat_7),
            ~ ifelse(
              variable == "marital" & grepl("with", label), "-", .x
            )
          ),
          across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%", "-", .x
            )
          )
        )

      # Diff and p-value columns
      tbl <- tbl %>%
        rowwise() %>%
        mutate(
          stat_diff = case_when(
            label == "Observations"                                              ~ "",
            variable == "gun_prev"                                               ~ gun_diff_str,
            row_type == "label" & !is.na(variable)                              ~ "",
            variable == "gender"  & grepl("Other",   label)                     ~ "-",
            variable == "marital" & grepl("partner", label, ignore.case = TRUE) ~ "-",
            row_type == "level" ~ {
              lv      <- str_trim(label)
              matched <- diff_lookup %>%
                filter(variable == .data$variable, level == lv)
              if (nrow(matched) == 1) matched$diff_str else "-"
            },
            TRUE ~ ""
          ),
          stat_pval = case_when(
            label == "Observations"                 ~ "",
            row_type == "label" & !is.na(variable) ~ format_pval(chisq_pvals[variable]),
            TRUE                                    ~ ""
          )
        ) %>%
        ungroup()

      tbl
    }
  ) %>%
  modify_header(
    stat_diff ~ "2025--2013",
    stat_pval ~ "p-value"
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    label     = "SummaryYearTest",
    linesep   = "",
    caption   = "Demographic Composition of US Adult Gun Owners by Survey Year, 2013-2025",
    align     = "lccccccccc"
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 9
  ) %>%
  column_spec(1,   width = "4.5cm") %>%
  column_spec(2:8, width = "1.3cm") %>%
  column_spec(9,   width = "1.3cm") %>%
  column_spec(10,  width = "1.3cm") %>%
  row_spec(1, bold = TRUE) %>%
  footnote(
    number           = note,
    threeparttable   = TRUE,
    escape           = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/summary-year-2.tex")
)
```

## Full sample summary

```{r}
overview <-
  ds %>%
  tbl_svysummary(
    include   = demovars,
    statistic = list(all_categorical() ~ "{p}%")
  )

overview_tbl <-
  overview %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "Overall"
  ) %>%
  modify_table_body(
    ~ .x %>%
      filter(
        !(variable == "nranow" & label == "No"),
        !grepl("Ques", label)
      ) %>%
      mutate(
        label = ifelse(
          row_type == "level",
          paste0("  ", label),
          label
        ),
        stat_0 = case_when(
          variable == "marital" & grepl("with", label) ~ ".",
          TRUE ~ stat_0
        )
      )
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Demographic Characteristics of U.S. Gun Owners",
    linesep   = "",
    align     = "lc",
    label     = "Summary"
  ) %>%
  kable_styling(
    latex_options = c("HOLD_position", "scale_down"),
    font_size = 8
  ) %>%
  column_spec(1, width = "6cm") %>%
  column_spec(2, width = "2cm") %>%
  footnote(
    number =
      paste(
        "Values are weighted percentages",
        "describing the demographic composition of U.S.",
        "gun owners between 2013 and 2025.",
        "Due to differences in the survey questions",
        "over time, ``Some college or associates degree''",
        "also includes responses for vocational or technical schools."
      ),
    threeparttable   = TRUE,
    escape           = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  overview_tbl,
  file = here(path_ol, "tables/summary.tex")
)
```
