# Summary Statistics

Survey weights were normalized to the sample size; therefore, population counts were obtained by multiplying weighted proportions by external Census population totals.

**Input**

-   DemographyGunOwners/data/clean/gs-all.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   01/28/26

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
all <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-all.rds"
      )
  )
```

## Set up survey design

This uses the package `survey`.

```{r}
des <- svydesign(
  ids = ~1,                 # no clustering
  weights = ~ weight,   # your post-strat weight
  data = all
) 
```

### Play around

```{r}
svymean(~ gunhome, des)
```

```{r}
svyby(
  ~gunhome,
  ~gender,
  des,
  svymean
)
```

## Create basic demographics table 


```{r}
demovars <-
  c(
    "gunhome",
    "party7",
    "age4",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income4",
    "internet"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  all,
  vars = demovars,
  group = "year",
  group.weights = "weight",
  summ = summstats,
  summ.names = summnames,
  title = "Sample characteristics",
  numformat = "comma",
  labels =  T,
  out = "latex",
  file = 
    here(
      path_ol, 
      "tables",
      "sample-chars.tex"
    )
  )
```
