# Clean 2013 data

This script cleans the 2013 survey data.

List of variables are different/not available in comparison to later survey years:

-   Different

    -   `income`

    -   `racethnicity`

        -   No level for Asian, non-Hispanic

    -   `hh02` vs `hh01` in later years

-   Not available here but available for later years

    -   `phoneservice`
    -   `nraever`

-   Not available in later years

    -   `core_par`

    -   `ever_par`

    -   `hhhead`

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2013.rds

**Output**

-   DemographyGunOwners/data/gunsurvey/2013.rds

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
source <-
  read_dta(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2013.dta"
      )
  ) %>% 
  clean_names() %>% 
  # many variable names have the pp prefix, so remove it
  rename_with(
    ~ str_remove_all(.x, "pp"), 
    everything()
  )
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2013 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|--------------------------|---------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                            |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Select variables

```{r}
demovars <-
  c(
    "age",
    "agecat",
    "agect4",
    "educ",
    "educat",
    "ethm",
    "gender",
    # Household 
    "hhhead",
    "hhsize",
    "t01",
    "t25",
    "t1317",
    "t18ov",
    "t612",
    "core_par",
    "ever_par",
    "marit",
    # Region
    "staten",
    "reg4",
    "reg9",
    "msacat",
    # House type
    "house",
    "rent",
    # Income
    "incimp",
    # Internet
    "net",
    # Employment
    "work",
    "partyid7"
  )

gun <-
  c(
    # Respondent personally owns a gun
    "xpa0122",
    # Respondent has a gun in household but is not owner
    "xpa0117",
    # Type of gun
    "pa0118",
    "pa0119",
    "pa0120",
    "pa0121",
    "pa0128",
    # NRA
    "q30",
    #Victim of gun violence
    "q31"
  )
```

```{r}
raw <-
  source %>% 
  clean_names() %>% 
  rename(
    # used to select general population sample
    sample = xsample 
  ) %>% 
  select(
    starts_with(c("caseid", "weight", "sample")),
    all_of(c(demovars, gun))
  )
```

##  Sample

This variable is used to restrict the sample to the general population sample or the gun augment sample. We use the first when we are making general population inference.

```{r}
raw %>% tabyl(sample)
```

```{r}
raw <- 
  raw %>% 
  mutate(sample = as_factor(sample)) %>% 
  set_variable_labels(
    sample = "Type of Respondents"
  )
```

```{r}
raw %>% tabyl(sample)
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender = droplevels(as_factor(gender))
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(agecat)
raw %>% tabyl(agect4)
```

```{r}
raw <-
  raw %>% 
  rename(age4 = agect4,
         age7 = agecat) %>% 
  mutate(
    across(
      c("age4", "age7"),
      ~ droplevels(as_factor(.x))
    )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

Note that 2013 data does not include Asian, non-Hispanic as a category.

```{r}
raw %>% tabyl(ethm)
```

```{r}
raw <-
  raw %>% 
  rename(racethnicity = ethm) %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:5,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+ Races, non-Hispanic"
            )
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

We drop `educ` because it corresponds to 4 levels of education, but in order to match later survey years, we define education using 5 levels.

```{r}
raw %>% tabyl(educ)
raw %>% tabyl(educat)
```

```{r, warning = F}
raw <-
  raw %>% 
  mutate(
    educ5 =
      case_when(
       educ %in% 1:8 ~ 1,
       educ == 9 ~ 2,
       educ %in% 10:11 ~ 3,
       educ == 12 ~ 4,
       TRUE ~ 5
       ) %>% 
      factor(
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Some college/associates degree",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      ),
    educ = 
      factor(
        educ,
        levels = 1:14,
        labels = 
          c("No formal education",
            "1st, 2nd, 3rd, or 4th grade",
            "5th or 6th grade",
            "7th or 8th grade",
            "9th grade",
            "10th grade",
            "11th grade",
            "12th grade (no diploma)",
            "HS graduate, diploma, or GED",
            "Some college, no degree",
            "Associate degree",
            "Bachelors degree",
            "Masters degree",
            "Professional or Doctorate degree")
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education",
    educ = "Highest level of education"
  ) %>% 
  select(-"educat")
```

```{r}
walk(
  c("educ", "educ5"),
  ~ print(raw %>% tabyl(.x))
)
```

## Martial status

```{r}
raw %>% tabyl(marit)
```

```{r}
raw <-
  raw %>% 
  rename(marital = marit) %>% 
  mutate(
    marital = droplevels(as_factor(marital))
    ) %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(work)
```

```{r}
raw <-
  raw %>% 
  rename(employ = work) %>% 
  mutate(employ = droplevels(as_factor(employ))) %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

We construct 2 variables: `income4` and `income5` to match 2021 and 2023 data.

Note that in later years, the income variable splits 175-199,999 and then 200,000+ but here it is 175,000+. We leave this as is because we will likely never use this variable, rather we will use the income4 or income9 variables. This income variable also splits the sample into different categories as the later years. We attempt to harmonize it as much as possible.

```{r}
raw %>% tabyl(incimp)
```

```{r}
raw <-
  raw %>% 
  rename(income = incimp) %>% 
  mutate(
    income4 = 
      case_when(
        income %in% 1:8 ~ 1,
        income %in% 9:12 ~ 2,
        income %in% 13:15 ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      case_when(
        income %in% 1:3 ~ 1,
        income %in% 4:6 ~ 2,
        income %in% 7:8 ~ 3,
        income %in% 9:10 ~ 4,
        income == 11 ~ 5,
        income %in% 12:13 ~ 6,
        income %in% 14:15 ~ 7,
        income %in% 16:17 ~ 8,
        TRUE ~ 9
      ) %>% 
      factor(
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      ),
    income = 
      case_when(
        income == 1 ~ 1,
        income %in% 2:3 ~ 2,
        income %in% 4:5 ~ 3,
        TRUE ~ income - 2
      ) %>% 
      factor(
        levels = 1:17,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000 or more"
            )
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
walk(
    raw %>% select(starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## State

```{r}
raw %>% tabyl(staten)
```

```{r}
raw <-
  raw %>%
  rename(state = staten) %>% 
  mutate(state = droplevels(as_factor(state))) %>% 
  set_variable_labels(
    state = "State"
  )
```

```{r}
raw %>% tabyl(state)
```

## Region

```{r}
raw %>% tabyl(reg4)
raw %>% tabyl(reg9)
```

```{r}
raw <-
  raw %>% 
  rename(region4 = reg4, region9 = reg9) %>% 
  mutate(
    across(starts_with("region"), ~ droplevels(as_factor(.x)))
    )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )

```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(msacat)
```

```{r}
raw <-
  raw %>% 
  rename(metro = msacat) %>% 
  mutate(
    metro = droplevels(as_factor(metro)) 
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(net)
```

```{r}
raw <-
  raw %>% 
  rename(internet = net) %>% 
  mutate(
    internet = 
      factor(
        internet,
        levels = 0:1,
        labels = c("Non-internet household",
                   "Internet household")
      )
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(rent)
```

```{r}
raw <-
  raw %>% 
  rename(housing = rent) %>% 
  mutate(
    housing = droplevels(as_factor(housing))
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(house)
```

```{r}
raw <- 
  raw %>% 
  rename(home_type = house) %>% 
  mutate(
    home_type = droplevels(as_factor(home_type))
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```


## Are you a member of the National Rifle Association--also known as the NRA? (Q30)

```{r}
raw %>% tabyl(q30)
```

```{r}
raw <-
  raw %>% 
  rename(nranow = q30) %>% 
  mutate(
    nranow = 
      factor(
        nranow,
        levels = c(1:2, -1),
        labels =
           c("Yes", "No", "Unknown")
      )
    ) %>% 
  set_variable_labels(
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow)
```

## Political party

We construct `party5`

```{r}
raw %>% tabyl(partyid7)
```

```{r}
raw <-
  raw %>% 
  rename(party7 = partyid7) %>% 
  mutate(
    party5 =
      case_when(
        party7 %in% 6:7 ~ 1,
        party7 == 5 ~ 2,
        party7 == 4 ~ 3,
        party7 == 3 ~ 4,
        party7 %in% 1:2 ~ 5
      ) %>% 
      factor(
        levels = 1:5,
        labels = 
          c("Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Republican"
            )
      ),
    party7 = droplevels(as_factor(party7)),
    party3 = 
      case_when(
        grepl("demo", party7, ignore.case = T) ~ 1,
        grepl("repub", party7, ignore.case = T) ~ 2,
        grepl("ind", party7, ignore.case = T) ~ 3
      ) %>% 
      factor(
        levels = 1:3,
        labels = c(
          "Democrat",
          "Republican",
          "Independent"
        )
      )
  ) %>% 
  set_variable_labels(
    party7 = "7-level political affiliation",
    party5 = "5-level political affiliation",
    party3 = "3-level political affiliation"
  )
```

```{r}
raw %>% tabyl(party5)
raw %>% tabyl(party7, party5)
raw %>% tabyl(party3, party7)
```

## Gun ownership

-   Respondent personally owns a gun (`xpa0122`)

-   Respondent has a gun in household but is not owner (`xpa0117`)

In the later survey years, this corresponds to

-   `xpa0122` = `gunhomeper`

-   `xpa0117` not exactly `gunhome`

    -   `gunhome` = Do you happen to have in your home or garage any guns or revolvers?

    -   So in this survey you cant own a gun (`gunhome`) and have a gun in the household and not be the owner (`xpa0117`)

For the purpose of our analysis, we will only be using `gunhomeper` because we just care about people who personally own guns. So we will drop the variable `xpa0117` to minimize confusion. Also in 2015, the only variable they have is "Respondent personally owns a gun", so although we have the variable `gunhome` for later years, it doesn't appear in 2015, so we can't use it for the longitudinal analysis, so we don't lose anything from dropping it.

In the [Barry paper](https://www.sciencedirect.com/science/article/pii/S0091743515001668?via=ihub), they consider "non-gun owners" as people who answer no to being a gun owner but have a gun in their home (N=843). For the purposes of our analysis just looking at gun owners, it doesnt matter. But for the sake of comparing national prevalence estimates across years to check the quality of my data cleaning, it is useful to know why it may differ. However, in 2013, my estimate of 22\% matches Barry's.

```{r}
raw %>% tabyl(xpa0122, xpa0117)
```

```{r} 
raw %>%
  filter(xpa0122 == 2 & xpa0117 == 1) %>% 
  nrow() == 843
```

```{r}
raw <-
  raw %>% 
  rename(gunhomeper = xpa0122) %>% 
  mutate(
    gunhomeper = as_factor(gunhomeper)
  ) %>% 
  set_variable_labels(
    gunhomeper = "Respondent personally owns a gun"
  ) %>% 
  select(-"xpa0117")
```

```{r}
tabyl(raw, gunhomeper)
```

## Type of gun ownership

Do you personally own any of the following types of guns?

-   `pa0118` = pistol/handgun

-   `pa0119` = shotguns

-   `pa0120` = rifles

-   `pa0121` other (binary yes no) and `pa0128` other (string)

```{r}
walk(
  c("pa0118", "pa0119", "pa0120", "pa0121"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <-
  raw %>% 
  rename(
   ownhandgun = pa0118,
   ownrifle = pa0119, 
   ownshot = pa0120,
   ownother = pa0121
  ) %>% 
  mutate(
    across(
        starts_with("own"),
        ~ case_when(
          is.na(.x) & gunhomeper != "Yes" ~ 100,
          is.na(.x) ~ 99,
          TRUE ~ .x
        ) %>% 
          factor(
            levels = c(1, 0, 99:100),
            labels =
              c("Yes", 
                "No",
                "Unknown",
                "Non-gun owner"
                )
          )
    )
  )  %>% 
    set_variable_labels(
      ownhandgun = "Do you personally own any hanguns?",
      ownrifle = "Do you personally own any rifles?", 
      ownshot = "Do you personally own any shotguns?",
      ownother = "Do you personally own any other types of guns?"
    ) %>% 
  select(-pa0128)
```

```{r}
walk(
    raw %>% select(starts_with("own")) %>% names(),
  ~ print(raw %>% tabyl(.x))
)
```

## Victim of gun violence (Q31)

```{r}
raw %>% tabyl(q31)
```

```{r}
raw <-
  raw %>% 
  rename(victim = q31) %>% 
  mutate(
    victim = 
      factor(
        victim,
        levels = c(1:2, -1),
         labels =
            c("Yes", 
              "No",
              "Unknown")
      )
  ) %>% 
  set_variable_labels(
      victim = "Have you ever been the victim of a crime involving a gun?"
  )
```

```{r}
raw %>% tabyl(victim)
```

## Household variables

### Household size

We have

-   `t01` = Presence of Household Members - Children 0-2

In later survey years we have

-   `hh01` = age 0-1

This difference is okay because we will define 2 grouped variables in `9-merge.Rmd`

1.  `hh06` = age 0-6

2.  hh18un = Under 18

```{r}
raw %>% tabyl(hhsize)
```

```{r}
raw <-
  raw %>% 
  rename(
    hh02 = t01,
    hh25 = t25,
    hh612 = t612,
    hh1317 = t1317,
    hh18ov = t18ov,
  ) %>% 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh02 = "Number of HH members age 0-2",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+"
  )
```

### Parental status

```{r}
walk(
  c("core_par", "ever_par", "hhhead"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
vname <-
   c("corepar", "everpar", "hhead")

raw <-
  raw %>% 
  rename(
    hhead = hhhead,
    corepar = core_par,
    everpar = ever_par
  ) %>% 
  mutate(
    across(
      all_of(vname), 
      ~ droplevels(as_factor(.x))
      ),
    corepar = 
      fct_na_value_to_level(
        corepar, 
        level = "Unknown"
        )
  ) %>% 
  set_variable_labels(
    corepar = "Is respondent a parent or legal guardian?",
    everpar = "Has respondent ever been a parent or legal guardian?",
    hhead = "Is respondent the household head?"
  )
```

```{r}
walk(
 vname, ~ print(raw %>% tabyl(.x))
)
```

## Weights

Notice here that the weight numbering here is different to 2021- waves of the survey. `weight2` is the only weight with no missing values.

-   `weight1` = Post-Stratification Weight for 18+ US adult gen pop sample.

    -   For respondents in the genpop sample (n=1454)

-   `weight2` = Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs

    -   A total weight consisting of gen pop + augment samples (controlling demographics to the benchmarks from created from the KnowledgePanel) was then created to scale to the total respondent sample size (n=2728), that is weight2.

    -   We use this when looking at the full sample

-   `weight3` = Post-Stratification Weight for gun owners

    -   This total respondent weight was then used to scale to the sample size of gun owners respondents (n=947)

    -   **We use this when looking at gun owners.**

-   `weight4` = Post-Stratification Weight for gun HHs

    -   This total respondent weight was then used to scale to respondents with a gun in the household (n=843)

```{r}
clean <-
  raw %>% 
  set_variable_labels(
    weight1 = "Post-stratification weights - 18+ general population",
    weight2 = "Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs",
weight3 = "Post-stratified weights for gun owners",
weight4 = "Post-stratified weights for gun households"
  )
```

Based on [the Barry paper](https://www.sciencedirect.com/science/article/pii/S0091743515001668?via=ihub), we should have 2703 in total, with 843 non-gunowners and 947 gun owners.

```{r}
clean %>% tabyl(gunhomeper) %>% adorn_totals("row")
```

We have more non-gun than the paper.

It looks like they are using the gun owner vs non-gun owner weights to compare gun owners to non gun owners, but the full weight to explore the whole sample.

We can see that by restrict the sample to only the observations with non missing values for the gun owner weight (weight3) and non-gun owner weight (weight4). This matches the numbers in Table 1 of the Barry paper.

```{r}
clean %>%
  filter(
    !is.na(weight3) | !is.na(weight4)
  ) %>% 
  tabyl(gunhomeper) %>% 
  adorn_totals("row")
```

## Check for missing values

```{r}
colSums(is.na(clean))
```

## Explore demographic variables

```{r}
clean %>% 
  mutate(
    g = as.numeric(gunhomeper == "Yes")
  ) %>% 
  summarise(
    across(
      c("weight1", "weight2"), 
      ~ (sum(g*.x, na.rm = T)/sum(.x, na.rm = T)) %>%  round(2) *100
    )
  ) 
```

### Gun owners

```{r}
clean %>% tabyl(sample, gunhomeper)
```

```{r}
demovars <-
  c(
    "age4",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income4",
    "internet",
    "home_type",
    "party7"
  )

summstats <-
  c("notNA(x)", "mean(x)")
summnames <-
  c("Observations", "Mean")
```

```{r, results = "asis", warning = F, message = F}
st(
  filter(
    clean,
    sample == "Gun Owners",
    ),
  vars = demovars,
  group.weights = "weight3",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among gun owners",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

### Full sample

```{r}
demovars <-
  c(
    "gunhomeper",
    "age4",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income4",
    "internet",
    "home_type",
    "party7"
  )
```


```{r, results = "asis", warning = F, message = F}
st(
  filter(clean, sample == "General population"),
  vars = demovars,
  group.weights = "weight1",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among full sample",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2013.rds"
    )
  ) 
```

```{r}
nrow(clean)
```
