# Clean 2013 data

This script cleans the 2013 survey data.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2013.rds

**Output**

-   DemographyGunOwners/data/gunsurvey/2013.rds

**Last ran**

-   01/31/26

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
source <-
  read_dta(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2013.dta"
      )
  ) %>% 
  clean_names() %>% 
  # many variable names have the pp prefix, so remove it
  rename_with(
    ~ str_remove_all(.x, "pp"), 
    everything()
  )
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2013 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-------------------------|-----------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                            |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Select variables

```{r}
demovars <-
  c(
    "age",
    "agecat",
    "agect4",
    "educ",
    "educat",
    "ethm",
    "gender",
    # Household 
    "hhhead",
    "hhsize",
    "t01",
    "t25",
    "t1317",
    "t18ov",
    "core_par",
    "ever_par",
    "marit",
    # Region
    "staten",
    "reg4",
    "reg9",
    "msacat",
    # House type
    "house",
    "rent",
    # Income
    "incimp",
    # Internet
    "net",
    # Employment
    "work",
    "partyid7"
  )

gun <-
  c(
    # Respondent personally owns a gun
    "xpa0122",
    # Respondent has a gun in household but is not owner
    "xpa0117",
    # Type of gun
    "pa0118",
    "pa0119",
    "pa0121",
    "pa0128",
    # NRA
    "q30",
    #Victim of gun violence
    "q31"
  )
```

```{r}
raw <-
  source %>% 
  clean_names() %>% 
  select(
    starts_with(c("caseid", "weight")),
    all_of(c(demovars, gun))
  )
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender =
      factor(
        gender,
        levels = 1:2,
        labels = c("Male", "Female")
      )
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(agecat)
raw %>% tabyl(agect4)
```

```{r}
raw <-
  raw %>% 
  rename(age4 = agect4,
         age7 = agecat) %>% 
  mutate(
    age4 = 
      factor(
        age4,
        levels = 1:4,
        labels =
          c("18-29",
            "30-44",
            "45-59",
            "60+")
      ),
    age7 = 
      factor(
        age7,
        levels = 1:7,
        labels =
          c("18-24",
            "25-34",
            "35-44",
            "45-54",
            "55-64",
            "65-74",
            "75+")
      )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

Note that 2013 data does not include Asian, non-Hispanic as a category.

```{r}
raw %>% tabyl(ethm)
```

```{r}
raw <-
  raw %>% 
  rename(racethnicity = ethm) %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:5,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+, non-Hispanic"
            )
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

```{r}
raw %>% tabyl(educ)
raw %>% tabyl(educat)
```

```{r}
raw <-
  raw %>% 
  mutate(
    educ5 =
      case_when(
       educ %in% 1:8 ~ 1,
       educ == 9 ~ 2,
       educ %in% 10:11 ~ 3,
       educ == 12 ~ 4,
       TRUE ~ 5
       ) %>% 
      factor(
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Vocational/tech school/some college/ associates",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      ),
    educ = 
      factor(
        educ,
        levels = 1:14,
        labels = 
          c("No formal education",
            "1st, 2nd, 3rd, or 4th grade",
            "5th or 6th grade",
            "7th or 8th grade",
            "9th grade",
            "10th grade",
            "11th grade",
            "12th grade (no diploma)",
            "High school graduate, diploma, or GED",
            "Some college, no degree",
            "Associate degree",
            "Bachelors degree",
            "Masters degree",
            "Professional or Doctorate degree")
      ),
    educ4 = 
      factor(
        educ4, 
        levels = 1:4,
        labels = 
          c("No HS diploma",
            "HS graduate or equivalent",
            "Some college",
            "BA or above")
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education",
    educ4 = "4-level highest level of education",
    educ = "Highest level of education"
  )
```

```{r}
walk(
  c("educ", "educ5","educ4"),
  ~ print(raw %>% tabyl(.x))
)
```

## Martial status

```{r}
raw %>% tabyl(marital)
```

```{r}
raw <-
  raw %>% 
  mutate(
    marital = 
      factor(
        marital, 
        levels = 1:6,
        labels = 
          c(
            "Married",
            "Widowed",
            "Divorced",
            "Separated",
            "Never married",
            "Living with partner"
          )
      )
  )  %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(employ)
```

```{r}
raw <-
  raw %>% 
  mutate(
    employ = 
      factor(
        employ, 
        levels = 1:7, 
        labels = 
          c(
            "Working - as a paid employee",
            "Working - self-employed",
            "Not working - on temporary layoff from a job",
            "Not working - looking for work",
            "Not working - retired",
            "Not working - disabled",
            "Not working - other"
          )
      )
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

We construct 2 variables: income4 and income5 to match 2021 and 2023 data.

```{r}
raw %>% tabyl(income)
```

```{r}
raw <-
  raw %>% 
  mutate(
    income4 = 
      case_when(
        income %in% 1:6 ~ 1,
        income %in% 7:10 ~ 2,
        income %in% 11:13 ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      case_when(
        income %in% 1:2 ~ 1,
        income %in% 3:4 ~ 2,
        income %in% 5:6 ~ 3,
        income %in% 7:8 ~ 4,
        income == 9 ~ 5,
        income %in% 10:11 ~ 6,
        income %in% 12:13 ~ 7,
        income %in% 14:15 ~ 8,
        TRUE ~ 9
      ) %>% 
      factor(
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      ),
    income = 
      factor(
        income, 
        levels = 1:18,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000-199,999",
            "$200,000 or more")
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
walk(
    raw %>% select(starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## State

```{r}
raw %>% tabyl(state)
```

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    state = "State"
  )
```

## Region

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    region4 = 
      factor(
        region4, 
        levels = 1:4,
        labels =
          c("Northeast",
            "Midwest",
            "South",
            "West")
      ),
    region9 = 
      factor(
        region9, 
        levels = 1:9,
        labels =
          c("New England",
            "Mid-Atlantic",
            "East North Central",
            "West North Central",
            "South Atlantic",
            "East South Central",
            "West South Central",
            "Mountain",
            "Pacific")
      )
  )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )

```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(msacat)
```

```{r}
raw <-
  raw %>% 
  mutate(
    metro = 
      factor(
        metro, 
        levels = 0:1, 
        labels = c("Non-Metro Area", "Metro Area")
      )
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(internet)
```

```{r}
raw <-
  raw %>% 
  mutate(
    internet = 
      factor(
        internet,
        levels = 0:1,
        labels = c("Non-internet household",
                   "Internet household")
      )
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(housing)
```

```{r}
raw <-
  raw %>% 
  mutate(
    housing = 
      factor(
        housing, 
        levels = 1:3,
        labels = 
            c(
              "Owned or being bought by you or someone in your household",
              "Rented for cash",
              "Occupied without payment of cash rent"
              )
      )
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(home_type)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    home_type = 
      factor(
        home_type, 
        levels = 1:5, 
        labels = 
          c("A one-family house detached from any other house",
            "A one-family house attached to one or more houses",
            "A building with 2 or more apartments",
            "A mobile home or trailer",
            "Boat, RV, van, etc")
      )
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```

## Phone service

```{r}
raw %>% tabyl(phoneservice)
```

```{r}
raw <-
  raw %>% 
  mutate(
    phoneservice = 
      factor(
        phoneservice,
        levels = 1:5,
        labels = 
          c("Landline telephone only",
            "Have a landline, but mostly use cellphone",
            "Have cellphone, but mostly use landline",
            "Cellphone only",
            "No telephone service"
            )
      )
  ) %>% 
  set_variable_labels(
    phoneservice = "Telephone service for the household"
  )
```

```{r}
raw %>% tabyl(phoneservice)
```

## Political party

```{r}
raw %>% tabyl(p_partyid)
```

```{r}
raw <-
  raw %>% 
  rename(party7 = p_partyid) %>% 
  mutate(
    party7 =
      factor(
        as.numeric(party7), 
        levels = c(-1, 1:7),
        labels = 
          c("Unknown",
            "Strong Democrat",
            "Not so strong Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Not so strong Republican",
            "Strong Republican")
      )
  ) %>% 
  set_variable_labels(
    party7 = "7-level political affiliation"
  )
```

```{r}
raw %>% tabyl(party7)
```


## Are you a member of the National Rifle Association--also known as the NRA?  (q30)


## Household size

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh01 = "Number of HH members age 0-1",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+"
  )
```

## Weights

Notice here that the weight numbering here is different to 2021- waves of the survey

-   `weight1` = Post-Stratification Weight for 18+ US adult gen pop sample

-   `weight2` = Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs

-   `weight3` = Post-Stratification Weight for gun owners

-   `weight4` = Post-Stratification Weight for gun HHs

```{r}
clean <-
  raw %>% 
  rename(weight = weight1) %>% 
  set_variable_labels(
    weight = "Post-stratification weights - 18+ general population",
    weight2 = "Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs",
weight3 = "Post-stratified weights - scaled to 2 groups (gun owners vs not gun owners)"
  )
```

## Explore demographic variables

```{r}
demovars <-
  c(
    "age",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type",
    "party7",
    "gunhome"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  clean,
  vars = demovars,
  group.weights = "weight",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2017.rds"
    )
  ) 
```
