# Clean 2015 data

This script cleans the 2015 survey data. This survey is very similar to 2013.

List of variables are different/not available in comparison to later survey years:

-   Different

    -   `income`

    -   `racethnicity`

        -   No level for Asian, non-Hispanic

    -   `hh02` vs `hh01` in later years

-   Not available here but available for other years

    -   `phoneservice`

    -   `nraever`

    -   `party7`

    -   `gunhome` = Do you happen to have in your home or garage any guns or revolvers?

    -   `core_par`

    -   `ever_par`

    -   `hhhead`
    
    - types of guns a person owns

-   Not available in later years

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2015.rds

**Output**

-   DemographyGunOwners/data/gunsurvey/2015.rds

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
source <-
  read_dta(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2015.dta"
      )
  ) %>% 
  clean_names() %>% 
  # many variable names have the pp prefix, so remove it
  rename_with(
    ~ str_remove_all(.x, "pp"), 
    everything()
  )
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2015 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|---------------------------|---------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                            |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Select variables

```{r}
demovars <-
  c(
    "age",
    "agecat",
    "agect4",
    "educ",
    "educat",
    "ethm",
    "gender",
    # Household 
    "hhhead",
    "hhsize",
    "t01",
    "t25",
    "t1317",
    "t18ov",
    "t612",
    "marit",
    # Region
    "staten",
    "reg4",
    "reg9",
    "msacat",
    # House type
    "house",
    "rent",
    # Income
    "incimp",
    # Internet
    "net",
    # Employment
    "work"
  )

gun <-
  c(
    # Respondent personally owns a gun
    "xpa0122",
    # NRA
    "q23"
  )
```

```{r}
raw <-
  source %>% 
  rename(
    caseid = case_id,
    # used to select general population sample
    sample = xgun 
    ) %>%
  select(
    starts_with(c("caseid", "weight", "sample")),
    all_of(c(demovars, gun))
  )
```

##  Sample

This variable is used to restrict the sample to the general population sample or the gun augment sample. We use the first when we are making general population inference.

```{r}
raw %>% tabyl(sample)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    sample = 
      factor(
        sample,
        levels = 1:2,
        labels = 
          c("General population",
            "Gun Owners")
      )) %>% 
  set_variable_labels(
    sample = "Type of Respondents"
  )
```

```{r}
raw %>% tabyl(sample)
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender = droplevels(as_factor(gender))
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(agecat)
raw %>% tabyl(agect4)
```

```{r}
raw <-
  raw %>% 
  rename(age4 = agect4,
         age7 = agecat) %>% 
  mutate(
    across(
      c("age4", "age7"), 
      ~ droplevels(as_factor(.x))
    )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

Note that 2013-2017 data does not include Asian, non-Hispanic as a category.

```{r}
raw %>% tabyl(ethm)
```

```{r}
raw <-
  raw %>% 
  rename(racethnicity = ethm) %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:5,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+ Races, non-Hispanic"
            )
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

We drop `educ` because it corresponds to 4 levels of education, but in order to match later survey years, we define education using 5 levels.

```{r}
raw %>% tabyl(educ)
raw %>% tabyl(educat)
```

```{r, warning = F}
raw <-
  raw %>% 
  mutate(
    educ5 =
      case_when(
       educ %in% 1:8 ~ 1,
       educ == 9 ~ 2,
       educ %in% 10:11 ~ 3,
       educ == 12 ~ 4,
       TRUE ~ 5
       ) %>% 
      factor(
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Some college/associates degree",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      ),
    educ = 
      factor(
        educ,
        levels = 1:14,
        labels = 
          c("No formal education",
            "1st, 2nd, 3rd, or 4th grade",
            "5th or 6th grade",
            "7th or 8th grade",
            "9th grade",
            "10th grade",
            "11th grade",
            "12th grade (no diploma)",
            "HS graduate, diploma, or GED",
            "Some college, no degree",
            "Associate degree",
            "Bachelors degree",
            "Masters degree",
            "Professional or Doctorate degree")
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education",
    educ = "Highest level of education"
  ) %>% 
  select(-"educat")
```

```{r}
walk(
  c("educ", "educ5"),
  ~ print(raw %>% tabyl(.x))
)
```

## Martial status

```{r}
raw %>% tabyl(marit)
```

```{r}
raw <-
  raw %>% 
  rename(marital = marit) %>% 
  mutate(
    marital = droplevels(as_factor(marital))
    ) %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(work)
```

```{r}
raw <-
  raw %>% 
  rename(employ = work) %>% 
  mutate(
    employ = droplevels(as_factor(employ))
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

We construct 2 variables: `income4` and `income5` to match 2021 and 2023 data.

Note that in later years, the income variable splits 175-199,999 and then 200,000 + but here it is 175,000+. We leave this as is because we will likely never use this variable, rather we will use the income4 or income9 variables. This income variable also splits the sample into different categories as the later years. We attempt to harmonize it as much as possible.

```{r}
raw %>% tabyl(incimp)
```

```{r}
raw <-
  raw %>% 
  rename(income = incimp) %>% 
  mutate(
    income4 = 
      case_when(
        income %in% 1:8 ~ 1,
        income %in% 9:12 ~ 2,
        income %in% 13:15 ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      case_when(
        income %in% 1:3 ~ 1,
        income %in% 4:6 ~ 2,
        income %in% 7:8 ~ 3,
        income %in% 9:10 ~ 4,
        income == 11 ~ 5,
        income %in% 12:13 ~ 6,
        income %in% 14:15 ~ 7,
        income %in% 16:17 ~ 8,
        TRUE ~ 9
      ) %>% 
      factor(
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      ),
    income = 
      case_when(
        income == 1 ~ 1,
        income %in% 2:3 ~ 2,
        income %in% 4:5 ~ 3,
        TRUE ~ income - 2
      ) %>% 
      factor(
        levels = 1:17,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000 or more"
            )
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
walk(
    raw %>% select(starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## State

```{r}
raw %>% tabyl(staten)
```

```{r}
raw <-
  raw %>%
  rename(state = staten) %>% 
  mutate(state = droplevels(as_factor(state))) %>% 
  set_variable_labels(state = "State")
```

```{r}
raw %>% tabyl(state)
```

## Region

```{r}
raw %>% tabyl(reg4)
raw %>% tabyl(reg9)
```

```{r}
raw <-
  raw %>% 
  rename(region4 = reg4, region9 = reg9) %>% 
  mutate(
    across(starts_with("region"), ~ droplevels(as_factor(.x)))
    )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )
```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(msacat)
```

```{r}
raw <-
  raw %>% 
  rename(metro = msacat) %>% 
  mutate(
    metro = as_factor(metro)
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(net)
```

```{r}
raw <-
  raw %>% 
  rename(internet = net) %>% 
  mutate(
    internet = 
      factor(
        internet,
        levels = 0:1,
        labels = c("Non-internet household",
                   "Internet household")
      )
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(rent)
```

```{r}
raw <-
  raw %>% 
  rename(housing = rent) %>% 
  mutate(
    housing = droplevels(as_factor(housing))
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(house)
```

```{r}
raw <- 
  raw %>% 
  rename(home_type = house) %>% 
  mutate(
    home_type = droplevels(as_factor(home_type))
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```

## Are you a member of the National Rifle Association--also known as the NRA? (Q23)

```{r}
raw %>% tabyl(q23)
```

```{r}
raw <-
  raw %>% 
  rename(nranow = q23) %>% 
  mutate(
    nranow = 
      factor(
        nranow,
        levels = c(1:2, -1),
        labels =
           c("Yes", "No", "Unknown")
      )
    ) %>% 
  set_variable_labels(
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow)
```

## Gun ownership

-   Respondent personally owns a gun (`xpa0122`)

```{r}
raw %>% tabyl(xpa0122)
```

```{r}
raw <-
  raw %>% 
  rename(
    gunhomeper = xpa0122,
    ) %>% 
  mutate(
    gunhomeper = 
      factor( 
        gunhomeper,
        levels = c(1:2),
        labels =
          c("Yes", "No")
        )
    ) %>% 
  set_variable_labels(
    gunhomeper = "Respondent personally owns a gun"
    ) 
```

```{r}
raw %>% tabyl(gunhomeper)
```

## Household variables

### Household size

We have

-   `t01` = Presence of Household Members - Children 0-2

In later survey years we have

-   `hh01` = age 0-1

This difference is okay because we will define 2 grouped variables in `9-merge.Rmd`

1.  `hh06` = age 0-6

2.  hh18un = Under 18

```{r}
raw %>% tabyl(hhsize)
```

```{r}
raw <-
  raw %>% 
  rename(
    hh02 = t01,
    hh25 = t25,
    hh612 = t612,
    hh1317 = t1317,
    hh18ov = t18ov,
  ) %>% 
  rowwise() |> 
  mutate(hh18un = sum(c(hh02, hh25, hh612, hh1317))) |> 
  ungroup() |> 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh02 = "Number of HH members age 0-2",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+",
    hh18un = "Number of HH members under 18"
  ) 
```


```{r}
colSums(is.na(raw))
```

```{r}
hist(raw$hh18un)
```

```{r}
nrow(filter(raw, hh18un < 0) ) == 0
```

### Parental status

```{r}
raw %>% tabyl(hhhead)
```

```{r}
raw <-
  raw %>% 
  rename(hhead = hhhead) %>% 
  mutate(
    hhead = droplevels(as_factor(hhead))
  ) %>% 
  set_variable_labels(
    hhead = "Is respondent the household head?"
  )
```

```{r}
raw %>% tabyl(hhead)
```

## Weights

All respondents (general population and gun owner augment) were weighted together to look like the age 18+ U.S. population  using the categories above, with demographics controlled within gun owners and non-owners. The benchmarks for gun owners and non-owners were derived from weighted KnowledgePanel data. 

-   `weight1` = Post-Stratification Weight for 18+ US adult gen pop sample.

    -   For respondents in the genpop sample (n=1454)

-   `weight2` = Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs

    -   The weights were then trimmed and scaled to sum to the sample size of the total respondents (weight2; n=1326)

    -   We use this when looking at the full sample

-   `weight3` = Post-Stratification Weight for gun owners

    -   This total respondent weight was then used to scale to the sample size ofb y gun owners and non-owners (weight3; n=524 gun-owners and n=802 non-owners)

    -   **We use this when looking at gun owners.**
    
There are 25 rows with missing data on `weight3`. According to the documentation, the sample size of the total respondents is the number of observations in the data set minus 25 = 1326. So we drop these observations.

```{r}
clean <-
  raw %>% 
  filter(!is.na(weight2)) %>% 
  set_variable_labels(
    weight1 = "Post-stratification weights - 18+ general population",
    weight2 = "Post-Stratification Weight for total sample, 18+ US adults with no guns in HH, gun owners, and guns in HHs",
weight3 = "Post-stratified weights for gun owners"
  )
```

The final sample size should be 1326, with 802 non-gun owners and 524 gun owners.

```{r}
clean %>% tabyl(gunhomeper) %>% adorn_totals("row")
```

## Check for missing values

```{r}
colSums(is.na(clean))
```

## Explore demographic variables

### Gun owners

```{r}
demovars <-
  c(
    "age4",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income9",
    "internet",
    "home_type"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  filter(clean, sample == "Gun Owners"),
  vars = demovars,
  group.weights = "weight3",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among gun owners",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

### Full sample

```{r}
demovars <-
  c(
    "gunhomeper",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income9",
    "internet",
    "home_type"
  )
```

```{r, results = "asis", warning = F, message = F}
st(
  filter(clean, sample == "General population"),
  vars = demovars,
  group.weights = "weight1",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among full sample",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2015.rds"
    )
  ) 
```

```{r}
nrow(clean)
```
