# Clean 2017 data

This script cleans the 2017 survey data.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2017.rds

**Output**

-   DemographyGunOwners/data/gunsurvey/2017.rds

**Last ran** 

- 12/13/2025

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "weights",
    "kableExtra",
    "survey",
    "skimr",
    "psych",
    "dplyr",
    "here",
    "vtable",
    "fixest",
    "lfe",
    "broom"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
source <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2017.rds"
      )
  ) %>% 
  clean_names()
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2017 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name | Value |
|------------------------|-----------------------------------------------|
| Number of observations | `r nrow(raw)` |
| Number of unique cases | `r n_distinct(source$case_id)` |
| Sum of weights | `r sum(``source``$weight1)` |
| Number of variables | `r length(``source``)` |
| Variable names | `r``source``%>% names() %>% paste(collapse = ", ")` |

## Clean names and remove Stata encoding

```{r}
raw <-
  source %>% 
  mutate(
    across(
      where(is.havenlab), ~ factor(haven::as_factor(.x))
    )
  )
```

```{r}
demovars <-
  c(
    "age",
    "hhsize",
    "gender",
    "raceethnicity",
    "educ",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  raw,
  vars = demovars,
  group.weights = "weight1",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary",
  out = "kable",
  numformat = "comma"
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

## Data quality

### Duration of survey

Figure \@ref(fig:durhist) shows that over half of survey respondents
have a survey duration of greater than 3 hours (180 minutes), which seems suspicious. 

```{r}
summary(raw$duration)
```


```{r durhist, fig.cap="Distribution of survey duration",  fig.topcaption=TRUE}
raw %>% 
  mutate(
    duration = 
      ifelse(duration > 180, 180, duration)
    ) %>% 
  ggplot( 
    aes(x = duration, weight = weight1)
    ) +
  geom_histogram(
    aes(y = after_stat(count / sum(count))),
    bins = 150,
    boundary = 0,
    closed = "left"
  ) +
  scale_y_continuous(
    labels = scales::percent
    ) +
  scale_x_continuous(n.breaks = 12) +
  theme +
  labs(
    x = "Weighted distribution of survey duration (in minutes)",
    y = "Weighted percent of respondents"
  )
```

Age is positively correlated with duration of survey.

```{r}
felm(
  duration ~ gender +  income + age | state,
  data = raw
)
```

```{r}
t.test(duration ~ surv_mode, data = raw )

```

```{r}
raw %>% 
  group_by(surv_mode) %>% 
  summarise(
    mean = sum(weight1*duration)/sum(weight1)
  )
```
