# Clean 2017 data

This script cleans the 2017 survey data.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2017.rds

**Output**

-   DemographyGunOwners/data/gunsurvey/2017.rds

**Last ran** 

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```


## Import data

```{r}
source <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2017.rds"
      )
  ) %>% 
  clean_names()
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2017 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-------------------------|-----------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                           |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Select variables

```{r}
demovars <-
  c(
    "age",
    "age4",
    "age7",
    "gender",
    "racethnicity",
    "party_id7",
    "educ",
    "educ4",
    "marital",
    "employ",
    "income",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov"
  )

gun <-
  c(
    "q29",
    "q30",
    "q31"
  )
```

```{r}
raw <-
  source %>% 
  clean_names() %>% 
  select(
    starts_with(c("caseid", "weight")),
    all_of(c(demovars, gun))
  )
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender = droplevels(as_factor(gender))
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

```{r}
raw <-
  raw %>% 
   mutate(
    across(
      c("age4", "age7"), 
      ~ droplevels(as_factor(.x))
    )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

```{r}
raw %>% tabyl(racethnicity)
```

```{r}
raw <-
  raw %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:5,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+ Races, non-Hispanic"
            )
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

We construct a new variable `educ5` using `educ` which matches the variable from 2021 and 2023.

```{r}
raw %>% tabyl(educ)
raw %>% tabyl(educ4)
```

```{r}
raw <-
  raw %>% 
  mutate(
    educ5 =
      case_when(
       educ %in% 1:8 ~ 1,
       educ == 9 ~ 2,
       educ %in% 10:11 ~ 3,
       educ == 12 ~ 4,
       TRUE ~ 5
       ) %>% 
      factor(
        levels = 1:5,
        labels = 
           c("Less than HS",
            "HS graduate",
            "Some college/associates degree",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      ),
    educ = 
      factor(
        educ,
        levels = 1:14,
        labels = 
          c("No formal education",
            "1st, 2nd, 3rd, or 4th grade",
            "5th or 6th grade",
            "7th or 8th grade",
            "9th grade",
            "10th grade",
            "11th grade",
            "12th grade (no diploma)",
            "HS graduate, diploma, or GED",
            "Some college, no degree",
            "Associate degree",
            "Bachelors degree",
            "Masters degree",
            "Professional or Doctorate degree"
            )
        ),
    educ4 = 
      factor(
        educ4, 
        levels = 1:4,
        labels = 
          c("No HS diploma",
            "HS graduate or equivalent",
            "Some college",
            "BA or above")
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education",
    educ4 = "4-level highest level of education",
    educ = "Highest level of education"
  )
```

```{r}
walk(
  c("educ", "educ5","educ4"),
  ~ print(raw %>% tabyl(.x))
)
```

## Martial status

```{r}
raw %>% tabyl(marital)
```

```{r}
raw <-
  raw %>% 
  mutate(
    marital = droplevels(as_factor(marital))
    ) %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(employ)
```

```{r}
raw <-
  raw %>% 
  mutate(
    employ = droplevels(as_factor(employ))
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

We construct 2 variables: income4 and income5 to match 2021 and 2023 data.

```{r}
raw %>% tabyl(income)
```

```{r}
raw <-
  raw %>% 
  mutate(
    income4 = 
      case_when(
        income %in% 1:6 ~ 1,
        income %in% 7:10 ~ 2,
        income %in% 11:13 ~ 3,
        TRUE ~ 4
      ) %>% 
      factor(
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      case_when(
        income %in% 1:2 ~ 1,
        income %in% 3:4 ~ 2,
        income %in% 5:6 ~ 3,
        income %in% 7:8 ~ 4,
        income == 9 ~ 5,
        income %in% 10:11 ~ 6,
        income %in% 12:13 ~ 7,
        income %in% 14:15 ~ 8,
        TRUE ~ 9
      ) %>% 
      factor(
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      ),
    income = 
      factor(
        income, 
        levels = 1:18,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000-199,999",
            "$200,000 or more")
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
walk(
    raw %>% select(starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## State

```{r}
raw %>% tabyl(state)
```

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    state = "State"
  )
```

## Region

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    across(starts_with("region"), ~ droplevels(as_factor(.x)))
    )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )
```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(metro)
```

```{r}
raw <-
  raw %>% 
  mutate(
    metro = as_factor(metro)
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(internet)
```

```{r}
raw <-
  raw %>% 
  mutate(
    internet = as_factor(internet)
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(housing)
```

```{r}
raw <-
  raw %>% 
  mutate(housing = as_factor(housing)) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(home_type)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    home_type = as_factor(home_type)
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```

## Phone service

```{r}
raw %>% tabyl(phoneservice)
```

```{r}
raw <-
  raw %>% 
  mutate(
    phoneservice = as_factor(phoneservice)
  ) %>% 
  set_variable_labels(
    phoneservice = "Telephone service for the household"
  )
```

```{r}
raw %>% tabyl(phoneservice)
```


## Political party

We construct `party5`
```{r}
raw %>% tabyl(party_id7)
```

```{r}
raw <-
  raw %>% 
  rename(party7 = party_id7) %>% 
  mutate(
    party5 =
      case_when(
        party7 < 1 | party7 == 8 ~ -1,
        party7 %in% 1:2 ~ 1,
        party7 == 3 ~ 2,
        party7 == 4 ~ 3,
        party7 == 5 ~ 4,
        TRUE ~ 5
      ) %>% 
      factor(
        levels = c(1:5,-1),
        labels = 
          c("Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Republican",
            "Unknown"
            )
      ),
    party7 =
      ifelse(party7 == 8, -1, party7) %>% 
      factor(
        levels = c(1:7, -1),
        labels = 
          c(
            "Strong Democrat",
            "Not so strong Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Not so strong Republican",
            "Strong Republican",
            "Unknown"
            )
      )
  ) %>% 
  set_variable_labels(
    party7 = "7-level political affiliation",
    party5 = "5-level political affiliation"
  )
```

```{r}
raw %>% tabyl(party7)
```

## Are you a member of the National Rifle Association (Q29)

Are you a member of the National Rifle Association--also known as the NRA? = `q29` = `nranow`

```{r}
raw %>% tabyl(q29)
```

```{r}
raw <-
  raw %>% 
  rename(nranow = q29) %>% 
  mutate(
    nranow = as.numeric(nranow),
    nranow =
      ifelse(nranow > 2, 99, nranow) %>% 
      factor(
        levels = c(1:2, 99),
        labels =
          c("Yes", "No", "Unknown")
      )
  ) %>% 
  set_variable_labels(
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow)
```

## Gun ownership (Q30 & 31)

-   Do you happen to have in your home or garage any guns or revolvers? = `q30` = `gunhome`

Do any of these guns personally belong to you? = `gunhomeper`  = `q31`

```{r}
raw %>% tabyl(q30)
raw %>% tabyl(q31)
```

```{r}
raw <-
  raw %>% 
  rename(
    gunhome = q30,
    gunhomeper = q31
    ) %>% 
  mutate(
    gunhomeper = 
      case_when(
       is.na(gunhomeper) ~  as.character(gunhome), 
        TRUE ~ as.character(gunhomeper)
      ) %>% as.numeric,
    across(
      starts_with("gunhome"),
      ~ ifelse(.x > 2, 99, .x) %>% 
        factor(
          levels = c(1:2, 99),
          labels =
            c("Yes", "No", "Unknown")
        )
      )
  ) %>% 
  set_variable_labels(
    gunhome = "Do you happen to have in your home or garage any guns or revolvers?",
    gunhomeper = "Do any of guns or revolvers in your home or garagepersonally belong to you?"
  )
```

```{r}
raw %>% tabyl(gunhome, gunhomeper)
```
## Household size

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh01 = "Number of HH members age 0-1",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+"
  )
```

## Weights

Unlike 2021, and 2023, the only weight available is "Post-stratification weights - 18+ general population (N=2,124)". 

I am unsure if this can be used for analysis among gun owners.

*Or I have to define my own weights.*

```{r}
clean <-
  raw %>% 
  rename(weight3 = weight1) %>% 
  set_variable_labels(
    weight3 = "Post-stratification weights - 18+ general population (N=2,124)"
  )
```

## Check for missing values

```{r}
colSums(is.na(clean))
```

## Explore demographic variables

### Gun owners

```{r}
demovars <-
  c(
    "age4",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  clean %>% filter(gunhomeper == "Yes"),
  vars = demovars,
  group.weights = "weight3",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among gun owners",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

### Full sample

```{r}
demovars <-
  c(
    "gunhomeper",
    "age4",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type"
  )
```

```{r, results = "asis", warning = F, message = F}
st(
  clean,
  vars = demovars,
  group.weights = "weight3",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among full sample",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2017.rds"
    )
  ) 
```




