# Clean 2023 data

We use this [codebook.](https://livejohnshopkins-my.sharepoint.com/:x:/r/personal/jdada3_jh_edu/Documents/Research/DemographyGunOwners/documentation/9558_JHU%20Gun%20Policy%202023_codebook.xlsx?d=w551a28b764fc4672a2bfc152b24b6e04&csf=1&web=1&e=4Jb9Vt).

Note that all of the demographic variables are the same as the 2021 survey.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2023.rds

**Output**

-   DemographyGunOwners/data/clean/gs-2023.rds

**Last ran**

-   01/27/26

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "haven",
    "labelled",
    "tidyr",
    "janitor",
    "weights",
    "conflicted",
    "kableExtra",
    "skimr",
    "purrr",
    "dplyr",
    "here",
    "vtable",
    "broom"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)

conflicts_prefer(dplyr::filter)
```

## Import data

```{r}
source <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2023.rds"
      )
  ) %>% 
  clean_names()
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2023 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-------------------------|-----------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight1)`                           |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Drop order variables

In the survey, the questions are ordered randomly. We remove the variables that denote the order of the questions.

```{r}
raw <- 
  source %>% 
  select(
    -c(
      matches("order"),
      "rnd_01"
      )
  )
```

## Select variables

We select demographic variables and a handful of the questions that pertain to a political identity in order to characterize the typical gun owner.

```{r}
demovars <-
  c(
    "age",
    "age4",
    "age7",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "income4",
    "income9",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov",
    "veteran_1",
    "veteran2_1"
  )

politics <-
  c(
    "party_id7",
    "party_id5",
    "qcandi20",
    "qvotediff",
    "q_vote22",
    "qvotediff2"
  )

gun <-
  c(
    "gun",
    "qguntypea",
    "qguntypeb",
    "qguntypec",
    "qguntyped", 
    "q39a",
    "q39b",
    "q40_23",
    "q40",
    "q41",
    "q41_23",
    "q42_23",
    "q43_23a",
    "q43_23b",
    "q43_23c",
    "q47_23"
  )
```

```{r}
raw <-
  source %>% 
  dplyr::select(
    starts_with(c("caseid", "weight")), 
    all_of(c(demovars, politics, gun))
  )
```

## Gun

Proportion-based simple random imputation used to determine gun ownership for those with missing data. Used for weighting purposes.

```{r}
raw %>% tabyl(gun)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gun = 
      factor(
        gun,
        levels = 1:2,
        labels = 
          c("Gun owner",
            "Non-gun owner")
      )
  ) %>% 
  set_variable_labels(
    gun = "Proportion-based simple random imputation used to determine gun ownership for those with missing data. Used for weighting purposes"
  )
```

```{r}
raw %>% tabyl(gun)
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender = 
      factor(
        gender, 
        levels = 1:2,
        labels = c("Male", "Female")
      )
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

```{r}
raw <-
  raw %>% 
  mutate(
    age4 = 
      factor(
        age4,
        levels = 1:4,
        labels =
          c("18-29",
            "30-44",
            "45-59",
            "60+")
      ),
    age7 = 
      factor(
        age7,
        levels = 1:7,
        labels =
          c("18-24",
            "25-34",
            "35-44",
            "45-54",
            "55-64",
            "65-74",
            "75+")
      )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

```{r}
raw %>% tabyl(racethnicity)
```

```{r}
raw <-
  raw %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:6,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+, non-Hispanic",
            "Asian, non-Hispanic")
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

```{r}
raw %>% tabyl(educ5)
```

```{r}
raw <-
  raw %>% 
  mutate(
    educ5 = 
      factor(
        educ5,
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Vocational/tech school/some college/ associates",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education"
  )
```

```{r}
raw %>% tabyl(educ5)
```

## Martial status

```{r}
raw %>% tabyl(marital)
```

```{r}
raw <-
  raw %>% 
  mutate(
    marital = 
      factor(
        marital, 
        levels = 1:6,
        labels = 
          c(
            "Married",
            "Widowed",
            "Divorced",
            "Separated",
            "Never married",
            "Living with partner"
          )
      )
  )  %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(employ)
```

```{r}
raw <-
  raw %>% 
  mutate(
    employ = 
      factor(
        employ, 
        levels = 1:7, 
        labels = 
          c(
            "Working - as a paid employee",
            "Working - self-employed",
            "Not working - on temporary layoff from a job",
            "Not working - looking for work",
            "Not working - retired",
            "Not working - disabled",
            "Not working - other"
          )
      )
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

```{r}
raw %>% tabyl(income)
raw %>% tabyl(income4)
raw %>% tabyl(income9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    income = 
      factor(
        income, 
        levels = 1:18,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000-199,999",
            "$200,000 or more")
      ),
    income4 = 
      factor(
        income4, 
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      factor(
        income9, 
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
raw %>% tabyl(income)
raw %>% tabyl(income4)
raw %>% tabyl(income9)
```

## State

```{r}
raw %>% tabyl(state)
```

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    state = "State"
  )
```

## Region

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    region4 = 
      factor(
        region4, 
        levels = 1:4,
        labels =
          c("Northeast",
            "Midwest",
            "South",
            "West")
      ),
    region9 = 
      factor(
        region9, 
        levels = 1:9,
        labels =
          c("New England",
            "Mid-Atlantic",
            "East North Central",
            "West North Central",
            "South Atlantic",
            "East South Central",
            "West South Central",
            "Mountain",
            "Pacific")
      )
  )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )

```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(metro)
```

```{r}
raw <-
  raw %>% 
  mutate(
    metro = 
      factor(
        metro, 
        levels = 0:1, 
        labels = c("Non-Metro Area", "Metro Area")
      )
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(internet)
```

```{r}
raw <-
  raw %>% 
  mutate(
    internet = 
      factor(
        internet,
        levels = 0:1,
        labels = c("Non-internet household",
                   "Internet household")
      )
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(housing)
```

```{r}
raw <-
  raw %>% 
  mutate(
    housing = 
      factor(
        housing, 
        levels = 1:3,
        labels = 
            c(
              "Owned or being bought by you or someone in your household",
              "Rented for cash",
              "Occupied without payment of cash rent"
              )
      )
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(home_type)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    home_type = 
      factor(
        home_type, 
        levels = 1:5, 
        labels = 
          c("A one-family house detached from any other house",
            "A one-family house attached to one or more houses",
            "A building with 2 or more apartments",
            "A mobile home or trailer",
            "Boat, RV, van, etc")
      )
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```

## Phone service

```{r}
raw %>% tabyl(phoneservice)
```

```{r}
raw <-
  raw %>% 
  mutate(
    phoneservice = 
      factor(
        phoneservice,
        levels = 1:5,
        labels = 
          c("Landline telephone only",
            "Have a landline, but mostly use cellphone",
            "Have cellphone, but mostly use landline",
            "Cellphone only",
            "No telephone service"
            )
      )
  ) %>% 
  set_variable_labels(
    phoneservice = "Telephone service for the household"
  )
```

```{r}
raw %>% tabyl(phoneservice)
```

## Gun ownership type

Do you personally own any of the following types of guns?

-   `qguntypea` = handguns

-   `qguntypeb` = rifles

-   `qguntypec` = shotguns

-   `qguntyped` = other

```{r}
walk(
  c("qguntypea", "qguntypeb", "qguntypec", "qguntyped"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <- 
  raw %>% 
  rename(
    ownhandgun = qguntypea,
    ownrifle = qguntypeb, 
    ownshot = qguntypec,
    ownother = qguntyped
  ) %>% 
  mutate(
    across(
      starts_with("own"),
      ~ case_when(
        is.na(.x) ~ 100,
        TRUE ~ .x
      ) %>% 
        factor(
          levels = c(1:2, 77, 98:100),
          labels =
            c("Yes", 
              "No",
              "Don't know",
              "Skipped on web",
              "Refused",
              "Non-gun owner")
        )
    )
  ) %>% 
    set_variable_labels(
      ownhandgun = "Do you personally own any hanguns?",
      ownrifle = "Do you personally own any rifles?", 
      ownshot = "Do you personally own any shotguns?",
      ownother = "Do you personally own any other types of guns?",
    )
```


```{r}
walk(
    raw %>% select(starts_with("own")) %>% names(),
  ~ print(raw %>% tabyl(.x))
)
```


## Q39a and b

Are you a member of the National Rifle Association--also known as the NRA? = `q39a` = `nranow`

Have you ever been a member of the National Rifle Association--also known as the NRA?  = `q39b` = `nraever`

There are `r nrow(filter(raw, q39b =="NA"))` observations with NA hardcoded for `q39b`. 
This corresponds to people who said in q39a that they are current members of the 
NRA or skipped/refused to answer q39a.

```{r}
raw %>% tabyl(q39a)
raw %>% tabyl(q39b)
raw %>% tabyl(q39a, q39b)
```

```{r, warning = F}
raw <-
  raw %>% 
  rename(
    nranow = q39a,
    nraever = q39b
    ) %>% 
  mutate(
    nraever = 
      case_when(
        nraever == "NA" ~ as.numeric(nranow),
        TRUE ~ as.numeric(nraever)
      ),
    across(
      starts_with("nra"),
      ~ factor(
        as.numeric(.x), 
        levels = c(1:2, 98:99),
        labels =
          c("Yes", "No", "Skipped on web", "Refused")
      )
    )
  ) %>% 
  set_variable_labels(
    nraever = "Have you ever been a member of the NRA?",
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow, nraever)
```
