# Clean 2023 data

We use this [codebook](https://livejohnshopkins-my.sharepoint.com/:x:/r/personal/jdada3_jh_edu/Documents/Research/DemographyGunOwners/documentation/9558_JHU%20Gun%20Policy%202023_codebook.xlsx?d=w551a28b764fc4672a2bfc152b24b6e04&csf=1&web=1&e=4Jb9Vt).

Note that all of the demographic variables are the same as the 2021 survey.

According to this [source](https://www.pewresearch.org/short-reads/2024/07/24/key-facts-about-americans-and-guns/), 32\% national gun ownership rates.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2023.rds

**Output**

-   DemographyGunOwners/data/clean/gs-2023.rds

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
packages <-
  c(
    "tidyverse",
    "haven",
    "labelled",
    "tidyr",
    "janitor",
    "weights",
    "conflicted",
    "kableExtra",
    "skimr",
    "purrr",
    "dplyr",
    "here",
    "vtable",
    "broom"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
source <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2023.rds"
      )
  ) %>% 
  clean_names()
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2023 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-----------------------|------------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                            |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Drop order variables

In the survey, the questions are ordered randomly. We remove the variables that denote the order of the questions.

```{r}
raw <- 
  source %>% 
  select(
    -c(
      matches("order"),
      "rnd_01"
      )
  )
```

## Select variables

We select demographic variables and a handful of the questions that pertain to a political identity in order to characterize the typical gun owner.

```{r}
demovars <-
  c(
    "age",
    "age4",
    "age7",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "income4",
    "income9",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov",
    "veteran_1",
    "veteran2_1"
  )

politics <-
  c(
    "party_id7",
    "party_id5",
    "qcandi20",
    "qvotediff",
    "q_vote22",
    "qvotediff2"
  )

gun <-
  c(
    "gun",
    "qguntypea",
    "qguntypeb",
    "qguntypec",
    "qguntyped", 
    "q39a",
    "q39b",
    "q40_23",
    "q40",
    "q41",
    "q41_23",
    "q42_23",
    "q43_23a",
    "q43_23b",
    "q43_23c",
    "q47_23"
  )
```

```{r}
raw <-
  source %>% 
  dplyr::select(
    starts_with(c("case_id", "weight")), 
    all_of(c(demovars, politics, gun))
  ) %>% 
  rename(caseid = case_id)
```

## Gun
Proportion-based simple random imputation used to determine gun ownership for those with missing data. Used for weighting purposes, however, all people who are gunowners (i.e., `gunhomper == 1`) are also identified as gun owners here. So we drop this variable
to remove confusion.

```{r}
raw %>% tabyl(gun)
```

```{r}
raw <-
  raw %>% 
  select(-gun)
```

## Gender

```{r}
raw %>% tabyl(gender)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gender = 
      factor(
        gender, 
        levels = 1:2,
        labels = c("Male", "Female")
      )
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender"
  )
```

```{r}
raw %>% tabyl(gender)
```

## Age

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

```{r}
raw <-
  raw %>% 
  mutate(
    age4 = 
      factor(
        age4,
        levels = 1:4,
        labels =
          c("18-29",
            "30-44",
            "45-59",
            "60+")
      ),
    age7 = 
      factor(
        age7,
        levels = 1:7,
        labels =
          c("18-24",
            "25-34",
            "35-44",
            "45-54",
            "55-64",
            "65-74",
            "75+")
      )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

```{r}
raw %>% tabyl(racethnicity)
```

```{r}
raw <-
  raw %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:6,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+ Races, non-Hispanic",
            "Asian, non-Hispanic")
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

```{r}
raw %>% tabyl(educ5)
```

```{r}
raw <-
  raw %>% 
  mutate(
    educ5 = 
      factor(
        educ5,
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Some college/associates degree",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education"
  )
```

```{r}
raw %>% tabyl(educ5)
```

## Martial status

```{r}
raw %>% tabyl(marital)
```

```{r}
raw <-
  raw %>% 
  mutate(
    marital = 
      factor(
        marital, 
        levels = 1:6,
        labels = 
          c(
            "Married",
            "Widowed",
            "Divorced",
            "Separated",
            "Never married",
            "Living with partner"
          )
      )
  )  %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(employ)
```

```{r}
raw <-
  raw %>% 
  mutate(
    employ = 
      factor(
        employ, 
        levels = 1:7, 
        labels = 
          c(
            "Working - as a paid employee",
            "Working - self-employed",
            "Not working - on temporary layoff from a job",
            "Not working - looking for work",
            "Not working - retired",
            "Not working - disabled",
            "Not working - other"
          )
      )
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

```{r}
raw %>% tabyl(income)
raw %>% tabyl(income4)
raw %>% tabyl(income9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    income = 
      factor(
        income, 
        levels = 1:18,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000-199,999",
            "$200,000 or more")
      ),
    income4 = 
      factor(
        income4, 
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      factor(
        income9, 
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
raw %>% tabyl(income)
raw %>% tabyl(income4)
raw %>% tabyl(income9)
```

## State

```{r}
raw %>% tabyl(state)
```

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    state = "State"
  )
```

## Region

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

```{r}
raw <-
  raw %>% 
  mutate(
    region4 = 
      factor(
        region4, 
        levels = 1:4,
        labels =
          c("Northeast",
            "Midwest",
            "South",
            "West")
      ),
    region9 = 
      factor(
        region9, 
        levels = 1:9,
        labels =
          c("New England",
            "Mid-Atlantic",
            "East North Central",
            "West North Central",
            "South Atlantic",
            "East South Central",
            "West South Central",
            "Mountain",
            "Pacific")
      )
  )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )

```

```{r}
raw %>% tabyl(region4)
raw %>% tabyl(region9)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(metro)
```

```{r}
raw <-
  raw %>% 
  mutate(
    metro = 
      factor(
        metro, 
        levels = 0:1, 
        labels = c("Non-Metro Area", "Metro Area")
      )
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(internet)
```

```{r}
raw <-
  raw %>% 
  mutate(
    internet = 
      factor(
        internet,
        levels = 0:1,
        labels = c("Non-internet household",
                   "Internet household")
      )
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(housing)
```

```{r}
raw <-
  raw %>% 
  mutate(
    housing = 
      factor(
        housing, 
        levels = 1:3,
        labels = 
            c(
              "Owned or being bought by you or someone in your household",
              "Rented for cash",
              "Occupied without payment of cash rent"
              )
      )
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(home_type)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    home_type = 
      factor(
        home_type, 
        levels = 1:5, 
        labels = 
          c("A one-family house detached from any other house",
            "A one-family house attached to one or more houses",
            "A building with 2 or more apartments",
            "A mobile home or trailer",
            "Boat, RV, van, etc")
      )
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```

## Phone service

```{r}
raw %>% tabyl(phoneservice)
```

```{r}
raw <-
  raw %>% 
  mutate(
    phoneservice = 
      factor(
        phoneservice,
        levels = 1:5,
        labels = 
          c("Landline telephone only",
            "Have a landline, but mostly use cellphone",
            "Have cellphone, but mostly use landline",
            "Cellphone only",
            "No telephone service"
            )
      )
  ) %>% 
  set_variable_labels(
    phoneservice = "Telephone service for the household"
  )
```

```{r}
raw %>% tabyl(phoneservice)
```


## Are you a member of the National Rifle Association (Q39a and b)

-   Are you a member of the National Rifle Association--also known as the NRA? = `q39a` = `nranow`

-   Have you ever been a member of the National Rifle Association--also known as the NRA? = `q39b` = `nraever`

There are some observations with NA hardcoded for `q39b`. This corresponds to people who said in q39a that they are current members of the NRA or skipped/refused to answer q39a.

I group together  "Skipped on web", "Refused" as "Unknown"

```{r}
raw %>% tabyl(q39a)
raw %>% tabyl(q39b)
raw %>% tabyl(q39a, q39b)
```

```{r, warning = F}
raw <-
  raw %>% 
  rename(
    nranow = q39a,
    nraever = q39b
    ) %>% 
  mutate(
    nraever = 
      case_when(
        is.na(nraever) ~ as.character(nranow),
        TRUE ~ as.character(nraever)
        ) %>% as.numeric,
    across(
      starts_with("nra"),
      ~ case_when(
          .x == 98 ~ 99,
          TRUE ~ .x
          ) %>% 
        as.numeric %>% 
        factor( 
          levels = c(1:2, 99),
          labels =
            c("Yes", "No", "Unknown")
          )
    )
  ) %>% 
  set_variable_labels(
    nraever = "Have you ever been a member of the NRA?",
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow, nraever)
```

## Are you a member of any other gun rights groups (Q40_23, Q41_23)

`Q40_23`= Are you a member of any other gun rights groups such as Gun Owners of America? = `goanow` i.e., GOA = gun owners of america

`Q41_23` = Are you a member of any gun safety groups such as Moms Demand Action or Everytown for Gun Safety? = `mdanow` = MDA = Moms demand action

```{r}
raw %>% tabyl(q40_23)
raw %>% tabyl(q41_23)
```

```{r}
raw <-
  raw %>% 
  rename(goanow = q40_23, mdanow = q41_23) %>% 
  mutate(
    across(
      c(goanow, mdanow),
      ~ factor(
        .x, 
        levels = c(1:2, 98),
        labels = c("Yes", "No", "Unknown") #"Skipped on web" recoded to unknown
      )
    )
  ) %>% 
  set_variable_labels(
    goanow = "Are you a member of any other gun rights groups such as Gun Owners of America?",
    mdanow = "Are you a member of any gun safety groups such as Moms Demand Action or Everytown for Gun Safety?"
  )
```

```{r}
raw %>% tabyl(goanow)
raw %>% tabyl(mdanow)
```

## Gun ownership (Q40 & 41)

-   Do you happen to have in your home or garage any guns or revolvers? = `q40` = `gunhome`

Do any of these guns personally belong to you? = `gunhomeper`. There are some observations with NA hardcoded for `q41`. This corresponds to people who said answered no to q40.

I group together  "Skipped on web", "Refused" as "No" for the definition of gun ownership, which matches what the [Gun center reported](https://publichealth.jhu.edu/center-for-gun-violence-solutions/americans-agree-on-effective-gun-policy-more-than-were-led-to-believe).

```{r}
raw %>% tabyl(q40)
raw %>% tabyl(q41)
```

```{r}
raw <-
  raw %>% 
  rename(
    gunhome = q40,
    gunhomeper = q41
    ) %>% 
  mutate(
    gunhomeper = 
      case_when(
        gunhomeper > 2 | is.na(gunhomeper) ~ 2,
        TRUE ~ gunhomeper
      )  %>% 
      factor(
          levels = c(1:2),
          labels =
            c("Yes", "No")
        ),
    gunhome = 
      ifelse(gunhome > 2, 99, gunhome) %>% 
      factor(
        levels = c(1:2, 99),
        labels =
          c("Yes", "No", "Unknown")
        )
  ) %>% 
  set_variable_labels(
    gunhome = "Do you happen to have in your home or garage any guns or revolvers?",
    gunhomeper = "Do any of guns or revolvers in your home or garage personally belong to you?"
  )
```

```{r}
raw %>% tabyl(gunhome, gunhomeper) %>% adorn_totals("row")
```

## Type of gun ownership

Do you personally own any of the following types of guns?

-   `qguntypea` = handguns

-   `qguntypeb` = rifles

-   `qguntypec` = shotguns

-   `qguntyped` = other

```{r}
walk(
  c("qguntypea", "qguntypeb", "qguntypec", "qguntyped"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <- 
  raw %>% 
  rename(
    ownhandgun = qguntypea,
    ownrifle = qguntypeb, 
    ownshot = qguntypec,
    ownother = qguntyped
  ) %>% 
  mutate(
    across(
      starts_with("own"),
      ~ case_when(
        is.na(.x) & gunhomeper == "No" ~ 100,
        is.na(.x) | .x == 77 | .x %in% 98:99 ~ 99,
        TRUE ~ .x
      ) %>% 
        factor(
          levels = c(1:2, 99:100),
          labels =
            c("Yes", 
              "No",
              "Unknown",
              "Non-gun owner")
        )
    )
  ) %>% 
    set_variable_labels(
      ownhandgun = "Do you personally own any hanguns?",
      ownrifle = "Do you personally own any rifles?", 
      ownshot = "Do you personally own any shotguns?",
      ownother = "Do you personally own any other types of guns?"
    )
```

```{r}
walk(
    raw %>% select(starts_with("own")) %>% names(),
  ~ print(raw %>% tabyl(.x))
)
```



##  Have you bought any guns since January 1, 2020? (Q42_23, Q43_23a-c)

-   `q42_23` = Have you bought any guns since January 1, 2020? = `gun2020`

-   `q43_23a` = Did you buy your first gun after January 1, 2020? = `firstgun2020`

-   `q43_23b` = Was your purchase after January 1, 2020 motivated by concerns of racial violence? = `racialgun2020`

-   `q43_23c` = Was your purchase after January 1, 2020 motivated by concerns of political violence? = `polgun2020`

- Don't know and Refused grouped as Unknown.

```{r}
walk(
  c("q42_23", "q43_23a", "q43_23b", "q43_23c"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <- 
  raw %>% 
  rename(
    gun2020 = q42_23,
    firstgun2020 = q43_23a,
    racialgun2020 = q43_23b,
    polgun2020 = q43_23c
  ) %>% 
  mutate(
    across(
      ends_with("gun2020"),
      ~ case_when(
        is.na(.x) ~ 100,
        .x  == 98 ~ 77,
        TRUE ~ .x
      ) %>% 
        factor(
          levels = c(1:2, 77, 100),
          labels = 
            c("Yes",
              "No",
              "Unknown",
              "Did not purchase a gun since January 1, 2020")
        )
    )
  ) %>% 
  set_variable_labels(
    gun2020 = "Have you bought any guns since January 1, 2020?",
    firstgun2020 = "Did you buy your first gun after January 1, 2020?",
    racialgun2020 = "Was your purchase after January 1, 2020 motivated by concerns of racial violence?",
    polgun2020 = "Was your purchase after January 1, 2020 motivated by concerns of political violence?"
  )
```

```{r}
walk(
  raw %>% select(ends_with("gun2020")) %>% names ,
  ~ print(raw %>% tabyl(.x))
)
```

## Which of the following is your primary reason for owning guns? (Q47_23)

Which of the following is your primary reason for owning guns? = `gunreason`

Don't know, Skipped on web, and Refused grouped as unknown.

```{r}
raw %>% tabyl(q47_23)
raw %>% tabyl(q47_23)
```

```{r}
raw <-
  raw %>% 
  rename(gunreason = q47_23) %>% 
  mutate(
    gunreason = 
      case_when(
        is.na(gunreason) ~ 100,
        gunreason %in% c(77, 98) ~ 99,
        TRUE ~ gunreason
      ) %>% 
      factor(
        levels = c(1:10, 99:100),
        labels = 
            c("For protection from dangerous people or situations when I am at home",
              "For protection from dangerous people or situations when I am away from home",
            "For protection at public events, such as sporting events or concerts",
            "For protection during political activities, such as while voting or listening to speeches",
            "For protection at demonstrations, rallies, or protests",
            "For protection against people who do not share my beliefs",
            "For protection against police violence",
            "For situations where I think that force or violence is justified to advance an important political objective",
            "For hunting",
            "For other recreational activities, such as target shooting",
            "Unknown",
            "Non-gun owner")
      )
  ) %>% 
  set_variable_labels(
    gunreason = "Which of the following is your primary reason for owning guns?"
  )
```

```{r}
raw %>% tabyl(gunreason)
```

## Veteran

-   `veteran_1` = Have you ever served on active duty in the U.S. Armed Forces, military Reserves, or National Guard? = `vet`

-   `veteran2_1` = Are you currently on active duty in the U.S Armed Forces, military Reserves, or National Guard? = `activevet`

```{r}
raw %>% tabyl(veteran_1, veteran2_1)
```

```{r}
raw <-
  raw %>% 
  rename(vet = veteran_1, activevet = veteran2_1) %>% 
  mutate(
    activevet =
      case_when(
        is.na(activevet) ~ vet,
        TRUE ~ activevet
      ),
    across(
      ends_with("vet"),
      ~ factor(
        .x, 
        levels = c(1:2, 98),
        labels = c("Yes", "No", "Unknown")
      )
    )
  ) %>% 
  set_variable_labels(
    activevet = "Are you currently on active duty in the U.S Armed Forces, military Reserves, or National Guard?",
    vet = "Have you ever served on active duty in the U.S. Armed Forces, military Reserves, or National Guard?"
  )
```

```{r}
raw %>% tabyl(vet, activevet)
```

## Political party

```{r}
raw %>% tabyl(party_id7, party_id5)
```

```{r}
raw <-
  raw %>% 
  rename(party7 = party_id7, party5 = party_id5) %>% 
  mutate(
    party7 =
      factor(
        as.numeric(party7), 
        levels = c(1:7, -1),
        labels = 
          c("Strong Democrat",
            "Not so strong Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Not so strong Republican",
            "Strong Republican",
            "Unknown")
      ),
    party5 =
      factor(
        party5,
        levels = c(1:5, -1),
        labels = 
          c("Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Republican",
            "Unknown")
      ),
    party3 = 
      case_when(
        grepl("demo", party5, ignore.case = T) ~ 1,
        grepl("rep", party5, ignore.case = T) ~ 2,
        grepl("inde", party5, ignore.case = T) ~ 3,
        TRUE ~ -1
      ) %>% 
       factor(
        party5,
        levels = c(1:3, -1),
        labels = 
          c("Democrat",
            "Republican",
            "Independent",
            "Unknown"
            )
      )
  ) %>% 
  set_variable_labels(
    party7 = "7-level political affiliation",
    party5 = "5-level political affiliation",
    party3 = "3-level political affiliation"
  )
```

```{r}
raw %>% tabyl(party7, party5)
raw %>% tabyl(party7, party3)
```

## Elections

-   `qvotediff` = How easy or difficult was it for you to vote in the 2020 election? = `votediff20`

-   `q_vote22` = In talking to people about the 2020 election, we often find that a lot of people were not able to vote because they weren’t registered, they were sick, or they just didn’t have time. Which one of the following statements best describes you? = `votereason22`

-   `qcandi20` = Who did you vote for in the 2020 election = `votecandi20`

-   `qvotediff2` = How easy or difficult was it for you to vote in the 2020 election? = `votediff22`

Grouped "Don't know", "Skipped on web" into Unknown.

```{r}
walk(
  c("qvotediff", "qcandi20", "q_vote22", "qvotediff2"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <-
  raw %>% 
  rename(
    votediff20 = qvotediff,
    votereason22 = q_vote22,
    votecandi20 = qcandi20,
    votediff22 = qvotediff2
  ) %>% 
  mutate(
    across(
      starts_with("voted"),
      ~ case_when(
        is.na(.x) ~ 100,
        .x == 77 ~ 98,
        TRUE ~ .x
      ) %>% 
        factor(
        levels = c(1:5, 100, 98),
        labels = 
          c("Very easy",
            "Easy",
            "Neither easy or difficult",
            "Difficult",
            "Very difficult",
            "Did not vote in this race",
            "Unknown")
      )
    ),
    votecandi20 =
      ifelse(votecandi20 == 98, 99, votecandi20) %>% 
      factor(
        levels = c(1:4, 99),
        labels = 
        c("Joe Biden",
          "Donald Trump",
          "Someone else",
          "Did not vote in this race",
          "Unknown"
        )
      ),
    votereason22 = 
      ifelse(votereason22 > 4, 99, votereason22) %>% 
      factor(
        levels = c(1:4, 99),
        labels =
          c(
            "I did not vote in the 2022 election",
            "I thought about voting in the 2022 election, but didn't",
            "I usually vote, but I didn't in the 2022 election",
            "I'm sure I voted",
            "Unknown"
          )
      )
  ) %>% 
  set_variable_labels(
    votediff20 = "How easy or difficult was it for you to vote in the 2020 election?",
    votereason22 = "In talking to people about elections, we often find that a lot of people were not able to vote because they weren’t registered, they were sick, or they just didn’t have time. Which one of the following statements best describes you?",
    votecandi20 = "Who did you vote for in the 2020 election?",
    votediff22 = "How easy or difficult was it for you to vote in the 2022 election?"
  )
```

```{r}
walk(
  raw %>% select(matches("vote")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## Household size
```{r}
raw <-
  raw %>%  
  rowwise() |> 
  mutate(hh18un = sum(c(hh01, hh25, hh612, hh1317))) |> 
  ungroup() |> 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh01 = "Number of HH members age 0-1",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+",
    hh18un = "Number of HH members under 18"
  )
```


```{r}
colSums(is.na(raw))
```

```{r}
hist(raw$hh18un)
```

```{r}
nrow(filter(raw, hh18un < 0) ) == 0
```

## Weights

`weight1` = general population analysis
`weight3` = gun owner specific analysis

```{r}
clean <-
  raw %>% 
  rename(weight1 = weight) %>% 
  set_variable_labels(
    weight1 = "Post-stratification weights - 18+ general population (N=3,096)",
    weight2 = "Post-stratified weights - scaled to 4 race groups (NH-Black, Hispanic, AAPI, NH-All Other) (N=3,096)",
weight3 = "Post-stratified weights - scaled to 2 groups (gun owners vs not gun owners) (N=3,096)"
  )
```

## Check for missing values

```{r}
colSums(is.na(clean))
```


## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2023.rds"
    )
  ) 
```


## Explore demographic variables

### Gun owners

```{r}
demovars <-
  c(
    "age4",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type"
  )

summstats <-
  c("notNA(x)", "mean(x)")
summnames <-
  c("Observations", "Mean")
```

```{r, results = "asis", warning = F, message = F}
st(
  clean %>% filter(gunhomeper == "Yes"),
  vars = demovars,
  group.weights = "weight1",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among gun owners",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

### Full sample

```{r}
demovars <- c("gunhomeper", demovars)
```

```{r, results = "asis", warning = F, message = F}
st(
  clean,
  vars = demovars,
  group.weights = "weight1",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among full sample",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

