# Clean 2025 data

We use this [codebook](https://livejohnshopkins-my.sharepoint.com/:x:/r/personal/jdada3_jh_edu/Documents/Research/DemographyGunOwners/documentation/2025/A118%20JHU%20Gun%20Policy%202025_Codebook.xlsx?d=wfaba06bcf9be4a6eabfffee71f62b7f1&csf=1&web=1&e=InolJ9).

Note that all of the demographic variables are the same as the 2021 survey.

**Input**

-   DemographyGunOwners/data/raw/gunsurvey/2025.rds

**Output**

-   DemographyGunOwners/data/clean/gs-2025.rds

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "labelled",
    "tidyr",
    "janitor",
    "weights",
    "conflicted",
    "kableExtra",
    "survey",
    "skimr",
    "psych",
    "dplyr",
    "here",
    "vtable",
    "fixest",
    "lfe",
    "broom"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)

conflicts_prefer(dplyr::filter)
```

## Import data

```{r}
source <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2025.rds"
      )
  ) %>% 
  clean_names()
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey 2025 Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|------------------------|---------------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight1)`                           |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |

## Select variables

All questions are encoded, so we decode them using the [codebook](https://livejohnshopkins-my.sharepoint.com/:x:/r/personal/jdada3_jh_edu/Documents/Research/DemographyGunOwners/documentation/2021%20Documentation/8999_JHU_Gun%20Policy%202021_Codebook.xlsx?d=w75cec2186ce04efd95f19e64b723f885&csf=1&web=1&e=YHbc6U).

We focus on demographic variables and the main variables about gun ownership.

```{r}
demovars <-
  c(
    "age",
    "age4",
    "age7",
    "hhsize",
    "racethnicity",
    "party_id7",
    "party_id5",
    "veteran_1", 
    "veteran2_1",
    "educ5",
    "marital",
    "employ",
    "income",
    "income4",
    "income9",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov",
    # Only in 2025
    "gender_birth1",
    "gender_current1",
    "gender_trans1",
    "lgbt"  
  )

gun <-
  c(
    "gun",
    "q39a",
    "q39b",
    "q40",
    "q41",
    "qguntypea", 
    "qguntypeb",
    "qguntypec", 
    "qguntyped",
    # 2025
    "firstgun_age"
  )
```

```{r}
raw <-
  source %>% 
  dplyr::select(
    starts_with(c("case_id", "weight")), 
    all_of(c(demovars, gun))
  ) %>% 
  rename(caseid = case_id)
```

## Gun

Proportion-based simple random imputation used to determine gun ownership for those with missing data. Used for weighting purposes.

```{r}
raw %>% tabyl(gun)
```

```{r}
raw <-
  raw %>% 
  mutate(
    gun = 
      factor(
        gun,
        levels = 1:2,
        labels = 
          c("Gun owner",
            "Non-gun owner")
      )
  ) %>% 
  set_variable_labels(
    gun = "Proportion-based simple random imputation used to determine gun ownership for those with missing data. Used for weighting purposes"
  )
```

```{r}
raw %>% tabyl(gun)
```

## Gender and sexuality

I choose to label `gender_current1` as the primary gender variable for the analysis.

For LGBT, I recoding "something else" and skipped on web as unknown. I do similarly throughout cleaning this data for the simplicity of analysis.

```{r}
walk(
  c("gender_birth1", "gender_current1", "gender_trans1", "lgbt"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <-
  raw %>% 
  rename(
    gender = gender_current1, 
    gender_as = gender_birth1,
    gender_tr = gender_trans1
  ) %>% 
  mutate(
    gender = 
      factor(
        gender, 
        levels = c(1:4, 98),
        labels = c(
          "Male",
          "Female",
          "Transgender",
          "Do not identify as male, female, or transgender",
          "Unknown" # Recoded from "Skipped on web"
          )
      ),
    gender_as = 
      factor(
        gender_as, 
        levels = 1:2,
        labels = c("Male", "Female")
      ),
    gender_tr =
      ifelse(is.na(gender_tr), 999, gender_tr) %>% 
      factor(
        levels = c(1:4, 999),
        labels = 
          c("Male-to-Female", 
            "Female-to-Male", 
            "Transgender",
            "Other", 
            "Not transgender")
      ),
    lgbt = 
      ifelse(lgbt == 5, 98, lgbt) %>% 
      factor( 
        levels = c(2, 1, 3:4, 98),
        labels = 
          c("Straight", "Gay or Lesbian", "Bisexual", "Other", "Unknown")
        )
  ) %>% 
  set_variable_labels(
    gender = "Respondent gender",
    gender_as = "Respondent gender, assigned at birth",
    gender_tr = "Respondent transgender identity",
    lgbt = "Respondent sexual identity"
  )
```

```{r}
walk(
  c("gender", "gender_as", "gender_tr", "lgbt"),
  ~ print(raw %>% tabyl(.x))
)
```

## Age

Note that there are clearly imputation issues with the continuous age variable, 
with a maximum value of 998.

```{r}
hist(raw$age)
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

```{r}
raw <-
  raw %>% 
  mutate(
    age4 = 
      factor(
        age4,
        levels = 1:4,
        labels =
          c("18-29",
            "30-44",
            "45-59",
            "60+")
      ),
    age7 = 
      factor(
        age7,
        levels = 1:7,
        labels =
          c("18-24",
            "25-34",
            "35-44",
            "45-54",
            "55-64",
            "65-74",
            "75+")
      )
  ) %>% 
  set_variable_labels(
    age = "Age",
    age4 = "Age - 4 Categories",
    age7 = "Age - 7 Categories"
  )
```

```{r}
raw %>% tabyl(age4)
raw %>% tabyl(age7)
```

## Race and ethnicity

```{r}
raw %>% tabyl(racethnicity)
```

```{r}
raw <-
  raw %>% 
  mutate(
    racethnicity = 
      factor(
        racethnicity,
        levels = 1:6,
        labels =
          c("White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "Hispanic",
            "2+, non-Hispanic",
            "Asian, non-Hispanic")
      )
  ) %>% 
  set_variable_labels(
    racethnicity = "Combined race/ethnicity"
  )
```

```{r}
raw %>% tabyl(racethnicity)
```

## Education

```{r}
raw %>% tabyl(educ5)
```

```{r}
raw <-
  raw %>% 
  mutate(
    educ5 = 
      factor(
        educ5,
        levels = 1:5,
        labels = 
          c("Less than HS",
            "HS graduate",
            "Some college/associates degree",
            "Bachelors degree",
            "Post grad study/professional degree"
            )
      )
  ) %>% 
  set_variable_labels(
    educ5 = "Highest level of education"
  )
```

```{r}
raw %>% tabyl(educ5)
```

## Martial status

```{r}
raw %>% tabyl(marital)
```

```{r}
raw <-
  raw %>% 
  mutate(
    marital = droplevels(as_factor(marital))
    ) %>% 
  set_variable_labels(
    marital = "Marital status"
  )
```

```{r}
raw %>% tabyl(marital)
```

## Employment

```{r}
raw %>% tabyl(employ)
```

```{r}
raw <-
  raw %>% 
  mutate(
    employ = droplevels(as_factor(employ))
  )   %>% 
  set_variable_labels(
    employ = "Current employment status"
  )
```

```{r}
raw %>% tabyl(employ)
```

## Income

```{r}
walk(
  select(raw, starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <-
  raw %>% 
  mutate(
    income = 
      factor(
        income, 
        levels = 1:18,
        labels = 
          c("Less than $5,000",
            "$5,000-9,999",
            "$10,000-14,999",
            "$15,000-19,999",
            "$20,000-24,999",
            "$25,000-29,999",
            "$30,000-34,999",
            "$35,000-39,999",
            "$40,000-49,999",
            "$50,000-59,999",
            "$60,000-74,999",
            "$75,000-84,999",
            "$85,000-99,999",
            "$100,000-124,999",
            "$125,000-149,999",
            "$150,000-174,999",
            "$175,000-199,999",
            "$200,000 or more")
      ),
    income4 = 
      factor(
        income4, 
        levels = 1:4, 
        labels = 
          c("Less than $30,000",
            "$30,000-60,000",
            "$60,000-100,000",
            "$100,000 or more")
      ),
    income9 = 
      factor(
        income9, 
        levels = 1:9, 
        labels = 
          c("Under $10,000",
            "$10,000-20,000",
            "$20,000-30,000",
            "$30,000-40,000",
            "$40,000-50,000",
            "$50,000-75,000",
            "$75,000-100,000",
            "$100,000-150,000",
            "$150,000 or more")
      )
  )  %>% 
  set_variable_labels(
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories"
  )
```

```{r}
walk(
  select(raw, starts_with("income")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## State

```{r}
raw %>% tabyl(state)
```

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    state = "State"
  )
```

## Region

```{r}
walk(
  select(raw, starts_with("region")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <-
  raw %>%  
  mutate(
    across(starts_with("region"), ~ droplevels(as_factor(.x)))
    )  %>% 
  set_variable_labels(
    region4 = "4-level region",
    region9 = "9-level region"
  )
```

```{r}
walk(
  select(raw, starts_with("region")) %>% names,
  ~ print(raw %>% tabyl(.x))
)
```

## Metropolitan area flag

```{r}
raw %>% tabyl(metro)
```

```{r}
raw <-
  raw %>% 
  mutate(
    metro = droplevels(as_factor(metro)) 
  ) %>% 
  set_variable_labels(
    metro = "Metropolitan area flag"
  )
```

```{r}
raw %>% tabyl(metro)
```

## Internet

```{r}
raw %>% tabyl(internet)
```

```{r}
raw <-
  raw %>% 
  mutate(
    internet = droplevels(as_factor(internet)) 
  ) %>% 
  set_variable_labels(
    internet = "HH internet access via dial-up, DSL, or cable broadband at home"
  )
```

```{r}
raw %>% tabyl(internet)
```

## Housing

```{r}
raw %>% tabyl(housing)
```

```{r}
raw <-
  raw %>% 
  mutate(
    housing = droplevels(as_factor(housing))
  ) %>% 
  set_variable_labels(
    housing = "Home ownership"
  )
```

```{r}
raw %>% tabyl(housing)
```

## Home type

```{r}
raw %>% tabyl(home_type)
```

```{r}
raw <- 
  raw %>% 
  mutate(
    home_type = as_factor(home_type)
  ) %>% 
  set_variable_labels(
    home_type = "Type of building of panelists' residence"
  )
```

```{r}
raw %>% tabyl(home_type)
```


## Phone service

```{r}
raw %>% tabyl(phoneservice)
```

```{r}
raw <-
  raw %>% 
  mutate(
    phoneservice = as_factor(phoneservice)
  ) %>% 
  set_variable_labels(
    phoneservice = "Telephone service for the household"
  )
```

```{r}
raw %>% tabyl(phoneservice)
```


## Are you a member of the National Rifle Association (Q39a and b)

Are you a member of the National Rifle Association--also known as the NRA? = `q39a` = `nranow`

Have you ever been a member of the National Rifle Association--also known as the NRA? = `q39b` = `nraever`

There are `r nrow(filter(raw, q39b =="NA"))` observations with NA hardcoded for `q39b`. This corresponds to people who said in `q39a` that they are current members of the NRA or skipped/refused to answer `q39a`.

I group together  "Skipped on web", "Refused" as "Unknown"

```{r}
raw %>% tabyl(q39a, q39b)
```

```{r, warning = F}
raw <-
  raw %>% 
  rename(
    nranow = q39a,
    nraever = q39b
    ) %>% 
  mutate(
    nraever = 
      case_when(
        is.na(nraever) ~ as.numeric(nranow),
        TRUE ~ as.numeric(nraever)
      ),
    across(
      starts_with("nra"),
      ~ case_when(
          .x  == 98 ~ 99,
          TRUE ~ .x) %>% 
        factor( 
          levels = c(1:2, 99),
          labels =
            c("Yes", "No", "Unknown")
          )
    )
  ) %>% 
  set_variable_labels(
    nraever = "Have you ever been a member of the NRA?",
    nranow = "Are you a member of the NRA?"
  )
```

```{r}
raw %>% tabyl(nranow, nraever)
```


## Gun ownership (Q40 & 41)

Do you happen to have in your home or garage any guns or revolvers? = `q40` = `gunhome`

Do any of these guns personally belong to you? = `gunhomeper`. There are `r nrow(filter(raw, q41 =="NA"))` observations with NA hardcoded for `q41`. This corresponds to people who said answered no to q40.

```{r}
raw %>% tabyl(q40)
raw %>% tabyl(q41)
```

```{r}
raw <-
  raw %>% 
  rename(
    gunhome = q40,
    gunhomeper = q41
    ) %>% 
  mutate(
    gunhomeper = 
      case_when(
       is.na(gunhomeper) ~  as.character(gunhome), 
        TRUE ~ as.character(gunhomeper)
      ) %>% as.numeric,
    across(
      starts_with("gunhome"),
      ~ case_when(
          .x == 98 ~ 99,
          TRUE ~ .x
          ) %>%  
        factor(
          levels = c(1:2, 99),
          labels =
            c("Yes", "No", "Unknown")
        )
      )
  ) %>% 
  set_variable_labels(
    gunhome = "Do you happen to have in your home or garage any guns or revolvers?",
    gunhomeper = "Do any of guns or revolvers in your home or garage personally belong to you?"
  )
```

```{r}
raw %>% tabyl(gunhome, gunhomeper)
```


## Type of gun ownership

Do you personally own any of the following types of guns?

-   `qguntypea` = handguns

-   `qguntypeb` = rifles

-   `qguntypec` = shotguns

-   `qguntyped` = other

```{r}
walk(
  c("qguntypea", "qguntypeb", "qguntypec", "qguntyped"),
  ~ print(raw %>% tabyl(.x))
)
```

```{r}
raw <- 
  raw %>% 
  rename(
    ownhandgun = qguntypea,
    ownrifle = qguntypeb, 
    ownshot = qguntypec,
    ownother = qguntyped
  ) %>% 
  mutate(
    across(
      starts_with("own"),
      ~ case_when(
        is.na(.x) ~ 100,
        .x %in% c(77, 98:99) ~ 77,
        TRUE ~ .x
      ) %>% 
        factor(
          levels = c(1:2, 77, 100),
          labels =
            c("Yes", 
              "No",
              "Unknown",
              "Non-gun owner")
        )
    )
  ) %>% 
    set_variable_labels(
      ownhandgun = "Do you personally own any hanguns?",
      ownrifle = "Do you personally own any rifles?", 
      ownshot = "Do you personally own any shotguns?",
      ownother = "Do you personally own any other types of guns?"
    )
```

```{r}
walk(
    raw %>% select(starts_with("own")) %>% names(),
  ~ print(raw %>% tabyl(.x))
)
```


## Veteran

-   `veteran_1` = Have you ever served on active duty in the U.S. Armed Forces, military Reserves, or National Guard? = `vet`

-   `veteran2_1` = Are you currently on active duty in the U.S Armed Forces, military Reserves, or National Guard? = `activevet`

```{r}
raw %>% tabyl(veteran_1, veteran2_1)
```


```{r}
raw <-
  raw %>% 
  rename(vet = veteran_1, activevet = veteran2_1) %>% 
  mutate(
    activevet =
      case_when(
        is.na(activevet) ~ vet,
        TRUE ~ activevet
      ),
    across(
      ends_with("vet"),
      ~ factor(
        .x, 
        levels = c(1:2, 98),
        labels = c("Yes", "No", "Unknown")
      )
    )
  ) %>% 
  set_variable_labels(
    activevet = "Are you currently on active duty in the U.S Armed Forces, military Reserves, or National Guard?",
    vet = "Have you ever served on active duty in the U.S. Armed Forces, military Reserves, or National Guard?"
  )
```

```{r}
raw %>% tabyl(vet, activevet)
```


## Political party

```{r}
raw %>% tabyl(party_id7, party_id5)
```

```{r}
raw <-
  raw %>% 
  rename(party7 = party_id7, party5 = party_id5) %>% 
  mutate(
    party7 =
      factor(
        as.numeric(party7), 
        levels = c(1:7, -1),
        labels = 
          c("Strong Democrat",
            "Not so strong Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Not so strong Republican",
            "Strong Republican",
            "Unknown")
      ),
    party5 =
      factor(
        party5,
        levels = c(1:5, -1),
        labels = 
          c("Democrat",
            "Lean Democrat",
            "Don't Lean/Independent/None",
            "Lean Republican",
            "Republican",
            "Unknown")
      )
  ) %>% 
  set_variable_labels(
    party7 = "7-level political affiliation",
    party5 = "5-level political affiliation"
  )
```

```{r}
raw %>% tabyl(party7, party5)
```

## At what age did you buy your first gun?

We define a categorical variable to match the age4 with an added category for < 18 and 
for not purchasing a gun (missing values). Some observations have firstgun == 998, 
which I assume to mean gun owners so I code them as unknown.

```{r}
raw %>% tabyl(firstgun_age)
```

```{r}
raw <-
  raw %>% 
  rename(firstgun = firstgun_age) %>% 
  mutate(
    firstgun7 = 
      case_when(
        firstgun < 18 ~ "Under 18",
        firstgun %in% 18:29 ~ "18-29",
        firstgun %in% 30:44 ~ "30-44",
        firstgun %in% 45:59 ~ "45-59",
        firstgun %in% 60:100 ~ "60+",
        firstgun == 998 ~ "Unknown",
        TRUE ~ "Never purchased a gun"
      )
  ) %>% 
  set_variable_labels(
    firstgun7 = "At what age did you buy your first gun? 7 categories",
    firstgun = "At what age did you buy your first gun?"
  )
```

```{r}
raw %>% tabyl(firstgun7)
```
## Household size

```{r}
raw <-
  raw %>% 
  set_variable_labels(
    hhsize = "Household size (including children)",
    hh01 = "Number of HH members age 0-1",
    hh25 = "Number of HH members age 2-5",
    hh612 = "Number of HH members age 6-12",
    hh1317 = "Number of HH members age 13-17",
    hh18ov = "Number of HH members age 18+"
  )
```

## Weights

```{r}
clean <-
  raw %>% 
  rename(weight = weight1) %>% 
  set_variable_labels(
    weight = "Post-stratification weights - 18+ general population",
    weight2 = "Post-stratified weights - scaled to 3 race groups (NH-Black, Hispanic, NH-All Other)",
weight3 = "Post-stratified weights - scaled to 2 groups (gun owners vs not gun owners)"
  )
```

## Explore demographic variables

### Gun owners

```{r}
demovars <-
  c(
    "age",
    "hhsize",
    "gender",
    "raceethnicity",
    "educ",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type",
    "party7"
  )

summstats <-
  c("notNA(x)", "mean(x)", "sd(x)")
summnames <-
  c("Observations", "Mean", "SD")
```

```{r, results = "asis", warning = F, message = F}
st(
  clean %>% filter(gunhomeper == "Yes"),
  vars = demovars,
  group.weights = "weight3",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```

### Full sample

```{r}
demovars <-
  c(
    "gunhomeper",
    "age",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income",
    "internet",
    "home_type",
    "party7"
  )
```


```{r, results = "asis", warning = F, message = F}
st(
  clean,
  vars = demovars,
  group.weights = "weight2",
  summ = summstats,
  summ.names = summnames,
  title = "Weighted demographic summary among full sample",
  out = "kable",
  numformat = "comma",
  labels =  T
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
  ) %>% 
  scroll_box(width = "800px", height = "500px")
```


## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "gs-2025.rds"
    )
  ) 
```
