# Merge survey years

This analysis creates a combined data set of the survey years.


**Input**

-   DemographyGunOwners/data/clean/gs-`i`.rds for `i %in% 2019, 2021, 2023`

**Output**


**Last ran** 

-   01/28/26

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)

cleanfold <-
  here(
    path_od, 
    "data",
    "clean"
  )
```


## Import data

```{r}
years <- c(2019, 2021, 2023)
```

```{r}
source <-
  map_dfr(
    years, 
    ~ read_rds(here(cleanfold,paste0("gs-", .x, ".rds"))) %>% 
      mutate(year = .x)
  )
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-------------------------|-----------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                           |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |
| Years of data          | `r source %>% distinct(year) %>% pull(year) %>% paste(collapse = ", ")` |

## Create unique ID

Within years, the `caseid` is the ID variable. We redefine `caseid` to be unique
across years.

```{r}
merge <-
  source %>% 
  select(-caseid) %>% 
  arrange(year, weight) %>% 
  mutate(caseid = row_number())
```

## Check that the common columns are all non-missing

Different survey years have different questions, but they key demographic variables
are the same across years and should not be missing. We also need survey weights to be non-missing.

```{r}
commonvars <-
  c(
    "weight",
    "caseid",
    "year",
    "nranow",
    "gunhome",
    "gunhomeper",
    "party7",
    "age",
    "age4",
    "age7",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income9",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov"
  )
```

```{r}
merge %>% 
  select(all_of(commonvars)) %>% 
  filter(if_any(commonvars, is.na)) %>% 
  nrow() == 0
```


## Export data

```{r}
write_rds(
  merge,
  here(
    cleanfold,
    "gs-all.rds"
  )
)
```
