# Merge survey years

This analysis creates a combined data set of the survey years.


**Input**

-   DemographyGunOwners/data/clean/gs-`i`.rds for `i %in% 2019, 2021, 2023`

**Output**


**Last ran** 

-   01/28/26

```{r,  message = FALSE, warning = FALSE, echo = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)

cleanfold <-
  here(
    path_od, 
    "data",
    "clean"
  )
```


## Import data

```{r}
years <- c(2019, 2021, 2023)
```

```{r}
source <-
  map_dfr(
    years, 
    ~ read_rds(here(cleanfold,paste0("gs-", .x, ".rds"))) %>% 
      mutate(year = .x)
  )
```

## Explore data

```{r}
source %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

### Summary table

| Name                   | Value                                             |
|-------------------------|-----------------------------------------------|
| Number of observations | `r nrow(source)`                                  |
| Number of unique cases | `r n_distinct(source$case_id)`                    |
| Sum of weights         | `r sum(source$weight)`                           |
| Number of variables    | `r length(source)`                                |
| Variable names         | `r source %>% names() %>% paste(collapse = ", ")` |
| Years of data          | `r source %>% distinct(year) %>% pull(year) %>% paste(collapse = ", ")` |

## Create unique ID

Within years, the `caseid` is the ID variable. We redefine `caseid` to be unique
across years.

```{r}
merge <-
  source %>% 
  select(-caseid) %>% 
  arrange(year, weight) %>% 
  mutate(caseid = row_number())
```

## Check that the common columns are all non-missing

Different survey years have different questions, but they key demographic variables
are the same across years and should not be missing. We also need survey weights to be non-missing.

```{r}
commonvars <-
  c(
    "weight",
    "caseid",
    "year",
    "nranow",
    "gunhome",
    "gunhomeper",
    "party7",
    "age",
    "age4",
    "age7",
    "hhsize",
    "gender",
    "racethnicity",
    "educ5",
    "marital",
    "employ",
    "income9",
    "state",
    "region4",
    "region9",
    "metro",
    "internet",
    "housing",
    "home_type",
    "phoneservice",
    "hhsize",
    "hh01",
    "hh25",
    "hh612",
    "hh1317",
    "hh18ov"
  )
```

```{r}
merge %>% 
  select(all_of(commonvars)) %>% 
  filter(if_any(commonvars, is.na)) %>% 
  nrow() == 0
```

## Create household size grouping variables

`hh06` = age 0-6

`hh18un` = Under 18

## Relabel

```{r}
merge <-
  merge %>% 
  set_variable_labels(
    gender = "Respondent gender",
    racethnicity = "Combined race/ethnicity",
    marital = "Marital status",
    income = "Household income",
    income4 = "Household income: 4 categories",
    income9 = "Household income: 9 categories",
    region4 = "4-level region",
    region9 = "9-level region",
    metro = "Metropolitan area flag",
    internet = "HH internet access via dial-up, DSL, or cable broadband at home",
    home_type = "Type of building of panelists' residence",
    phoneservice = "Telephone service for the household",
    ownhandgun = "Do you personally own any hanguns?",
    ownrifle = "Do you personally own any rifles?", 
    ownshot = "Do you personally own any shotguns?",
    ownother = "Do you personally own any other types of guns?",
    nraever = "Have you ever been a member of the NRA?",
    nranow = "Are you a member of the NRA?",
    goanow = "Are you a member of any other gun rights groups such as Gun Owners of America?",
    mdanow = "Are you a member of any gun safety groups such as Moms Demand Action or Everytown for Gun Safety?",
    gunhome = "Do you happen to have in your home or garage any guns or revolvers?",
    gunhomeper = "Do any of guns or revolvers in your home or garagepersonally belong to you?",
    gun2020 = "Have you bought any guns since January 1, 2020?",
    firstgun2020 = "Did you buy your first gun after January 1, 2020?",
    racialgun2020 = "Was your purchase after January 1, 2020 motivated by concerns of racial violence?",
    polgun2020 = "Was your purchase after January 1, 2020 motivated by concerns of political violence?",
    gunreason = "Which of the following is your primary reason for owning guns?",
    activevet = "Are you currently on active duty in the U.S Armed Forces, military Reserves, or National Guard?",
    vet = "Have you ever served on active duty in the U.S. Armed Forces, military Reserves, or National Guard?",
    party7 = "7-level political affiliation",
    party5 = "5-level political affiliation",
    votediff20 = "How easy or difficult was it for you to vote in the 2020 election?",
    votereason22 = "In talking to people about elections, we often find that a lot of people were not able to vote because they weren’t registered, they were sick, or they just didn’t have time. Which one of the following statements best describes you?",
    votecandi20 = "Who did you vote for in the 2020 election?",
    votediff22 = "How easy or difficult was it for you to vote in the 2022 election?",
    weight = "Post-stratification weights - 18+ general population (N=2,778)",
    weight2 = "Post-stratified weights - scaled to 3 race groups (NH-Black, Hispanic, NH-All Other) (N=2,778)",
weight3 = "Post-stratified weights - scaled to 2 groups (gun owners vs not gun owners) (N=2,778)"
  )
```

## Export data

```{r}
write_rds(
  merge,
  here(
    cleanfold,
    "gs-all.rds"
  )
)
```
