## 2015

No data on 2015

The only reasonable output for 2015 in this data is 19.8\% using `weight2`.
This implies that gun ownership decreased from 2013 and 2015. 

[JHU Gun Center 2015 Survey Results](https://publichealth.jhu.edu/2015/large-majority-of-americans-includin-gun-owners-support-stronger-gun-safety-policies)

[Barry paper on 2015 Survey](https://www.sciencedirect.com/science/article/pii/S0091743515001668?via=ihub).

The sample counts of gun owners and non-gun owners match the Barry paper. The Barry paper does not include an estimate for the percentage of US gun owners but because the sample
counts are the same, I am more confident that this is accurate.

```{r,  message = FALSE, warning = FALSE, echo = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "skimr",
    "psych",
    "kableExtra",
    "dplyr",
    "vtable",
    "here",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```


## Load data

```{r}
data <-
  read_rds(
   here(
    path_od, 
    "data",
    "clean",
    "gs-2015.rds"
    )
  )
```

This matches exactly.

```{r}
data %>%  tabyl(sample, gunhomeper) %>% adorn_totals("row")
```

## National prevalence

```{r}
data %>% tabyl(sample, gunhomeper)
```

```{r}
data %>% 
  mutate(
    g = as.numeric(gunhomeper == "Yes")
  ) %>% 
  summarise(
    across(
      c("weight1", "weight2", "weight3"), 
      ~ sum(g*.x, na.rm = T)/sum(.x, na.rm = T) %>%  round(2) *100
    )
  )
```

```{r}
filter(data, is.na(weight2)) %>% nrow() == 0
```
