# NRA Membership

This script regresses NRA membership on demographic characteristics

**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/data/raw/gunsurvey/...rds

**Last ran**

- 03/08/2026

```{r,  message = FALSE, warning = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "dplyr",
    "here",
    "survey",
    "broom",
    "kableExtra",
    "modelsummary",
    "labelled"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
gun <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```


```{r}
gun %>% tabyl(nranow)
```


## Define binary indicator for NRA membership

```{r}
nra <-
  gun %>% 
  mutate(
    nra = (nranow == "Member") %>% as.numeric %>% factor()
  )
```

```{r}
nra %>% tabyl(nra)
```



## Redefine demographic variables 

- Gender = binary indicator for male

- Race = use white as baseline

- party5 = democrat as base level

- age4 = baseline 

- year = factor

```{r, warning = F}
nra <-
  nra %>% 
  mutate(
    Male = 
      (gender == "Male") %>% 
      as.numeric,
    party = 
      party5 %>% 
      fct_relevel("Republican") %>% 
      fct_relevel("Unkown", after = Inf),
    Year = factor(year) %>% 
      fct_relevel("2013")
  ) %>% 
  set_variable_labels(
    party = "Political Affiliation"
  )
```

```{r}
nra %>% tabyl(Male)
```

```{r}
contrasts(nra$party)
```



## Select demographic controls

```{r}
controls <-
  c(
    "Male",
    "age4",
    "educ",
    "party",
    "income4",
    "racethnicity",
    "employ"
  )
```


```{r}
nra <-
  nra %>% 
  select(
    nra, 
    Year,
    year,
    weight, 
    all_of(controls)
  )
```

2015 observations are dropped because political party is missing

```{r}
colSums(is.na(nra))
```


## Set up survey design

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = nra,
    weights = ~ weight
  )
```


## Run regression

```{r}
fml <-
  paste0(
    "nra ~ ",
    paste(controls, collapse = " + "), 
    " + Year"
  ) %>%  as.formula()
```

```{r}
mod <-
  svyglm(
    fml,
    design = ds,
    family = quasibinomial() # logit
  )
```

```{r}
summary(mod)
```

Only political party and gender seem important

## Save results

```{r}
write_rds(
  mod,
   here(
    path_od, 
    "data",
    "models",
    "nra.rds"
    )
)
```


## NRA by year

```{r}
fml <-
  paste0(
    "nra ~ ",
    paste(controls, collapse = " + ")
  ) %>%  as.formula()
```

```{r}
models <- 
  lapply(
    seq(2013, 2025, 2), 
    function(y) {
      
      if (y == 2015) {
        fml <-
          paste0(
            "nra ~ ",
            paste(controls[controls != "party"], collapse = " + ")
          )
                
      }
      
      mod <-
        svyglm(
          fml,
          design = subset(ds, year == y),
          family = quasibinomial()
          )
          
       write_rds(
        mod,
         here(
          path_od, 
          "data",
          "models",
          paste0("nra-", y, ".rds")
          )
        )
       
       mod
      })
```

```{r}


```
