# NRA Membership

This script regresses NRA membership on demographic characteristics

**Input**

-   DemographyGunOwners/data/clean/gs-gun-owners.rds

**Output**

-   DemographyGunOwners/data/raw/gunsurvey/...rds

**Last ran**

- 03/11/2026


Main specification is a pooled logistic regression with adjusted weights and a
year indicator variable.

$$
\text{logit}(P(NRA_{it} = 1)) = \beta_0 + \beta_1 X_{it} + \gamma_t + \varepsilon_{it}
$$

```{r,  message = FALSE, warning = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "tidyr",
    "janitor",
    "dplyr",
    "here",
    "survey",
    "broom",
    "kableExtra",
    "modelsummary",
    "labelled",
    "gtsummary"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
raw <-
  read_rds(
     here(
      path_od, 
      "data",
      "clean",
      "gs-gun-owners.rds"
      )
  )
```


```{r}
raw %>% tabyl(nranow)
```


## Define binary indicator for NRA membership

```{r}
gun <-
  raw %>% 
  mutate(
    nra = (nranow == "Member") %>% as.numeric %>% factor()
  )

nra <- gun %>% filter(nra == 1)
```

```{r}
gun %>% tabyl(nra)
```


## Select demographic controls

```{r}
controls <-
  c(
    "Male",
    "age4",
    "racethnicity",
    "Republican",
    "educ",
    "employ",
    "income4",
    "marital",
    "region4"
  )
```

```{r}
demovars <-
  c(
    "Male" = "Male",
    "age4" = "Age",
    "racethnicity" = "Race and ethnicity",
    "Republican" = "Republican",
    "educ" = "Highest level of education",
    "employ" = "Current employment status",
    "income4" = "Household income",

    "region4" = "Region"
  )

  
```

## Redefine demographic variables 

```{r, warning = F}
gun <-
  gun %>% 
  mutate(
    Male = 
      (gender == "Male") %>% 
      as.numeric,
    # Group because of sample size (independent + democrat)
    Republican = (party3 == "Republican") %>% 
      as.numeric,
    # Combine separated and divorced, married 
    # and Living with partner 
    marital = 
      case_when(
        marital == "Separated" ~ "Divorced",
        marital == "Living with partner" ~ "Married",
        TRUE ~ marital
      ) %>% 
      factor(
        levels =
          c(
            "Married",
            "Divorced",
            "Never married",
            "Widowed"
          )
      ),
    # group asian and other
    racethnicity = 
      case_when(
        grepl("Asian", racethnicity) ~ "Other, non-Hispanic",
        TRUE ~ racethnicity
      ) %>% 
      factor(
        levels =
          c(
            "White, non-Hispanic",
            "Black, non-Hispanic",
            "Other, non-Hispanic",
            "2+ Races, non-Hispanic",
            "Hispanic"
          )
      ),
    Year = factor(year)
  ) %>% 
  set_variable_labels(
       racethnicity = "Race and ethnicity",
       Republican = "Republican",
       marital = "Marital status"
  )
```

```{r}
gun %>% tabyl(Male)
gun %>% tabyl(Republican)
gun %>% tabyl(marital)
gun %>% tabyl(gun)
```




```{r}
gun <-
  gun %>% 
  select(
    nra, 
    Year,
    year,
    weight, 
    all_of(controls)
  ) %>% 
  filter(!(year == 2015))
```

2015 observations are dropped because political party is missing

```{r}
colSums(is.na(gun))
```

## Set up survey design

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = gun,
    weights = ~ weight
  )
```

## Run regression

Instead of rescaling, interact variables with year.

```{r}
int <-
  map(
    controls, ~ paste0(.x, "*Year")
  ) %>% 
  paste(collapse = " + ")

fml <-
  paste0(
    "nra ~ ",
    int
  ) %>%  as.formula()
```

```{r}
mod <-
  svyglm(
    fml,
    design = ds,
    family = quasibinomial() 
  )
```

```{r}
tbl_regression(mod, exponentiate = TRUE) 
```


## Save results

```{r}
write_rds(
  mod,
   here(
    path_od, 
    "data",
    "models",
    "nra.rds"
    )
)
```
