# Gun Owner analysis using general population weights

This chapter creates a demographic summary table of gun owners, but instead of using the
gun owner specific weights, we use the general population weights. This analysis is performed in 
response to Julie's comments:

How do these distributions compare to the demographic distributions of gun ownership in the weights applied for gun owner comparisons on these surveys? Are there groups that we may be systematically under-representing in more recent analyses?

The results are the same....

**Input**

-   DemographyGunOwners/data/clean/gs-gen-pop.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```



##  Load data

```{r}
genpop <-
  read_rds(
    here(
      path_od,
      "data",
      "clean",
      "gs-gen-pop.rds"
    )
  )
```

## Restrict to gun owners

```{r}
genpop <-
  genpop %>% 
  filter(gun == "Yes")
```

## Demographic variables

```{r}
demovars <-
  c(
    "gender",
    "age4",
    "racethnicity",
    "party3",
    "nranow",
    "educ",
    "employ",
    "income4",
    "marital",
    "metro",
    "region4"
  )
```

## Tabulate values

```{r}
walk(demovars, ~ print(tabyl(genpop, .x)))
```


## Yearly Summary

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = genpop, 
    strata = ~ year,
    weights = ~ weight
    ) 
```


```{r}
note <-
  paste0(
    "Note: ``Other'' gender category was not ",
    "included until the 2025 wave. ",
    "``Some college or associate’s degree'' also includes respondents ",
    "reporting vocational or technical school. ",
    "Political affiliation was not collected in the 2015 wave. ",
    "Asian, non-Hispanic was not included as a separate category until ",
    "the 2019 wave. ",
    "For marital status, ``Living with partner'' was not included ",
    "in the 2023 and 2025 waves."
  )
```

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  # add_ci() %>% 
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      
      tbl <- .x
      
      # compute unweighted yearly totals
      counts <- format(table(genpop$year), big.mark = ",")
      
      # create row with obs
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(
          as.list(counts),
          paste0("stat_", seq_along(counts))
          )
        ) %>% 
        mutate(across(starts_with("stat"), trimws)
          )

      bind_rows(n_row, tbl) %>%
        filter(
          ! grepl("Ques", label),
          ! (grepl("No", label) & variable == "nranow")
          ) %>% 
        mutate(
           label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label),
            label),
          stat_2 = ifelse(
               variable == "party5" & row_type == "level",
               "-", stat_2
               ),
           across(
             c(stat_6, stat_7),
             ~ ifelse(
               variable == "marital" & grepl("with", label),
               "-", .x
               )
             ),
          across(
            # Asian american not a category until > 2017
            c(stat_1, stat_2, stat_3),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label),
              "-", .x
              )
            ),
          # other gender category only 2025
          across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%",
              "-", .x
              )
          )
           )
    }
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "GenSummaryYear",
    linesep = "",
    caption = 
      "Demographic Composition of US Gun Owners using General Population Weights by Survey Year, 2013–2025",
    align = "lccccccc"
  ) %>%
  kable_styling(
    latex_options =
      c("hold_position", "scale_down"),
   font_size = 9
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(
    2:8, 
    width = "1.3cm"
    ) %>% 
  row_spec(1, bold = TRUE) %>% 
  footnote(
   number = note,
   threeparttable = TRUE,
   escape = FALSE,
   fixed_small_size = TRUE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/gen-pop-gun-summary-year.tex")
)
```

```{r}

```



