# Political Affiliation

This chapter explores time trends in the political affiliation of the US population.

**Input**

-   DemographyGunOwners/data/clean/gs-gen-pop.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```



##  Load data

```{r}
genpop <-
  read_rds(
    here(
      path_od,
      "data",
      "clean",
      "gs-gen-pop.rds"
    )
  )
```

## Redefine political affiliation

```{r}
genpop <- 
  genpop %>% 
  mutate(
    altparty = 
      case_when(
        grepl(
          "Strong Democrat", 
          party7, 
          ignore.case = T
          ) ~ "Democrat",
        grepl(
          "Strong Republican", 
          party7, 
          ignore.case = T
          ) ~ "Republican",
        grepl(
          "Unknown", 
          party7
        ) ~ "Unknown",
        year == 2015 ~ "Question not asked",
        TRUE ~ "Independent"
      ) %>% 
      factor()
  ) %>% 
  set_variable_labels(
    party3 = "Leaners as Partisans",
    party5 = "Strong and not so Strong as Partisans",
    altparty = "Leaners as Independents",
    party7 = "Seperate Categories"
  )
```

## Demographic variables

```{r}
demovars <-
  c(
    "party3",
    "altparty",
    "party5",
    "party7"
  )
```

## Tabulate values

```{r}
walk(demovars, ~ print(tabyl(genpop, .x)))
```


## Yearly Summary

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = genpop, 
    strata = ~ year,
    weights = ~ weight
    ) 
```
numbers are slightly off what is in Barry et al. 2018 so check survey weights and definitions

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  # add_ci() %>% 
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {
      
      tbl <- .x
      
      # compute unweighted yearly totals
      counts <- format(table(genpop$year), big.mark = ",")
      
      # create row with obs
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(
          as.list(counts),
          paste0("stat_", seq_along(counts))
          )
        ) %>% 
        mutate(across(starts_with("stat"), trimws)
          )

      bind_rows(n_row, tbl) %>%
        filter(
          ! grepl("Ques", label),
          ! (grepl("No", label) & variable == "nranow")
          ) %>% 
        mutate(
           label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", label),
            label),
          stat_2 = ifelse(
               variable == "party5" & row_type == "level",
               "-", stat_2
               ),
           across(
             c(stat_6, stat_7),
             ~ ifelse(
               variable == "marital" & grepl("with", label),
               "-", .x
               )
             ),
          across(
            # Asian american not a category until > 2017
            c(stat_1, stat_2, stat_3),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label),
              "-", .x
              )
            ),
          # other gender category only 2025
          across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%",
              "-", .x
              )
          )
           )
    }
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "GenSummaryYear",
    linesep = "",
    caption = 
      "Demographic Composition of US Adults by Survey Year, 2013–2025",
    align = "lccccccc"
  ) %>%
  kable_styling(
    latex_options =
      c("hold_position", "scale_down"),
   font_size = 9
  ) %>%
  column_spec(1, width = "4.5cm") %>%
  column_spec(
    2:8, 
    width = "1.3cm"
    ) %>% 
  row_spec(1, bold = TRUE) 
  # footnote(
  #  number = note,
  #  threeparttable = TRUE,
  #  escape = FALSE,
  #  fixed_small_size = TRUE
  # )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/political-summary-year.tex")
)
```

```{r}
