--- 
title: "Sports Betting"
author: "Joy Dada"
date: '`r format(Sys.Date(), format = "%A, %B %d, %Y")`'
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
---

# Introduction {-}

This book estimates the impact of the legalization of sports betting on gun ownership.

We use a DiD design, implemented by the `did` package, using the Callaway and Sant'Anna (2021) 
estimator.

```{r, warning = FALSE, message = FALSE, echo = F}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = TRUE
  )
```

<!--chapter:end:index.Rmd-->

# Legalisation Dates

This chapter prepares the dataset, including constructing and standardizing the dates of sports betting legalization for each state.

```{r,  message = FALSE, warning = FALSE, echo = F}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "readxl",
    "janitor",
    "here"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

Ben shared this document:
```{r}
dates <-
  read_excel(
     here(
      path_od, 
      "data",
      "raw",
      "sportsbettingdates.xlsx"
      ),
     sheet = "df"
  )
```

```{r}
dates %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Sports Betting Legalization Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```

## Tidy data

We convert the dates into the correct format and retrieve state name 
abbreviations.

```{r}
state_lookup <- setNames(state.abb, tolower(state.name))

clean <-
  dates %>% 
  rename(
    state_name = state
  ) %>% 
  mutate(
    across(
      ends_with("date"),
      ~excel_numeric_to_date(as.numeric(.x))
    ),
    state = state_lookup[tolower(trimws(state_name))]
  )
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od, 
    "data",
    "clean",
    "sportsbettingdates.rds"
    )
) 
```


<!--chapter:end:clean-dates.Rmd-->

# Create gun ownership panel

This chapter creates a state level panel of gun ownership by age and gender.

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "readxl",
    "janitor",
    "here",
    "gtsummary"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Load data

This data is seven waves of the [National Gun Policy Survey](https://publichealth.jhu.edu/center-for-gun-violence-solutions/data/national-survey-of-gun-policy) from 2013 to 2025.

We use survey weights to compute the share of individuals in a given state and year
that own a gun. This data is nationally representative.


```{r}
raw <-
  read_rds(
     here(
      path_od,
      "data",
      "clean",
      "gs-gen-pop.rds"
    )
  )
```


```{r}
raw %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey National Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```


## Create panel of gun ownership 

```{r}
des <- svydesign(
  ids = ~1,        
  weights = ~weight,
  data = raw
)

est <-
  svyby(
    ~gun,
    ~year + state,
    des,
    svymean,
    keep.var = FALSE,
    vartype = c("se", "ci"),
      level = 0.95
    ) %>% 
    clean_names()
```

## State-year level

```{r}
long <-
  est %>%
  tibble() %>% 
  rename(gun = statistic_gun_yes) %>% 
  select(
    state, year, gun
  )
```

## Check missing states

```{r}
long %>% group_by(state) %>% filter(n() != 7) %>% tabyl(state)
```

AK and WY not present every year because of small sample sizes, so we drop.

### Remove extreme values

We drop states with low effective sample sizes. This is because we cannot draw 
inference from these states. This reinforces the decision to drop observations
with only zeros and ones.

```{r}
valid_cells <-
  raw %>%
  group_by(year, state) %>%
  summarise(
    n_eff = (sum(weight)^2) / sum(weight^2),
    .groups = "drop"
  ) %>%
  filter(n_eff >= 30) %>% 
  select(-n_eff)

clean <-
  long %>%
  inner_join(valid_cells, by = c("year", "state"))
```

```{r}
n_distinct(clean$state)
```


## Save data

```{r}
write_rds(
  clean,
  here(
    path_od,
    "data",
    "clean",
    "gun-panel.rds"
  )
)
```


<!-- ## State characteristics -->

<!-- ```{r} -->
<!-- vars <- -->
<!--   c( -->
<!--     "gender", -->
<!--     "party3", -->
<!--     "metro" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ds <-  -->
<!--   svydesign( -->
<!--     ~ 1,  -->
<!--     data = raw,  -->
<!--     strata = ~ year + state, -->
<!--     weights = ~ weight -->
<!--     )  -->
<!-- ``` -->

<!-- ```{r} -->
<!-- share_df <- -->
<!--   raw %>% -->
<!--   pivot_longer( -->
<!--     all_of(vars),  -->
<!--     names_to = "variable",  -->
<!--     values_to = "level" -->
<!--     ) %>% -->
<!--   group_by(year, state, variable) %>% -->
<!--   mutate(total_w = sum(weight[!is.na(level)], na.rm = TRUE)) %>% -->
<!--   group_by(year, state, variable, level) %>% -->
<!--   summarise( -->
<!--     share = sum(weight, na.rm = TRUE) / first(total_w), -->
<!--     .groups = "drop" -->
<!--   ) %>% -->
<!--   filter( -->
<!--     level %in% c("Male", "Metro", "Republican") -->
<!--   ) %>%  -->
<!--   pivot_wider( -->
<!--     names_from = c(variable, level), -->
<!--     values_from = share -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- n_distinct(share_df$state) -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ``` -->

<!--chapter:end:clean-ownership.Rmd-->

# Create panel

This chapter merges gun survey data with the implementation dates to create
a panel for the did.

```{r}
gun <-
  read_rds(
     here(
      path_od,
      "data",
      "clean",
      "gun-panel.rds"
    )
  )
```

```{r}
dates <-
  read_rds(
    here(
      path_od, 
      "data",
      "clean",
      "sportsbettingdates.rds"
    )
  )
```

## Merge

```{r}
merge <-
  gun %>% 
  inner_join(dates, by = "state")
```

```{r}
n_distinct(merge$state)
```

## Define post treatment

These variables are defined so that they can be used with the `did` package, which
uses the Callaway and Sant'anna (2021) estimator.

```{r}
clean <-
  merge %>% 
  mutate(
    across(
      ends_with("date"),
      ~ year(.x),
      .names = "{str_replace_all(.col, 'date', 'year')}"
    ),
    across(
      ends_with("_year"),
      ~ case_when(
        !is.na(.x) ~ year >=.x,
        TRUE ~ F
      ) %>% as.numeric() %>% 
        factor(
          levels = 0:1,
          labels = c("Control", "Treatment")
        ),
      .names = "{str_replace_all(.col, 'year', 'post')}"
    ),
    # compute adjusted years so that implementation dates align with survey dates
    across(
      ends_with("_year"),
      ~ case_when(
        is.na(.x) ~ NA_real_,
        TRUE ~ .x + (.x %% 2 == 0)
      ),
      .names = "{str_replace_all(.col, 'year', 'year_adj')}"
    )
  ) %>% 
  group_by(year) %>% 
  mutate(t = cur_group_id()) %>% 
  ungroup() %>% 
  mutate(
     across(
      ends_with("_year_adj"),
      ~ case_when(
        .x == year ~ t,
        TRUE ~ NA_real_
      ),
      .names = "{str_replace_all(.col, 'year_adj', 'g')}"
    )
  ) %>% 
  group_by(state)  %>% 
   mutate(
    across(
      ends_with("g"), 
      ~ max(ifelse(is.na(.x), 0, max(.x, na.rm = T))) %>% as.numeric
      ),
    stateid = cur_group_id()
  )  %>% 
  group_by(state, year) %>% 
  mutate(cluster = cur_group_id()) %>% 
  ungroup()
```


## Check if we want a balanced panel

Balanced = Keep states that we observe for all 7 years.

```{r}
n_distinct(clean$state)
```

```{r}
balanced <-
  clean %>% 
  group_by(state) %>% 
  filter(n() == 7) %>% 
  ungroup()
```


```{r}
n_distinct(balanced$state)
```

```{r}
clean %>% tabyl(state, any_post)
```

In the balanced panel, we have 2 controls.

```{r}
balanced %>% tabyl(state, any_post)
```

## Export data

```{r}
write_rds(
  clean,
  here(
    path_od,
    "data",
    "clean",
    "betting-panel.rds"
  )
)
```


```{r}
write_rds(
  balanced,
  here(
    path_od,
    "data",
    "clean",
    "betting-panel-bal.rds"
  )
)
```

<!--chapter:end:construct-panel.Rmd-->

# Sample summary

This chapter summaries the states included in the main analysis.


```{r, echo = F}
library(kableExtra)
library(ggthemes)
```


```{r}
balanced <-
  read_rds(
    here(
    path_od,
    "data",
    "clean",
    "betting-panel-bal.rds"
    )
  ) 
```


## Sample sizes

```{r}
counts <- 
  balanced %>% 
    group_by(state) %>% 
    summarise(
       across(ends_with("_year"), max),
       n = n(),
       .groups = "drop"
    ) %>% 
  arrange(any_year) %>% 
    mutate(
      across(
        - c(state, n),
        ~ case_when(
          is.na(.x) ~ "-",
          TRUE ~ as.character(.x)
        )
      )
    ) %>% 
    bind_rows(
      balanced %>%
        mutate(state = "Overall (average)") %>% 
        group_by(state) %>% 
        summarise(
          across(
            ends_with("_year"), 
            ~ sprintf("%.0f", (mean(.x, na.rm = T)))
          ),
          n = n()
        )
    ) 
```



```{r, results = "asis"}
counts %>% 
  kbl(
    col.names = c(
      "State",
      "Any",
      "Retail",
      "Online",
      "iGaming",
      "Observations"
    ),
    format = "html",
    linesep = "",
    booktabs = TRUE,
    format.args = list(big.mark = ','),
    escape = F,
    caption = "Implementation Dates and Sample Counts"
  ) %>% 
  kable_classic(
    full_width = FALSE,
    position = "center",
    html_font = "Cambria"
    )  %>% 
  add_header_above(
    c(
      " " = 1,
      "Gambling Legalization Date" = 4,
      " " = 1
    )
  )
```

## Gun ownership

```{r, fig.cap = "Gun Ownership by State and Year"}
balanced %>% 
  ggplot(
    aes(
      x = year,
      y = gun,
      color = state
    )
  ) +
  geom_point() +
  geom_line(size = 1.15) +
  theme_event +
  theme(
    legend.title = element_blank(),
    legend.position = "bottom"
  ) +
  scale_color_economist() +
  scale_x_continuous(breaks = seq(2013, 2025, 2)) +
  scale_y_continuous(breaks = seq(0, 0.6, 0.1), limits = c(0, 0.6)) +
  labs(
    x = "Year",
    y = "Share of gun owners"
  ) +
  guides(
    color = guide_legend(
      nrow = 1,
      override.aes = list(
        linetype = 0,   
        shape = 15,    
        size = 5       
      )
    )
  )
```


## State averages

```{r, fig.cap = "Gun Ownership by State, 2013-2025"}
balanced %>% 
  bind_rows(balanced %>% mutate(state = "Overall")) %>% 
  group_by(state) %>% 
  summarise(
    across(
      gun,
      list(
        mean = ~ mean(.x),
        conf.low = ~ mean(.x) - 1.96 * sd(.x)/sqrt(7), # 7 observations per state
        conf.high = ~ mean(.x) + 1.96 * sd(.x)/sqrt(7) 
      ),
      .names = "{.fn}"
    )
  ) %>% 
  ggplot(
    aes(x = reorder(state, mean),
        y = mean,
        ymin = conf.low,
        ymax = conf.high)
  ) +
  geom_point() +
  geom_errorbar() +
  theme_event +
  scale_y_continuous(breaks = seq(0.1, 0.4, 0.05)) +
  labs(
    x = "State",
    y = "Share of gun owners"
  ) +
  coord_flip()
```


<!--chapter:end:sample-summary.Rmd-->

# DiD

This chapter runs a DiD estimation of the impact of legalization of sports betting on gun ownership at the state level.

This code currently does not include any control variables. 

```{r, echo = F}
packages <-
  c(
    "did",
    "paletteer"
  )

pacman::p_load(
  packages,
  character.only = T
)
```

## Load data

```{r}
panel <-
  read_rds(
    here(
    path_od,
    "data",
    "clean",
    "betting-panel.rds"
    )
  ) 
```

```{r}
balanced <-
  read_rds(
    here(
    path_od,
    "data",
    "clean",
    "betting-panel-bal.rds"
    )
  ) 
```


```{r}
clean %>% tabyl(state)
```

```{r}
balanced %>% tabyl(state)
```

## Simple OLS

```{r}
simple <-
  map(
    names(select(panel, ends_with("post"))),
    ~ {
      fml <- 
        paste0("gun ~ ", .x, "| state + year")  %>% 
        as.formula()
      
      fixest::feols(
          fml,
          cluster = ~ cluster,
          data = panel
        )
    }
  ) 
```

```{r}
map(simple, tbl_regression) %>% tbl_stack
```



From the simple model, post-online gambling legalization has the most significant
treatment effect. 

## Run DiD

```{r}
y_list <-  names(select(panel, ends_with("_g")))
```

```{r, warning = F}
att <-
  map(
    y_list,
    ~ att_gt(
      yname = "gun",
      tname = "t",
      idname = "stateid",
      gname = .x,
      data = balanced,
      print_details = FALSE,
      cores = 4,
      clustervars = "stateid",
      control_group = "notyettreated"
      )
  )
```

## Aggregate Group-Time Average Treatment Effects

No significant weighted average of all group-time average treatment effects.

```{r, fig.cap = "Aggregate Group-Time Average Treatment Effects by Policy"}
map(
  att,
  ~ aggte(
    .x, 
    type = "simple",
    na.rm = TRUE
    ) %>% tidy()
  ) %>% 
  bind_rows() %>% 
  mutate(
    policy = 
      y_list %>% 
      str_replace_all(
        c(
          "any_g" = "Any Sports Betting",
          "retail_g" = "Retail",
          "online_g" = "Online",
          "igaming_g" = "iGaming"
        )
      )
    ) %>% 
  ggplot(
    aes(
      x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = policy
    )
  ) +
  geom_point() +
  geom_errorbarh(width = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_event +
  labs(
    x = "Share of Gun Owners",
    y = "Legalization"
  )
```


## Event study

```{r}
events <-
  map(
    att, 
    ~ aggte(
      .x,
      type = "dynamic",
      na.rm = TRUE
      )
  )
```


```{r, warning = F}
plotdf <-
  map(
    seq_along(att),
    ~ aggte(
      att[[.x]], 
      type = "dynamic",
      na.rm = TRUE
      ) %>% 
      tidy() %>% 
      mutate(policy = y_list[[.x]])
    ) %>% 
    bind_rows() %>% 
    mutate(
      policy = 
        str_replace_all(
          policy,
          c(
            "any_g" = "Any Sports Betting",
            "retail_g" = "Retail",
            "online_g" = "Online",
            "igaming_g" = "iGaming"
          )
        )
      )
```

```{r, fig.width=8, fig.height=10}
plotdf %>% 
  ggplot(
    aes(
      x = event.time,
      y = estimate, 
      ymin = point.conf.low,
      ymax = point.conf.high
     # color = policy
      )
  ) +
  geom_point(size = 1) +
  geom_line() +
  geom_errorbar(width = 0.25) +
  geom_vline(
    xintercept =  0, 
    color = "black",
    linetype = "dashed"
    ) +
  geom_hline(
    yintercept = 0 ,
    linetype = "dashed",
    color = "black"
  ) +
  theme_event +
  scale_x_continuous(
    breaks = -5:5
    # limits = c(-5, 4)
    ) +
  scale_y_continuous(
    # limits = c(-0.2, 0.2), 
    breaks = seq(-0.2, 0.2, 0.05)
    ) +
  # scale_color_paletteer_d("MoMAColors::Althoff") +
  facet_wrap(~ policy, ncol = 1) +
  labs(
    x = "Event time",
    y = "Share of Gun Owners"
  )
```

<!--chapter:end:run-did.Rmd-->

