# Create gun ownership panel

This chapter creates a state level panel of gun ownership by age and gender.

```{r,  message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "readxl",
    "janitor",
    "here",
    "gtsummary"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Load data

This data is seven waves of the [National Gun Policy Survey](https://publichealth.jhu.edu/center-for-gun-violence-solutions/data/national-survey-of-gun-policy) from 2013 to 2025.

We use survey weights to compute the share of individuals in a given state and year
that own a gun. This data is nationally representative.


```{r}
raw <-
  read_rds(
     here(
      path_od,
      "data",
      "clean",
      "gs-gen-pop.rds"
    )
  )
```


```{r}
raw %>% 
  sample_n(10) %>% 
  head(10) %>% 
  kbl(
    caption = 
      "Gun Survey National Data",
    align = "c",
    format = "html",
    booktabs = TRUE
  ) %>% 
  kable_classic(
    full_width = FALSE,
    html_font = "Cambria"
    )  %>% 
  scroll_box(width = "800px", height = "500px")
```


## Create panel of gun ownership 

```{r}
des <- svydesign(
  ids = ~1,        
  weights = ~weight,
  data = raw
)

est <-
  svyby(
    ~gun,
    ~year + state,
    des,
    svymean,
    keep.var = FALSE,
    vartype = c("se", "ci"),
      level = 0.95
    ) %>% 
    clean_names()
```

## State-year level

```{r}
long <-
  est %>%
  tibble() %>% 
  rename(gun = statistic_gun_yes) %>% 
  select(
    state, year, gun
  )
```

## Check missing states

```{r}
long %>% group_by(state) %>% filter(n() != 7) %>% tabyl(state)
```

AK and WY not present every year because of small sample sizes, so we drop.

### Remove extreme values

We drop states with low effective sample sizes. This is because we cannot draw 
inference from these states. This reinforces the decision to drop observations
with only zeros and ones.

```{r}
valid_cells <-
  raw %>%
  group_by(year, state) %>%
  summarise(
    n_eff = (sum(weight)^2) / sum(weight^2),
    .groups = "drop"
  ) %>%
  filter(n_eff >= 30) %>% 
  select(-n_eff)

clean <-
  long %>%
  inner_join(valid_cells, by = c("year", "state"))
```

```{r}
n_distinct(clean$state)
```


## Save data

```{r}
write_rds(
  clean,
  here(
    path_od,
    "data",
    "clean",
    "gun-panel.rds"
  )
)
```


<!-- ## State characteristics -->

<!-- ```{r} -->
<!-- vars <- -->
<!--   c( -->
<!--     "gender", -->
<!--     "party3", -->
<!--     "metro" -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- ds <-  -->
<!--   svydesign( -->
<!--     ~ 1,  -->
<!--     data = raw,  -->
<!--     strata = ~ year + state, -->
<!--     weights = ~ weight -->
<!--     )  -->
<!-- ``` -->

<!-- ```{r} -->
<!-- share_df <- -->
<!--   raw %>% -->
<!--   pivot_longer( -->
<!--     all_of(vars),  -->
<!--     names_to = "variable",  -->
<!--     values_to = "level" -->
<!--     ) %>% -->
<!--   group_by(year, state, variable) %>% -->
<!--   mutate(total_w = sum(weight[!is.na(level)], na.rm = TRUE)) %>% -->
<!--   group_by(year, state, variable, level) %>% -->
<!--   summarise( -->
<!--     share = sum(weight, na.rm = TRUE) / first(total_w), -->
<!--     .groups = "drop" -->
<!--   ) %>% -->
<!--   filter( -->
<!--     level %in% c("Male", "Metro", "Republican") -->
<!--   ) %>%  -->
<!--   pivot_wider( -->
<!--     names_from = c(variable, level), -->
<!--     values_from = share -->
<!--   ) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- n_distinct(share_df$state) -->
<!-- ``` -->

<!-- ```{r} -->

<!-- ``` -->
