# DiD

This chapter runs a DiD estimation of the impact of legalization of sports betting on gun ownership at the state level.

This code currently does not include any control variables. 

```{r, echo = F}
packages <-
  c(
    "did",
    "paletteer"
  )

pacman::p_load(
  packages,
  character.only = T
)
```

## Load data

```{r}
panel <-
  read_rds(
    here(
    path_od,
    "data",
    "clean",
    "betting-panel.rds"
    )
  ) 
```

```{r}
balanced <-
  read_rds(
    here(
    path_od,
    "data",
    "clean",
    "betting-panel-bal.rds"
    )
  ) 
```


```{r}
clean %>% tabyl(state)
```

```{r}
balanced %>% tabyl(state)
```

## Simple OLS

```{r}
simple <-
  map(
    names(select(panel, ends_with("post"))),
    ~ {
      fml <- 
        paste0("gun ~ ", .x, "| state + year")  %>% 
        as.formula()
      
      fixest::feols(
          fml,
          cluster = ~ cluster,
          data = panel
        )
    }
  ) 
```

```{r}
map(simple, tbl_regression) %>% tbl_stack
```



From the simple model, post-online gambling legalization has the most significant
treatment effect. 

## Run DiD

```{r}
y_list <-  names(select(panel, ends_with("_g")))
```

```{r, warning = F}
att <-
  map(
    y_list,
    ~ att_gt(
      yname = "gun",
      tname = "t",
      idname = "stateid",
      gname = .x,
      data = balanced,
      print_details = FALSE,
      cores = 4,
      clustervars = "stateid",
      control_group = "notyettreated"
      )
  )
```

## Aggregate Group-Time Average Treatment Effects

No significant weighted average of all group-time average treatment effects.

```{r, fig.cap = "Aggregate Group-Time Average Treatment Effects by Policy"}
map(
  att,
  ~ aggte(
    .x, 
    type = "simple",
    na.rm = TRUE
    ) %>% tidy()
  ) %>% 
  bind_rows() %>% 
  mutate(
    policy = 
      y_list %>% 
      str_replace_all(
        c(
          "any_g" = "Any Sports Betting",
          "retail_g" = "Retail",
          "online_g" = "Online",
          "igaming_g" = "iGaming"
        )
      )
    ) %>% 
  ggplot(
    aes(
      x = estimate,
      xmin = conf.low,
      xmax = conf.high,
      y = policy
    )
  ) +
  geom_point() +
  geom_errorbarh(width = 0.5) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_event +
  labs(
    x = "Share of Gun Owners",
    y = "Legalization"
  )
```


## Event study

```{r}
events <-
  map(
    att, 
    ~ aggte(
      .x,
      type = "dynamic",
      na.rm = TRUE
      )
  )
```


```{r, warning = F}
plotdf <-
  map(
    seq_along(att),
    ~ aggte(
      att[[.x]], 
      type = "dynamic",
      na.rm = TRUE
      ) %>% 
      tidy() %>% 
      mutate(policy = y_list[[.x]])
    ) %>% 
    bind_rows() %>% 
    mutate(
      policy = 
        str_replace_all(
          policy,
          c(
            "any_g" = "Any Sports Betting",
            "retail_g" = "Retail",
            "online_g" = "Online",
            "igaming_g" = "iGaming"
          )
        )
      )
```

The implementation dates for retail and online gambling are the same I think.
```{r, fig.width=8, fig.height=10, fig.cap = "Event Study Estimates"}
plotdf %>% 
  ggplot(
    aes(
      x = event.time,
      y = estimate, 
      ymin = point.conf.low,
      ymax = point.conf.high
     # color = policy
      )
  ) +
  geom_point(size = 1) +
  geom_line() +
  geom_errorbar(width = 0.25) +
  geom_vline(
    xintercept =  0, 
    color = "black",
    linetype = "dashed"
    ) +
  geom_hline(
    yintercept = 0 ,
    linetype = "dashed",
    color = "black"
  ) +
  theme_event +
  scale_x_continuous(
    breaks = -5:5
    # limits = c(-5, 4)
    ) +
  scale_y_continuous(
    # limits = c(-0.2, 0.2), 
    breaks = seq(-0.2, 0.2, 0.05)
    ) +
  # scale_color_paletteer_d("MoMAColors::Althoff") +
  facet_wrap(~ policy, ncol = 1) +
  labs(
    x = "Event time",
    y = "Share of Gun Owners"
  )
```
