# Summary of Number of Exposed Children

This chapter creates a descriptive table that shows the demographics of households with children exposed to guns in the home. 

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "patchwork",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    ) |> 
  drop_na(hh18un)
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)
```


## Weighted share of exposed households with CIs

```{r}
children_exposure <- 
  svyby(
    ~gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>% 
  transmute(
    year,
    prev_pct = gunchild18un_yes,
    prev_se = se_gunchild18un_yes,
    prev_lb = ci_l_gunchild18un_yes,
    prev_ub = ci_u_gunchild18un_yes
  )
```

## Average number of children in exposed households

```{r}
avg_child <-
  svyby(
    ~hh18un,
    ~year,
    subset(ds, gunchild18un == "Yes"),
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>%
  mutate(across(-year, ~ .x)) %>%
  transmute(
    year,
    avg_child = hh18un,
    avg_child_se = se,
    avg_child_lb = ci_l,
    avg_child_ub = ci_u
  ) 
```

## Get population estimates from the ACS

I applied for an API key using this [website](https://api.census.gov/data/key_signup.html) and my jdada3@jh.edu email.

2025 estimates are not available, so we use the 2024 estimates for that year.

- [B11001_001](https://data.census.gov/table/ACSDT1Y2022.B11001)


```{r, message = F, warning = F}
hh_estimates <-
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = "B11001_001",
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
      transmute(
        year = .x,
        totalhh = estimate
      )
  )

child_estimates <-
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = "B09001_001",
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
      transmute(
        year = .x,
        children = estimate
      )
  )

census_estimates <-
  left_join(hh_estimates, child_estimates, by = "year") %>%
  mutate(
    year = if_else(year == 2024, 2025, year)
  )
```


## Combine Gun Survey and ACS estimates to compute the 95% CI of the number of children expose


```{r}
 # 2024 National Firearm Survey estimate of unsafe firearm storage
unsafe_prev <- 0.211
unsafe_lb   <- 0.183
unsafe_ub   <- 0.243
```

```{r}
exposure_estimates <-
  census_estimates %>%
  left_join(children_exposure, by = "year") %>%
  left_join(avg_child, by = "year") %>%
  mutate(

    # households with guns and children
    hh_exposed = totalhh * prev_pct,

    # children exposed to any gun
    child_est = hh_exposed * avg_child,

    # children exposed to loaded & unlocked firearms
    unsafe_child_est = child_est * unsafe_prev,
    
    # percentage of all US children exposed
    
    pct_children_exposed = child_est / children,
    pct_children_unsafe  = unsafe_child_est / children,
    

    # confidence intervals
    child_lb = totalhh * prev_lb * avg_child_lb,
    child_ub = totalhh * prev_ub * avg_child_ub,
    
    pct_children_exposed_lb = child_lb / children,
    pct_children_exposed_ub = child_ub / children,

    unsafe_child_lb = totalhh * prev_lb * avg_child_lb * unsafe_lb,
    unsafe_child_ub = totalhh * prev_ub * avg_child_ub * unsafe_ub,
    
    pct_children_unsafe_lb = unsafe_child_lb / children,
    pct_children_unsafe_ub = unsafe_child_ub / children
    
  ) %>%
  select(
    year,
    child_est,
    child_lb,
    child_ub,
    unsafe_child_est,
    unsafe_child_lb,
    unsafe_child_ub,
    pct_children_exposed,
    pct_children_exposed_lb,
    pct_children_exposed_ub,
    pct_children_unsafe,
    pct_children_unsafe_lb,
    pct_children_unsafe_ub
)
```

```{r}
exposure_estimates %>% 
  mutate(
    across(-year, ~ sjmisc::big_mark(.x))
  )
```


## Percentage of US households with guns and children in the home

```{r}
note <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns), while 2013, 2017, 2019, 2021, 2023, ",
    "and 2025 measured any guns in the home regardless of ownership. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying ",
    "the total number of US households (from the Census Bureau American Community Survey), ",
    "the prevalence of households with guns and children (survey estimate), and ",
    "the average number of children per exposed household (survey estimate). ",
    "Estimated children exposed to loaded and unlocked firearms additionally multiply ",
    "these estimates by the 2024 National Firearm Survey estimate that 21.1% ",
    "(95% CI, 18.3%–24.3%) of gun-owning households with children had at least one ",
    "firearm stored loaded and unlocked, reported by Miller et al. (2026). ",
    "Because 2025 American Community Survey estimates were not yet available, ",
    "2024 ACS household estimates were used for the 2025 calculations. ",
    "The 95 percent confidence intervals for estimated numbers of children use the conservative ",
    "multiplication rule."
  )
```

## Summary table

```{r}
# Calculate all proportion estimates with CIs
proportion_estimates <- 
  svyby(
    ~gunhousehold + child18un + gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names()

# Add observations
obs_counts <- tibble(
  year = as.numeric(names(table(child$year))),
  obs = as.numeric(table(child$year))
)

# Get unique years
years <- sort(unique(proportion_estimates$year))

# Create table structure
share_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has gun",
    "Household has child under 18",
    "Household has gun and child under 18",
    "Estimated children exposed",
    "Children exposed (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms",
    "Children exposed to loaded and unlocked firearms (\\% of US children)"
  )
)
# Add columns for each year
for (yr in years) {
  yr_data <- proportion_estimates %>% filter(year == yr)
  yr_obs <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  
  # Exposure estimates for this year
  yr_exposure <- exposure_estimates %>% filter(year == yr)
  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "M–", 
                        sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")
  
  # Gun
  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–", 
                   round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")
  
  # Child
  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "%–", 
                     round(yr_data$ci_u_child18un_yes * 100, 1), "%")
  
  # Gun and child
  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "%–", 
                        round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")
  
  # unsafe estimate
  unsafe_est <-
  paste0(
    sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)),
    " M"
  )

  unsafe_ci <-
    paste0(
      sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)),
      "M–",
      sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)),
      " M"
    )
  
  # Percent of all US children exposed
  exp_pct_est <-
    paste0(round(yr_exposure$pct_children_exposed * 100, 1), "%")
  
  exp_pct_ci <-
    paste0(
      round(yr_exposure$pct_children_exposed_lb * 100, 1),
      "%–",
      round(yr_exposure$pct_children_exposed_ub * 100, 1),
      "%"
    )
  
  unsafe_pct_est <-
    paste0(round(yr_exposure$pct_children_unsafe * 100, 1), "%")
  
  unsafe_pct_ci <-
    paste0(
      round(yr_exposure$pct_children_unsafe_lb * 100, 1),
      "%–",
      round(yr_exposure$pct_children_unsafe_ub * 100, 1),
      "%"
    )
  
  # Add columns (Est, CI pairs for this year)
  share_tbl_simple <- share_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has gun" ~ gun_est,
        characteristic == "Household has child under 18" ~ child_est,
        characteristic == "Household has gun and child under 18" ~ gunchild_est,
        characteristic == "Estimated children exposed" ~ exposure_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_est,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_est,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ 
          unsafe_pct_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has gun" ~ gun_ci,
        characteristic == "Household has child under 18" ~ child_ci,
        characteristic == "Household has gun and child under 18" ~ gunchild_ci,
        characteristic == "Estimated children exposed" ~ exposure_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_ci,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_ci,
        characteristic == "Children exposed to loaded and unlocked firearms (\\% of US children)"
        ~ unsafe_pct_ci
      )
    )
}

# Escape % for LaTeX
share_tbl_simple <- share_tbl_simple %>%
  mutate(
    across(
      starts_with("y"),
      ~ str_replace_all(., "%", "\\\\%")
    )
  )

# Create column names for header (without year labels, those go in add_header_above)
col_names <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years)))
align_string <- paste0("l", paste0(rep("c", length(years) * 2), collapse = ""))

share_tbl <- share_tbl_simple %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "Share of US Households with Guns and Children Under 18, 2013-2025",
    label = "ShareYear",
    col.names = col_names,
    linesep = "",
    align = align_string
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(
    row = 1,
    bold = T
  ) %>% 
  add_header_above(
    c(" " = 1, setNames(rep(2, length(years)), paste0(years)))
  ) %>%
  footnote(
    number = note,
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = TRUE
  )

# save_kable(
#   share_tbl,
#   file = here(path_ol, "tables/yearly-share.tex")
# )
```


## Use delta method on confidence interval

```{r}
ds_delta <- update(
  ds,
  child_exposed = ifelse(gunchild18un == "Yes", hh18un, 0)
)

## 2. A = E[children | any household]  (= p * m, with correct SE)
A_hat <- svyby(
  ~child_exposed, ~year, ds_delta, svymean,
  vartype = "se", na.rm = TRUE
) %>%
  clean_names() %>%
  transmute(year, A = child_exposed, A_se = se)

## 3. Storage prevalence: back out SE from the published CI
##    Miller et al. 2026: 21.1% (18.3-24.3%)
s_hat <- 0.211
s_se  <- (0.243 - 0.183) / (2 * qnorm(0.975))

## 4. Combine on the log scale
z <- qnorm(0.975)

exposure_delta <-
  census_estimates %>%
  left_join(A_hat, by = "year") %>%
  mutate(
    ## --- children exposed to any firearm ---
    child_est = totalhh * A,
    rel_var_A = (A_se / A)^2,
    child_lb  = child_est * exp(-z * sqrt(rel_var_A)),
    child_ub  = child_est * exp( z * sqrt(rel_var_A)),

    ## --- children exposed to loaded & unlocked ---
    unsafe_child_est = child_est * s_hat,
    rel_var_U        = rel_var_A + (s_se / s_hat)^2,
    unsafe_child_lb  = unsafe_child_est * exp(-z * sqrt(rel_var_U)),
    unsafe_child_ub  = unsafe_child_est * exp( z * sqrt(rel_var_U)),

    ## --- as share of all US children ---
    across(
      c(child_est, child_lb, child_ub,
        unsafe_child_est, unsafe_child_lb, unsafe_child_ub),
      ~ .x / children,
      .names = "pct_{.col}"
    )
  )

## 5. Sanity check against the old intervals
exposure_delta %>%
  transmute(
    year,
    est   = round(child_est / 1e6, 1),
    delta = paste0(round(child_lb / 1e6, 1), "-", round(child_ub / 1e6, 1)),
    A, A_se
  )
```


```{r}
exposure_estimates_delta <-
  exposure_delta %>%
  transmute(
    year,
    child_est,
    child_lb,
    child_ub,
    unsafe_child_est,
    unsafe_child_lb,
    unsafe_child_ub,
    pct_children_exposed    = pct_child_est,
    pct_children_exposed_lb = pct_child_lb,
    pct_children_exposed_ub = pct_child_ub,
    pct_children_unsafe     = pct_unsafe_child_est,
    pct_children_unsafe_lb  = pct_unsafe_child_lb,
    pct_children_unsafe_ub  = pct_unsafe_child_ub
  ) %>%
  filter(!is.na(child_est))


note_delta <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns) and was excluded from the analysis. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying ",
    "the total number of US households (from the Census Bureau American Community Survey) ",
    "by the weighted mean number of children living in households with both a gun and a ",
    "child under 18 (survey estimate), which is algebraically equal to the product of the ",
    "prevalence of such households and the mean number of children per such household. ",
    "Estimated children exposed to loaded and unlocked firearms additionally multiply ",
    "these estimates by the 2024 National Firearm Survey estimate that 21.1\\% ",
    "(95\\% CI, 18.3\\%–24.3\\%) of gun-owning households with children had at least one ",
    "firearm stored loaded and unlocked, reported by Miller et al. (2026). ",
    "Because 2025 American Community Survey estimates were not yet available, ",
    "2024 ACS household estimates were used for the 2025 calculations. ",
    "Confidence intervals for estimated numbers of children combine design-based ",
    "variance for the survey component with the published variance of the storage ",
    "estimate using the delta method on the log scale, treating American Community ",
    "Survey totals as fixed."
  )
```

```{r}
years_delta <- sort(unique(exposure_estimates_delta$year))

share_tbl_delta_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has gun",
    "Household has child under 18",
    "Household has gun and child under 18",
    "Estimated children exposed to firearms in household",
    "Children exposed (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms",
    "Children exposed to loaded and unlocked firearms (\\% of US children)"
  )
)

for (yr in years_delta) {

  yr_data     <- proportion_estimates       %>% filter(year == yr)
  yr_obs      <- obs_counts                 %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_exposure <- exposure_estimates_delta   %>% filter(year == yr)

  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "M–",
                         sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")

  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci  <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–",
                    round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")

  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci  <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "%–",
                      round(yr_data$ci_u_child18un_yes * 100, 1), "%")

  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci  <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "%–",
                         round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")

  unsafe_est <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)), " M")
  unsafe_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)), "M–",
                       sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)), " M")

  exp_pct_est <- paste0(round(yr_exposure$pct_children_exposed * 100, 1), "%")
  exp_pct_ci  <- paste0(round(yr_exposure$pct_children_exposed_lb * 100, 1), "%–",
                        round(yr_exposure$pct_children_exposed_ub * 100, 1), "%")

  unsafe_pct_est <- paste0(round(yr_exposure$pct_children_unsafe * 100, 1), "%")
  unsafe_pct_ci  <- paste0(round(yr_exposure$pct_children_unsafe_lb * 100, 1), "%–",
                           round(yr_exposure$pct_children_unsafe_ub * 100, 1), "%")

  share_tbl_delta_simple <- share_tbl_delta_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has gun" ~ gun_est,
        characteristic == "Household has child under 18" ~ child_est,
        characteristic == "Household has gun and child under 18" ~ gunchild_est,
        characteristic == "Estimated children exposed to firearms in household" ~ exposure_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_est,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_est,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ unsafe_pct_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has gun" ~ gun_ci,
        characteristic == "Household has child under 18" ~ child_ci,
        characteristic == "Household has gun and child under 18" ~ gunchild_ci,
        characteristic == "Estimated children exposed to firearms in household" ~ exposure_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_ci,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_ci,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ unsafe_pct_ci
      )
    )
}

share_tbl_delta_simple <- share_tbl_delta_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

col_names_delta   <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years_delta)))
align_string_delta <- paste0("l", paste0(rep("c", length(years_delta) * 2), collapse = ""))

share_tbl_delta <- share_tbl_delta_simple %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Share of US Households with Guns and Children Under 18, 2013-2025",
    label     = "ShareYearDelta",
    col.names = col_names_delta,
    linesep   = "",
    align     = align_string_delta
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  row_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, setNames(rep(2, length(years_delta)), paste0(years_delta)))) %>%
  footnote(number = note_delta, threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(
  share_tbl_delta,
  file = here(path_ol, "tables/yearly-share-delta.tex")
)
```

```{r}
left_join(
  exposure_estimates       %>% transmute(year, old = paste0(round(child_lb/1e6,1), "–", round(child_ub/1e6,1))),
  exposure_estimates_delta %>% transmute(year, delta = paste0(round(child_lb/1e6,1), "–", round(child_ub/1e6,1))),
  by = "year"
)
```


## Sensitivity analysis

Project project 2025 ACS totals from the 2013-2024 series instead of substituting the 2024 values

We fit the data using a log linear trend. 

```{R}
acs_series <- census_estimates %>%
  mutate(acs_year = if_else(year == 2025, 2024, year))

fit_hh    <- lm(log(totalhh)  ~ acs_year, data = acs_series)
fit_child <- lm(log(children) ~ acs_year, data = acs_series)

proj_2025 <- tibble(
  totalhh_proj  = exp(predict(fit_hh,    newdata = tibble(acs_year = 2025))),
  children_proj = exp(predict(fit_child, newdata = tibble(acs_year = 2025)))
)

growth_hh    <- exp(coef(fit_hh)["acs_year"])    - 1
growth_child <- exp(coef(fit_child)["acs_year"]) - 1

tibble(
  quantity = c("Households", "Children under 18"),
  acs_2024 = c(acs_series$totalhh[acs_series$acs_year == 2024],
               acs_series$children[acs_series$acs_year == 2024]),
  proj_2025 = c(proj_2025$totalhh_proj, proj_2025$children_proj),
  annual_growth = c(growth_hh, growth_child)
)
```


```{r}

A_2025 <- A_hat %>% filter(year == 2025)

recompute <- function(N_hh, N_child, A, A_se, s = s_hat, s_sd = s_se) {
  z         <- qnorm(0.975)
  rel_var_A <- (A_se / A)^2
  rel_var_U <- rel_var_A + (s_sd / s)^2

  tibble(
    child_est        = N_hh * A,
    child_lb         = N_hh * A * exp(-z * sqrt(rel_var_A)),
    child_ub         = N_hh * A * exp( z * sqrt(rel_var_A)),
    unsafe_child_est = N_hh * A * s,
    unsafe_child_lb  = N_hh * A * s * exp(-z * sqrt(rel_var_U)),
    unsafe_child_ub  = N_hh * A * s * exp( z * sqrt(rel_var_U))
  ) %>%
    mutate(
      across(everything(), ~ .x / N_child, .names = "pct_{.col}"),
      totalhh  = N_hh,
      children = N_child
    )
}

sens_2025 <- bind_rows(
  recompute(
    N_hh    = census_estimates$totalhh[census_estimates$year == 2025],
    N_child = census_estimates$children[census_estimates$year == 2025],
    A       = A_2025$A,
    A_se    = A_2025$A_se
  ) %>% mutate(spec = "2024 ACS estimates"),

  recompute(
    N_hh    = proj_2025$totalhh_proj,
    N_child = proj_2025$children_proj,
    A       = A_2025$A,
    A_se    = A_2025$A_se
  ) %>% mutate(spec = "Projected 2025 totals")
) %>%
  select(spec, everything())
```

```{r}
fmt_m   <- function(x) paste0(sjmisc::big_mark(round(x / 1e6, 1)), " M")
fmt_pct <- function(x) paste0(round(x * 100, 1), "\\%")

sens_tbl <- tibble(
  characteristic = c(
    "Total US households",
    "US population under 18",
    "Estimated children exposed",
    "Children exposed (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms",
    "Children exposed to loaded and unlocked firearms (\\% of US children)"
  )
)

for (sp in unique(sens_2025$spec)) {
  d <- sens_2025 %>% filter(spec == sp)

  sens_tbl <- sens_tbl %>%
    mutate(
      !!paste0(make.names(sp), "_est") := c(
        fmt_m(d$totalhh),
        fmt_m(d$children),
        fmt_m(d$child_est),
        fmt_pct(d$pct_child_est),
        fmt_m(d$unsafe_child_est),
        fmt_pct(d$pct_unsafe_child_est)
      ),
      !!paste0(make.names(sp), "_ci") := c(
        "", "",
        paste0(fmt_m(d$child_lb), "–", fmt_m(d$child_ub)),
        paste0(fmt_pct(d$pct_child_lb), "–", fmt_pct(d$pct_child_ub)),
        paste0(fmt_m(d$unsafe_child_lb), "–", fmt_m(d$unsafe_child_ub)),
        paste0(fmt_pct(d$pct_unsafe_child_lb), "-", fmt_pct(d$pct_unsafe_child_ub))
      )
    )
}

sens_note <- paste0(
  "Table compares estimated numbers of US children exposed to firearms in the home ",
  "in 2025 under two assumptions about American Community Survey totals, United States, 2025. ",
  "The first column pair substitutes 2024 American Community Survey estimates, which ",
  "is the specification used in the main analysis because 2025 estimates were not yet ",
  "available. The second projects 2025 totals by fitting a log-linear trend to the ",
  "2013--2024 American Community Survey series separately for total households and for ",
  "the population younger than 18 years, implying annual growth of ",
  sprintf("%.2f\\%%", growth_hh * 100), " in households and ",
  sprintf("%.2f\\%%", growth_child * 100), " in the child population. ",
  "Survey-based inputs are identical across columns; only the American Community Survey ",
  "totals differ. Confidence intervals use the delta method on the log scale and are ",
  "therefore of identical relative width in both columns."
)

sens_latex <- sens_tbl %>%
  kable(
    format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
    caption = "Sensitivity of 2025 Estimates to American Community Survey Totals",
    label = "SensACS",
    col.names = c("Characteristic", rep(c("Est.", "95\\% CI"), 2)),
    align = "lcccc"
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  add_header_above(
    c(" " = 1, "2024 ACS estimates" = 2, "Projected 2025 totals" = 2)
  ) %>%
  footnote(number = sens_note, threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(sens_latex, file = here(path_ol, "tables", "sensitivity-acs.tex"))
```


## Plot over time


```{r}
prev_long <-
  proportion_estimates %>%
  transmute(
    year,
    `Firearm in the home`        = gunhousehold_yes,
    `Firearm and child under 18` = gunchild18un_yes,
    lb_gun = ci_l_gunhousehold_yes, ub_gun = ci_u_gunhousehold_yes,
    lb_bth = ci_l_gunchild18un_yes, ub_bth = ci_u_gunchild18un_yes
  ) %>%
  pivot_longer(c(`Firearm in the home`, `Firearm and child under 18`),
               names_to = "measure", values_to = "est") %>%
  mutate(
    lb = if_else(measure == "Firearm in the home", lb_gun, lb_bth),
    ub = if_else(measure == "Firearm in the home", ub_gun, ub_bth),
    across(c(est, lb, ub), ~ .x * 100),
    measure = factor(measure, levels = c("Firearm in the home",
                                         "Firearm and child under 18"))
  ) %>%
  filter(year != 2015)

hh_cols <- c("Firearm in the home"        = "#1f4e5f",
             "Firearm and child under 18" = "#e9a13b")

## Panel A
prev_long <- prev_long %>% mutate(year_f = factor(year))

p_prev <-
  ggplot(prev_long, aes(year_f, est, color = measure, fill = measure,
                        group = measure)) +
  geom_ribbon(aes(ymin = lb, ymax = ub), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  # geom_text(
  #   data = prev_long %>% filter(year == max(year)),
  #   aes(label = measure), hjust = 1, vjust = -1.5,
  #   size = 3.5, family = "serif", show.legend = FALSE
  # ) +
  scale_color_manual(values = hh_cols) +
  scale_fill_manual(values = hh_cols) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0.08, 0.12))) +
  labs(x = NULL, y = "Share of US households") +
  theme +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "bottom",
        legend.title = element_blank(),
        text = element_text(size = 14 ),
        legend.text = element_text(size = 12)
        )

## Panel B

exp_df <-
  exposure_delta %>%
  filter(!is.na(child_est)) %>%
  transmute(year,
            est = pct_child_est * 100,
            lb  = pct_child_lb  * 100,
            ub  = pct_child_ub  * 100,
            year_f = factor(year))

p_exp <-
  ggplot(exp_df, aes(year_f, est, group = 1)) +
  geom_ribbon(aes(ymin = lb, ymax = ub), fill = "#1f4e5f", alpha = 0.15) +
  geom_line(color = "#1f4e5f", linewidth = 0.9) +
  geom_point(color = "#1f4e5f", size = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0.08, 0.12))) +
  labs(x = NULL, y = "Share of US children living\nin a household with a firearm") +
  theme +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = 14 ))

main_results_plot <- p_prev / p_exp

ggsave(
  here(path_ol, "figures", "household-prevalence-over-time.png"),
  p_prev, width = 8, height = 4, dpi = 300
)

ggsave(
  here(path_ol, "figures", "children-exposed-over-time.png"),
  p_exp, width = 8, height = 4, dpi = 300
)
```


## Appendix table: calculation inputs

Seeing the all of the raw inputs allows us to know what is driving the increase.

```{R}
inputs_years <- sort(unique(A_hat$year))

inputs_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Total US households",
    "Household has gun and child under 18",
    "Mean children per exposed household",
    "Mean exposed children per US household",
    "Estimated children exposed",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

for (yr in inputs_years) {

  yr_obs      <- obs_counts        %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_hh       <- census_estimates  %>% filter(year == yr)
  yr_prev     <- children_exposure %>% filter(year == yr)
  yr_m        <- avg_child         %>% filter(year == yr)
  yr_A        <- A_hat             %>% filter(year == yr)
  yr_exposure <- exposure_delta    %>% filter(year == yr)

  hh_est <- paste0(sjmisc::big_mark(round(yr_hh$totalhh / 1e6, 1)), " M")

  gc_est <- paste0(round(yr_prev$prev_pct * 100, 1), "%")
  gc_ci  <- paste0(round(yr_prev$prev_lb * 100, 1), "–",
                   round(yr_prev$prev_ub * 100, 1), "%")

  m_est <- sprintf("%.2f", yr_m$avg_child)
  m_ci  <- paste0(sprintf("%.2f", yr_m$avg_child_lb), "–",
                  sprintf("%.2f", yr_m$avg_child_ub))

  A_est <- sprintf("%.3f", yr_A$A)
  A_ci  <- paste0(sprintf("%.3f", yr_A$A - qnorm(0.975) * yr_A$A_se), "–",
                  sprintf("%.3f", yr_A$A + qnorm(0.975) * yr_A$A_se))

  exp_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exp_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "–",
                    sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")

  lu_est <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)), " M")
  lu_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)), "–",
                   sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)), " M")

  inputs_tbl_simple <- inputs_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Total US households" ~ hh_est,
        characteristic == "Household has gun and child under 18" ~ gc_est,
        characteristic == "Mean children per exposed household" ~ m_est,
        characteristic == "Mean exposed children per US household" ~ A_est,
        characteristic == "Estimated children exposed" ~ exp_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Total US households" ~ "",
        characteristic == "Household has gun and child under 18" ~ gc_ci,
        characteristic == "Mean children per exposed household" ~ m_ci,
        characteristic == "Mean exposed children per US household" ~ A_ci,
        characteristic == "Estimated children exposed" ~ exp_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_ci
      )
    )
}

inputs_tbl_simple <- inputs_tbl_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

col_names_inputs <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(inputs_years)))
align_inputs     <- paste0("l", paste0(rep("c", length(inputs_years) * 2), collapse = ""))

inputs_note <-
  paste0(
    "Table reports every quantity entering the estimated number of US children ",
    "exposed to firearms in the home, by survey wave, United States, 2013--2025. ",
    "Estimated children exposed = total US households (American Community Survey) ",
    "$\\times$ mean exposed children per US household (survey). The latter is ",
    "algebraically equal to the prevalence of households with a gun and a child ",
    "under 18 multiplied by the mean number of children in those households, both ",
    "of which are reported above so that changes across waves can be attributed to ",
    "each component; it is estimated directly so that its standard error accounts ",
    "for the covariance between them. Estimated children exposed to loaded and ",
    "unlocked firearms additionally multiplies by the 2024 National Firearm Survey ",
    "estimate that 21.1\\% (95\\% CI, 18.3\\%--24.3\\%) of gun-owning households with ",
    "children had at least one firearm stored loaded and unlocked (Miller et al.\\ 2026). ",
    "Total US household counts are American Community Survey 1-year estimates; 2024 ",
    "estimates were used for 2025. Confidence intervals for the derived counts use ",
    "the delta method."
  )

inputs_share_tbl <- inputs_tbl_simple %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Inputs to the Estimated Number of US Children Exposed to Firearms in the Home, 2013-2025",
    label     = "ExposureInputs",
    col.names = col_names_inputs,
    linesep   = "",
    align     = align_inputs
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  add_header_above(
    c(" " = 1, setNames(rep(2, length(inputs_years)), paste0(inputs_years)))
  ) %>%
  footnote(number = inputs_note, threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(inputs_share_tbl, file = here(path_ol, "tables/exposure-inputs.tex"))
```

## Test: does the association between having a firearm and a child in the household change over time

```{r}

ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)

## P(gun | child) by year, with CIs
gun_by_child <- svyby(
  ~gunhousehold, ~year + child18un, ds, svymean,
  vartype = c("se", "ci")
) %>% clean_names()

## Formal test: has the child-firearm association changed over time?
ds_ends <- subset(ds, year %in% c(2013, 2025))

m <- svyglm(
  I(gunhousehold == "Yes") ~ child18un * factor(year),
  design = ds_ends, family = quasibinomial()
)
summary(m)
```

