# Summary of Number of Exposed Children

This chapter creates a descriptive table that shows the demographics of households with children exposed to guns in the home. 

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "patchwork",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"

## kableExtra::footnote(threeparttable = TRUE) strips one level of backslashes
## from the note text even when escape = FALSE, so "\\%" reaches the .tex file
## as a bare "%", which comments out the rest of the footnote, and "$\\times$"
## arrives as "$times$". Escape any bare percent signs, then double every
## backslash so one survives.
esc_note <- function(x) {
  x %>%
    str_replace_all("(?<!\\\\)%", "\\\\%") %>%
    str_replace_all(fixed("\\"), "\\\\")
}
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    ) |> 
  drop_na(hh18un)
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)
```


## Weighted share of exposed households with CIs

```{r}
children_exposure <- 
  svyby(
    ~gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>% 
  transmute(
    year,
    prev_pct = gunchild18un_yes,
    prev_se = se_gunchild18un_yes,
    prev_lb = ci_l_gunchild18un_yes,
    prev_ub = ci_u_gunchild18un_yes
  )
```

## Average number of children in exposed households

```{r}
avg_child <-
  svyby(
    ~hh18un,
    ~year,
    subset(ds, gunchild18un == "Yes"),
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>%
  mutate(across(-year, ~ .x)) %>%
  transmute(
    year,
    avg_child = hh18un,
    avg_child_se = se,
    avg_child_lb = ci_l,
    avg_child_ub = ci_u
  ) 
```

## Get population estimates from the ACS

I applied for an API key using this [website](https://api.census.gov/data/key_signup.html) and my jdada3@jh.edu email.

2025 estimates are not available, so we use the 2024 estimates for that year.

- [B11001_001](https://data.census.gov/table/ACSDT1Y2022.B11001)


```{r, message = F, warning = F}
hh_estimates <-
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = "B11001_001",
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
      transmute(
        year = .x,
        totalhh = estimate
      )
  )

child_estimates <-
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = "B09001_001",
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
      transmute(
        year = .x,
        children = estimate
      )
  )

census_estimates <-
  left_join(hh_estimates, child_estimates, by = "year") %>%
  mutate(
    year = if_else(year == 2024, 2025, year)
  )
```


## Combine Gun Survey and ACS estimates to compute the 95% CI of the number of children expose


```{r}
 # 2024 National Firearm Survey estimate of unsafe firearm storage
unsafe_prev <- 0.211
unsafe_lb   <- 0.183
unsafe_ub   <- 0.243
```

```{r}
exposure_estimates <-
  census_estimates %>%
  left_join(children_exposure, by = "year") %>%
  left_join(avg_child, by = "year") %>%
  mutate(

    # households with guns and children
    hh_exposed = totalhh * prev_pct,

    # children exposed to any gun
    child_est = hh_exposed * avg_child,

    # children exposed to loaded & unlocked firearms
    unsafe_child_est = child_est * unsafe_prev,
    
    # percentage of all US children exposed
    
    pct_children_exposed = child_est / children,
    pct_children_unsafe  = unsafe_child_est / children,
    

    # confidence intervals
    child_lb = totalhh * prev_lb * avg_child_lb,
    child_ub = totalhh * prev_ub * avg_child_ub,
    
    pct_children_exposed_lb = child_lb / children,
    pct_children_exposed_ub = child_ub / children,

    unsafe_child_lb = totalhh * prev_lb * avg_child_lb * unsafe_lb,
    unsafe_child_ub = totalhh * prev_ub * avg_child_ub * unsafe_ub,
    
    pct_children_unsafe_lb = unsafe_child_lb / children,
    pct_children_unsafe_ub = unsafe_child_ub / children
    
  ) %>%
  select(
    year,
    child_est,
    child_lb,
    child_ub,
    unsafe_child_est,
    unsafe_child_lb,
    unsafe_child_ub,
    pct_children_exposed,
    pct_children_exposed_lb,
    pct_children_exposed_ub,
    pct_children_unsafe,
    pct_children_unsafe_lb,
    pct_children_unsafe_ub
)
```

```{r}
exposure_estimates %>% 
  mutate(
    across(-year, ~ sjmisc::big_mark(.x))
  )
```


## Percentage of US households with guns and children in the home

```{r}
note <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns), while 2013, 2017, 2019, 2021, 2023, ",
    "and 2025 measured any guns in the home regardless of ownership. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying ",
    "the total number of US households (from the Census Bureau American Community Survey), ",
    "the prevalence of households with guns and children (survey estimate), and ",
    "the average number of children per exposed household (survey estimate). ",
    "Estimated children exposed to loaded and unlocked firearms additionally multiply ",
    "these estimates by the 2024 National Firearm Survey estimate that 21.1% ",
    "(95% CI, 18.3%–24.3%) of gun-owning households with children had at least one ",
    "firearm stored loaded and unlocked, reported by Miller et al. (2026). ",
    "Because 2025 American Community Survey estimates were not yet available, ",
    "2024 ACS household estimates were used for the 2025 calculations. ",
    "The 95 percent confidence intervals for estimated numbers of children use the conservative ",
    "multiplication rule."
  )
```

## Summary table

```{r}
# Calculate all proportion estimates with CIs
proportion_estimates <- 
  svyby(
    ~gunhousehold + child18un + gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names()

# Add observations
obs_counts <- tibble(
  year = as.numeric(names(table(child$year))),
  obs = as.numeric(table(child$year))
)

# Get unique years
years <- sort(unique(proportion_estimates$year))

# Create table structure
# Row order is shared with the delta, ratio and post-stratified tables:
# percentage first, then the corresponding estimated count.
share_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has a firearm",
    "Household has a child under 18",
    "Household has a firearm and a child under 18",
    "Children exposed (\\% of US children)",
    "Estimated children exposed",
    "Children exposed to loaded and unlocked firearms (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)
# Add columns for each year
for (yr in years) {
  yr_data <- proportion_estimates %>% filter(year == yr)
  yr_obs <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  
  # Exposure estimates for this year
  yr_exposure <- exposure_estimates %>% filter(year == yr)
  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "M–", 
                        sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")
  
  # Gun
  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–", 
                   round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")
  
  # Child
  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "%–", 
                     round(yr_data$ci_u_child18un_yes * 100, 1), "%")
  
  # Gun and child
  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "%–", 
                        round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")
  
  # unsafe estimate
  unsafe_est <-
  paste0(
    sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)),
    " M"
  )

  unsafe_ci <-
    paste0(
      sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)),
      "M–",
      sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)),
      " M"
    )
  
  # Percent of all US children exposed
  exp_pct_est <-
    paste0(round(yr_exposure$pct_children_exposed * 100, 1), "%")
  
  exp_pct_ci <-
    paste0(
      round(yr_exposure$pct_children_exposed_lb * 100, 1),
      "%–",
      round(yr_exposure$pct_children_exposed_ub * 100, 1),
      "%"
    )
  
  unsafe_pct_est <-
    paste0(round(yr_exposure$pct_children_unsafe * 100, 1), "%")
  
  unsafe_pct_ci <-
    paste0(
      round(yr_exposure$pct_children_unsafe_lb * 100, 1),
      "%–",
      round(yr_exposure$pct_children_unsafe_ub * 100, 1),
      "%"
    )
  
  # Add columns (Est, CI pairs for this year)
  share_tbl_simple <- share_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has a firearm" ~ gun_est,
        characteristic == "Household has a child under 18" ~ child_est,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_est,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_est,
        characteristic == "Estimated children exposed" ~ exposure_est,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~
          unsafe_pct_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has a firearm" ~ gun_ci,
        characteristic == "Household has a child under 18" ~ child_ci,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_ci,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_ci,
        characteristic == "Estimated children exposed" ~ exposure_ci,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~
          unsafe_pct_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_ci
      )
    )
}

# Escape % for LaTeX
share_tbl_simple <- share_tbl_simple %>%
  mutate(
    across(
      starts_with("y"),
      ~ str_replace_all(., "%", "\\\\%")
    )
  )

# Create column names for header (without year labels, those go in add_header_above)
col_names <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years)))
align_string <- paste0("l", paste0(rep("c", length(years) * 2), collapse = ""))

share_tbl <- share_tbl_simple %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "Household Firearm Prevalence and Children's Firearm Exposure, 2013-2025",
    label = "ShareYear",
    col.names = col_names,
    linesep = "",
    align = align_string
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(
    row = 1,
    bold = T
  ) %>% 
  add_header_above(
    c(" " = 1, setNames(rep(2, length(years)), paste0(years)))
  ) %>%
  footnote(
    number = esc_note(note),
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  share_tbl,
  file = here(path_ol, "tables/yearly-share.tex")
)
```


## Use delta method on confidence interval

```{r}
ds_delta <- update(
  ds,
  child_exposed = ifelse(gunchild18un == "Yes", hh18un, 0)
)

## 2. A = E[children | any household]  (= p * m, with correct SE)
A_hat <- svyby(
  ~child_exposed, ~year, ds_delta, svymean,
  vartype = "se", na.rm = TRUE
) %>%
  clean_names() %>%
  transmute(year, A = child_exposed, A_se = se)

## 3. Storage prevalence: back out SE from the published CI
##    Miller et al. 2026: 21.1% (18.3-24.3%)
s_hat <- 0.211
s_se  <- (0.243 - 0.183) / (2 * qnorm(0.975))

## 4. Combine on the log scale
z <- qnorm(0.975)

exposure_delta <-
  census_estimates %>%
  left_join(A_hat, by = "year") %>%
  mutate(
    ## --- children exposed to any firearm ---
    child_est = totalhh * A,
    rel_var_A = (A_se / A)^2,
    child_lb  = child_est * exp(-z * sqrt(rel_var_A)),
    child_ub  = child_est * exp( z * sqrt(rel_var_A)),

    ## --- children exposed to loaded & unlocked ---
    unsafe_child_est = child_est * s_hat,
    rel_var_U        = rel_var_A + (s_se / s_hat)^2,
    unsafe_child_lb  = unsafe_child_est * exp(-z * sqrt(rel_var_U)),
    unsafe_child_ub  = unsafe_child_est * exp( z * sqrt(rel_var_U)),

    ## --- as share of all US children ---
    across(
      c(child_est, child_lb, child_ub,
        unsafe_child_est, unsafe_child_lb, unsafe_child_ub),
      ~ .x / children,
      .names = "pct_{.col}"
    )
  )

## 5. Sanity check against the old intervals
exposure_delta %>%
  transmute(
    year,
    est   = round(child_est / 1e6, 1),
    delta = paste0(round(child_lb / 1e6, 1), "-", round(child_ub / 1e6, 1)),
    A, A_se
  )
```


```{r}
exposure_estimates_delta <-
  exposure_delta %>%
  transmute(
    year,
    child_est,
    child_lb,
    child_ub,
    unsafe_child_est,
    unsafe_child_lb,
    unsafe_child_ub,
    pct_children_exposed    = pct_child_est,
    pct_children_exposed_lb = pct_child_lb,
    pct_children_exposed_ub = pct_child_ub,
    pct_children_unsafe     = pct_unsafe_child_est,
    pct_children_unsafe_lb  = pct_unsafe_child_lb,
    pct_children_unsafe_ub  = pct_unsafe_child_ub
  ) %>%
  filter(!is.na(child_est))


note_delta <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns) and was excluded from the analysis. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying ",
    "the total number of US households (from the Census Bureau American Community Survey) ",
    "by the weighted mean number of children living in households with both a gun and a ",
    "child under 18 (survey estimate), which is algebraically equal to the product of the ",
    "prevalence of such households and the mean number of children per such household. ",
    "Estimated children exposed to loaded and unlocked firearms additionally multiply ",
    "these estimates by the 2024 National Firearm Survey estimate that 21.1\\% ",
    "(95\\% CI, 18.3\\%–24.3\\%) of gun-owning households with children had at least one ",
    "firearm stored loaded and unlocked, reported by Miller et al. (2026). ",
    "Because 2025 American Community Survey estimates were not yet available, ",
    "2024 ACS household estimates were used for the 2025 calculations. ",
    "Confidence intervals for estimated numbers of children combine design-based ",
    "variance for the survey component with the published variance of the storage ",
    "estimate using the delta method on the log scale, treating American Community ",
    "Survey totals as fixed."
  )
```

```{r}
years_delta <- sort(unique(exposure_estimates_delta$year))

share_tbl_delta_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has a firearm",
    "Household has a child under 18",
    "Household has a firearm and a child under 18",
    "Children exposed (\\% of US children)",
    "Estimated children exposed",
    "Children exposed to loaded and unlocked firearms (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

for (yr in years_delta) {

  yr_data     <- proportion_estimates       %>% filter(year == yr)
  yr_obs      <- obs_counts                 %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_exposure <- exposure_estimates_delta   %>% filter(year == yr)

  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "M–",
                         sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")

  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci  <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–",
                    round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")

  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci  <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "%–",
                      round(yr_data$ci_u_child18un_yes * 100, 1), "%")

  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci  <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "%–",
                         round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")

  unsafe_est <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)), " M")
  unsafe_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)), "M–",
                       sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)), " M")

  exp_pct_est <- paste0(round(yr_exposure$pct_children_exposed * 100, 1), "%")
  exp_pct_ci  <- paste0(round(yr_exposure$pct_children_exposed_lb * 100, 1), "%–",
                        round(yr_exposure$pct_children_exposed_ub * 100, 1), "%")

  unsafe_pct_est <- paste0(round(yr_exposure$pct_children_unsafe * 100, 1), "%")
  unsafe_pct_ci  <- paste0(round(yr_exposure$pct_children_unsafe_lb * 100, 1), "%–",
                           round(yr_exposure$pct_children_unsafe_ub * 100, 1), "%")

  share_tbl_delta_simple <- share_tbl_delta_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has a firearm" ~ gun_est,
        characteristic == "Household has a child under 18" ~ child_est,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_est,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_est,
        characteristic == "Estimated children exposed" ~ exposure_est,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ unsafe_pct_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has a firearm" ~ gun_ci,
        characteristic == "Household has a child under 18" ~ child_ci,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_ci,
        characteristic == "Children exposed (\\% of US children)" ~ exp_pct_ci,
        characteristic == "Estimated children exposed" ~ exposure_ci,
        characteristic ==
          "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ unsafe_pct_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_ci
      )
    )
}

share_tbl_delta_simple <- share_tbl_delta_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

col_names_delta   <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years_delta)))
align_string_delta <- paste0("l", paste0(rep("c", length(years_delta) * 2), collapse = ""))

share_tbl_delta <- share_tbl_delta_simple %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Household Firearm Prevalence and Children's Firearm Exposure, 2013-2025",
    label     = "ShareYearDelta",
    col.names = col_names_delta,
    linesep   = "",
    align     = align_string_delta
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  row_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, setNames(rep(2, length(years_delta)), paste0(years_delta)))) %>%
  footnote(number = esc_note(note_delta), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(
  share_tbl_delta,
  file = here(path_ol, "tables/yearly-share-delta.tex")
)
```

```{r}
left_join(
  exposure_estimates       %>% transmute(year, old = paste0(round(child_lb/1e6,1), "–", round(child_ub/1e6,1))),
  exposure_estimates_delta %>% transmute(year, delta = paste0(round(child_lb/1e6,1), "–", round(child_ub/1e6,1))),
  by = "year"
)
```


## Sensitivity analysis

Project project 2025 ACS totals from the 2013-2024 series instead of substituting the 2024 values

We fit the data using a log linear trend. 

```{R}
acs_series <- census_estimates %>%
  mutate(acs_year = if_else(year == 2025, 2024, year))

fit_hh    <- lm(log(totalhh)  ~ acs_year, data = acs_series)
fit_child <- lm(log(children) ~ acs_year, data = acs_series)

proj_2025 <- tibble(
  totalhh_proj  = exp(predict(fit_hh,    newdata = tibble(acs_year = 2025))),
  children_proj = exp(predict(fit_child, newdata = tibble(acs_year = 2025)))
)

growth_hh    <- exp(coef(fit_hh)["acs_year"])    - 1
growth_child <- exp(coef(fit_child)["acs_year"]) - 1

tibble(
  quantity = c("Households", "Children under 18"),
  acs_2024 = c(acs_series$totalhh[acs_series$acs_year == 2024],
               acs_series$children[acs_series$acs_year == 2024]),
  proj_2025 = c(proj_2025$totalhh_proj, proj_2025$children_proj),
  annual_growth = c(growth_hh, growth_child)
)
```


```{r}

A_2025 <- A_hat %>% filter(year == 2025)

recompute <- function(N_hh, N_child, A, A_se, s = s_hat, s_sd = s_se) {
  z         <- qnorm(0.975)
  rel_var_A <- (A_se / A)^2
  rel_var_U <- rel_var_A + (s_sd / s)^2

  tibble(
    child_est        = N_hh * A,
    child_lb         = N_hh * A * exp(-z * sqrt(rel_var_A)),
    child_ub         = N_hh * A * exp( z * sqrt(rel_var_A)),
    unsafe_child_est = N_hh * A * s,
    unsafe_child_lb  = N_hh * A * s * exp(-z * sqrt(rel_var_U)),
    unsafe_child_ub  = N_hh * A * s * exp( z * sqrt(rel_var_U))
  ) %>%
    mutate(
      across(everything(), ~ .x / N_child, .names = "pct_{.col}"),
      totalhh  = N_hh,
      children = N_child
    )
}

sens_2025 <- bind_rows(
  recompute(
    N_hh    = census_estimates$totalhh[census_estimates$year == 2025],
    N_child = census_estimates$children[census_estimates$year == 2025],
    A       = A_2025$A,
    A_se    = A_2025$A_se
  ) %>% mutate(spec = "2024 ACS estimates"),

  recompute(
    N_hh    = proj_2025$totalhh_proj,
    N_child = proj_2025$children_proj,
    A       = A_2025$A,
    A_se    = A_2025$A_se
  ) %>% mutate(spec = "Projected 2025 totals")
) %>%
  select(spec, everything())
```

```{r}
fmt_m   <- function(x) paste0(sjmisc::big_mark(round(x / 1e6, 1)), " M")
fmt_pct <- function(x) paste0(round(x * 100, 1), "\\%")

sens_tbl <- tibble(
  characteristic = c(
    "Total US households",
    "US population under 18",
    "Estimated children exposed",
    "Children exposed (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms",
    "Children exposed to loaded and unlocked firearms (\\% of US children)"
  )
)

for (sp in unique(sens_2025$spec)) {
  d <- sens_2025 %>% filter(spec == sp)

  sens_tbl <- sens_tbl %>%
    mutate(
      !!paste0(make.names(sp), "_est") := c(
        fmt_m(d$totalhh),
        fmt_m(d$children),
        fmt_m(d$child_est),
        fmt_pct(d$pct_child_est),
        fmt_m(d$unsafe_child_est),
        fmt_pct(d$pct_unsafe_child_est)
      ),
      !!paste0(make.names(sp), "_ci") := c(
        "", "",
        paste0(fmt_m(d$child_lb), "–", fmt_m(d$child_ub)),
        paste0(fmt_pct(d$pct_child_lb), "–", fmt_pct(d$pct_child_ub)),
        paste0(fmt_m(d$unsafe_child_lb), "–", fmt_m(d$unsafe_child_ub)),
        paste0(fmt_pct(d$pct_unsafe_child_lb), "-", fmt_pct(d$pct_unsafe_child_ub))
      )
    )
}

sens_note <- paste0(
  "Table compares estimated numbers of US children exposed to firearms in the home ",
  "in 2025 under two assumptions about American Community Survey totals, United States, 2025. ",
  "The first column pair substitutes 2024 American Community Survey estimates, which ",
  "is the specification used in the main analysis because 2025 estimates were not yet ",
  "available. The second projects 2025 totals by fitting a log-linear trend to the ",
  "2013--2024 American Community Survey series separately for total households and for ",
  "the population younger than 18 years, implying annual growth of ",
  sprintf("%.2f\\%%", growth_hh * 100), " in households and ",
  sprintf("%.2f\\%%", growth_child * 100), " in the child population. ",
  "Survey-based inputs are identical across columns; only the American Community Survey ",
  "totals differ. Confidence intervals use the delta method on the log scale and are ",
  "therefore of identical relative width in both columns."
)

sens_latex <- sens_tbl %>%
  kable(
    format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
    caption = "Sensitivity of 2025 Estimates to American Community Survey Totals",
    label = "SensACS",
    col.names = c("Characteristic", rep(c("Est.", "95\\% CI"), 2)),
    align = "lcccc"
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  add_header_above(
    c(" " = 1, "2024 ACS estimates" = 2, "Projected 2025 totals" = 2)
  ) %>%
  footnote(number = esc_note(sens_note), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(sens_latex, file = here(path_ol, "tables", "sensitivity-acs.tex"))
```


## Plot over time


```{r}
prev_long <-
  proportion_estimates %>%
  transmute(
    year,
    `Firearm in the home`        = gunhousehold_yes,
    `Firearm and child under 18` = gunchild18un_yes,
    lb_gun = ci_l_gunhousehold_yes, ub_gun = ci_u_gunhousehold_yes,
    lb_bth = ci_l_gunchild18un_yes, ub_bth = ci_u_gunchild18un_yes
  ) %>%
  pivot_longer(c(`Firearm in the home`, `Firearm and child under 18`),
               names_to = "measure", values_to = "est") %>%
  mutate(
    lb = if_else(measure == "Firearm in the home", lb_gun, lb_bth),
    ub = if_else(measure == "Firearm in the home", ub_gun, ub_bth),
    across(c(est, lb, ub), ~ .x * 100),
    measure = factor(measure, levels = c("Firearm in the home",
                                         "Firearm and child under 18"))
  ) %>%
  filter(year != 2015)

hh_cols <- c("Firearm in the home"        = "#1f4e5f",
             "Firearm and child under 18" = "#e9a13b")

## Panel A
prev_long <- prev_long %>% mutate(year_f = factor(year))

p_prev <-
  ggplot(prev_long, aes(year_f, est, color = measure, fill = measure,
                        group = measure)) +
  geom_ribbon(aes(ymin = lb, ymax = ub), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  # geom_text(
  #   data = prev_long %>% filter(year == max(year)),
  #   aes(label = measure), hjust = 1, vjust = -1.5,
  #   size = 3.5, family = "serif", show.legend = FALSE
  # ) +
  scale_color_manual(values = hh_cols) +
  scale_fill_manual(values = hh_cols) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0.08, 0.12))) +
  labs(x = NULL, y = "Share of US households") +
  theme +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "bottom",
        legend.title = element_blank(),
        text = element_text(size = 14 ),
        legend.text = element_text(size = 12)
        )

## Panel B

exp_df <-
  exposure_delta %>%
  filter(!is.na(child_est)) %>%
  transmute(year,
            est = pct_child_est * 100,
            lb  = pct_child_lb  * 100,
            ub  = pct_child_ub  * 100,
            year_f = factor(year))

p_exp <-
  ggplot(exp_df, aes(year_f, est, group = 1)) +
  geom_ribbon(aes(ymin = lb, ymax = ub), fill = "#1f4e5f", alpha = 0.15) +
  geom_line(color = "#1f4e5f", linewidth = 0.9) +
  geom_point(color = "#1f4e5f", size = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0.08, 0.12))) +
  labs(x = NULL, y = "Share of US children living\nin a household with a firearm") +
  theme +
  theme(
    panel.grid.minor = element_blank(),
    text = element_text(size = 14 ))

main_results_plot <- p_prev / p_exp

ggsave(
  here(path_ol, "figures", "household-prevalence-over-time.png"),
  p_prev, width = 8, height = 4, dpi = 300
)

ggsave(
  here(path_ol, "figures", "children-exposed-over-time.png"),
  p_exp, width = 8, height = 4, dpi = 300
)
```


## Appendix table: calculation inputs

Seeing the all of the raw inputs allows us to know what is driving the increase.

```{R}
inputs_years <- sort(unique(A_hat$year))

inputs_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Total US households",
    "Household has a firearm and a child under 18",
    "Mean children per exposed household",
    "Mean exposed children per US household",
    "Estimated children exposed",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

for (yr in inputs_years) {

  yr_obs      <- obs_counts        %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_hh       <- census_estimates  %>% filter(year == yr)
  yr_prev     <- children_exposure %>% filter(year == yr)
  yr_m        <- avg_child         %>% filter(year == yr)
  yr_A        <- A_hat             %>% filter(year == yr)
  yr_exposure <- exposure_delta    %>% filter(year == yr)

  hh_est <- paste0(sjmisc::big_mark(round(yr_hh$totalhh / 1e6, 1)), " M")

  gc_est <- paste0(round(yr_prev$prev_pct * 100, 1), "%")
  gc_ci  <- paste0(round(yr_prev$prev_lb * 100, 1), "–",
                   round(yr_prev$prev_ub * 100, 1), "%")

  m_est <- sprintf("%.2f", yr_m$avg_child)
  m_ci  <- paste0(sprintf("%.2f", yr_m$avg_child_lb), "–",
                  sprintf("%.2f", yr_m$avg_child_ub))

  A_est <- sprintf("%.3f", yr_A$A)
  A_ci  <- paste0(sprintf("%.3f", yr_A$A - qnorm(0.975) * yr_A$A_se), "–",
                  sprintf("%.3f", yr_A$A + qnorm(0.975) * yr_A$A_se))

  exp_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exp_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "–",
                    sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")

  lu_est <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)), " M")
  lu_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)), "–",
                   sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)), " M")

  inputs_tbl_simple <- inputs_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Total US households" ~ hh_est,
        characteristic == "Household has a firearm and a child under 18" ~ gc_est,
        characteristic == "Mean children per exposed household" ~ m_est,
        characteristic == "Mean exposed children per US household" ~ A_est,
        characteristic == "Estimated children exposed" ~ exp_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Total US households" ~ "",
        characteristic == "Household has a firearm and a child under 18" ~ gc_ci,
        characteristic == "Mean children per exposed household" ~ m_ci,
        characteristic == "Mean exposed children per US household" ~ A_ci,
        characteristic == "Estimated children exposed" ~ exp_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_ci
      )
    )
}

inputs_tbl_simple <- inputs_tbl_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

col_names_inputs <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(inputs_years)))
align_inputs     <- paste0("l", paste0(rep("c", length(inputs_years) * 2), collapse = ""))

inputs_note <-
  paste0(
    "Table reports every quantity entering the estimated number of US children ",
    "exposed to firearms in the home, by survey wave, United States, 2013--2025. ",
    "Estimated children exposed = total US households (American Community Survey) ",
    "$\\times$ mean exposed children per US household (survey). The latter is ",
    "algebraically equal to the prevalence of households with a gun and a child ",
    "under 18 multiplied by the mean number of children in those households, both ",
    "of which are reported above so that changes across waves can be attributed to ",
    "each component; it is estimated directly so that its standard error accounts ",
    "for the covariance between them. Estimated children exposed to loaded and ",
    "unlocked firearms additionally multiplies by the 2024 National Firearm Survey ",
    "estimate that 21.1\\% (95\\% CI, 18.3\\%--24.3\\%) of gun-owning households with ",
    "children had at least one firearm stored loaded and unlocked (Miller et al.\\ 2026). ",
    "Total US household counts are American Community Survey 1-year estimates; 2024 ",
    "estimates were used for 2025. Confidence intervals for the derived counts use ",
    "the delta method."
  )

inputs_share_tbl <- inputs_tbl_simple %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Inputs to the Estimated Number of US Children Exposed to Firearms in the Home, 2013-2025",
    label     = "ExposureInputs",
    col.names = col_names_inputs,
    linesep   = "",
    align     = align_inputs
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  add_header_above(
    c(" " = 1, setNames(rep(2, length(inputs_years)), paste0(inputs_years)))
  ) %>%
  footnote(number = esc_note(inputs_note), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(inputs_share_tbl, file = here(path_ol, "tables/exposure-inputs.tex"))
```

## Test: does the association between having a firearm and a child in the household change over time

```{r}

ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)

## P(gun | child) by year, with CIs
gun_by_child <- svyby(
  ~gunhousehold, ~year + child18un, ds, svymean,
  vartype = c("se", "ci")
) %>% clean_names()

## Formal test: has the child-firearm association changed over time?
ds_ends <- subset(ds, year %in% c(2013, 2025))

m <- svyglm(
  I(gunhousehold == "Yes") ~ child18un * factor(year),
  design = ds_ends, family = quasibinomial()
)
summary(m)
```

## Ratio estimator

Our numerator comes from the survey. Our denominator comes from the ACS. And the survey and the ACS disagree about how many children exist in America so our estimates are biased.


```{r}
child %>%
  count(child18un, hh18un_zero = hh18un == 0, hh18un_na = is.na(hh18un))

svyby(~hh18un, ~year, ds, svymean, vartype = "se")
```



### Derive estimator


```{r}
z     <- qnorm(0.975)
s_hat <- 0.211
s_se  <- (0.243 - 0.183) / (2 * z)

ds_ratio <- update(
  ds,
  child_exposed = ifelse(gunhousehold == "Yes", hh18un, 0)
)

# P_t = sum(w * c * g) / sum(w * c)
# svyratio propagates the covariance between numerator and denominator
P_hat <- map_df(
  sort(unique(child$year)),
  function(yr) {
    r <- svyratio(~child_exposed, ~hh18un, subset(ds_ratio, year == yr))
    tibble(
      year = yr,
      P    = as.numeric(coef(r)),
      P_se = as.numeric(SE(r))
    )
  }
)

P_hat
```
### Combine with ACS

```{r}
exposure_ratio <-
  census_estimates %>%
  left_join(P_hat, by = "year") %>%
  mutate(
    rel_var_P = (P_se / P)^2,
    rel_var_U = rel_var_P + (s_se / s_hat)^2,

    ## share of US children exposed — pure survey quantity, no ACS
    pct_children_exposed    = P,
    pct_children_exposed_lb = P * exp(-z * sqrt(rel_var_P)),
    pct_children_exposed_ub = P * exp( z * sqrt(rel_var_P)),

    ## counts: ACS child population enters only here
    child_est = children * P,
    child_lb  = children * pct_children_exposed_lb,
    child_ub  = children * pct_children_exposed_ub,

    ## loaded & unlocked
    pct_children_unsafe    = P * s_hat,
    pct_children_unsafe_lb = P * s_hat * exp(-z * sqrt(rel_var_U)),
    pct_children_unsafe_ub = P * s_hat * exp( z * sqrt(rel_var_U)),

    unsafe_child_est = children * pct_children_unsafe,
    unsafe_child_lb  = children * pct_children_unsafe_lb,
    unsafe_child_ub  = children * pct_children_unsafe_ub
  ) %>%
  filter(!is.na(P))

```


### Main table

```{R}
years_r <- sort(unique(exposure_ratio$year))

tbl_ratio_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has a firearm",
    "Household has a child under 18",
    "Household has a firearm and a child under 18",
    "Children living with a firearm (\\% of US children)",
    "Estimated children exposed to firearms in household",
    "Children exposed to loaded and unlocked firearms (\\% of US children)",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

fmt_m   <- function(x) paste0(sjmisc::big_mark(round(x / 1e6, 1)), " M")
fmt_pct <- function(x) paste0(round(x * 100, 1), "%")

for (yr in years_r) {

  yr_data <- proportion_estimates %>% filter(year == yr)
  yr_obs  <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_exp  <- exposure_ratio %>% filter(year == yr)

  gun_est      <- fmt_pct(yr_data$gunhousehold_yes)
  gun_ci       <- paste0(fmt_pct(yr_data$ci_l_gunhousehold_yes), "–",
                         fmt_pct(yr_data$ci_u_gunhousehold_yes))
  child_est    <- fmt_pct(yr_data$child18un_yes)
  child_ci     <- paste0(fmt_pct(yr_data$ci_l_child18un_yes), "–",
                         fmt_pct(yr_data$ci_u_child18un_yes))
  gunchild_est <- fmt_pct(yr_data$gunchild18un_yes)
  gunchild_ci  <- paste0(fmt_pct(yr_data$ci_l_gunchild18un_yes), "–",
                         fmt_pct(yr_data$ci_u_gunchild18un_yes))

  exp_pct_est <- fmt_pct(yr_exp$pct_children_exposed)
  exp_pct_ci  <- paste0(fmt_pct(yr_exp$pct_children_exposed_lb), "–",
                        fmt_pct(yr_exp$pct_children_exposed_ub))
  exp_n_est   <- fmt_m(yr_exp$child_est)
  exp_n_ci    <- paste0(fmt_m(yr_exp$child_lb), "–", fmt_m(yr_exp$child_ub))

  uns_pct_est <- fmt_pct(yr_exp$pct_children_unsafe)
  uns_pct_ci  <- paste0(fmt_pct(yr_exp$pct_children_unsafe_lb), "–",
                        fmt_pct(yr_exp$pct_children_unsafe_ub))
  uns_n_est   <- fmt_m(yr_exp$unsafe_child_est)
  uns_n_ci    <- paste0(fmt_m(yr_exp$unsafe_child_lb), "–", fmt_m(yr_exp$unsafe_child_ub))

  tbl_ratio_simple <- tbl_ratio_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has a firearm" ~ gun_est,
        characteristic == "Household has a child under 18" ~ child_est,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_est,
        characteristic == "Children living with a firearm (\\% of US children)" ~ exp_pct_est,
        characteristic == "Estimated children exposed to firearms in household" ~ exp_n_est,
        characteristic == "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ uns_pct_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ uns_n_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has a firearm" ~ gun_ci,
        characteristic == "Household has a child under 18" ~ child_ci,
        characteristic == "Household has a firearm and a child under 18" ~ gunchild_ci,
        characteristic == "Children living with a firearm (\\% of US children)" ~ exp_pct_ci,
        characteristic == "Estimated children exposed to firearms in household" ~ exp_n_ci,
        characteristic == "Children exposed to loaded and unlocked firearms (\\% of US children)" ~ uns_pct_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ uns_n_ci
      )
    )
}

tbl_ratio_simple <- tbl_ratio_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

note_ratio <- paste0(
  "Data represent weighted estimates from the National Survey of Gun Policy, ",
  "United States, 2013--2025. Household-level rows report the weighted share of ",
  "US households with (1) a gun in the home, (2) a child under 18, and (3) both. ",
  "The share of US children living in a household with a firearm is estimated as a ",
  "ratio: the weighted number of children in respondent households with a firearm ",
  "divided by the weighted number of children in all respondent households. ",
  "This ratio form is used in place of scaling a per-household mean by American ",
  "Community Survey household totals, because the survey over-represents households ",
  "containing children relative to the American Community Survey and the degree of ",
  "over-representation varies across waves; the over-representation largely cancels ",
  "between the numerator and denominator of the ratio (Appendix Table). ",
  "Estimated counts of exposed children apply this share to the American Community ",
  "Survey population younger than 18 years. Children exposed to loaded and unlocked ",
  "firearms additionally multiply by the 2024 National Firearm Survey estimate that ",
  "21.1\\% (95\\% CI, 18.3\\%--24.3\\%) of gun-owning households with children had at ",
  "least one firearm stored loaded and unlocked (Miller et al.\\ 2026); this rate is ",
  "assumed constant across waves. Because 2025 American Community Survey estimates ",
  "were not yet available, 2024 estimates were used for 2025. The 2015 wave is excluded ",
  "because no household-level firearm ownership measure was collected. Confidence ",
  "intervals use the delta method, treating American Community Survey ",
  "totals as fixed and the storage estimate as independent of the survey estimates."
)

tbl_ratio <- tbl_ratio_simple %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    caption   = "Household Firearm Prevalence and Children's Firearm Exposure, 2013-2025",
    label     = "ShareYearRatio",
    col.names = c("Characteristic", rep(c("Est.", "95\\% CI"), length(years_r))),
    linesep   = "",
    align     = paste0("l", strrep("c", length(years_r) * 2))
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  row_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, setNames(rep(2, length(years_r)), paste0(years_r)))) %>%
  footnote(number = esc_note(note_ratio), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(tbl_ratio, file = here(path_ol, "tables/yearly-share-ratio.tex"))
```


### Calibration diagnostic

This is the table that justifies the switch. Put it in the appendix.

```{r}
## ---- Calibration of the survey household frame against the ACS -------------

## Survey-implied children per household, by wave
survey_cph <-
  svyby(~hh18un, ~year, ds, svymean, vartype = "se") %>%
  clean_names() %>%
  transmute(year, survey_cph = hh18un, survey_cph_se = se)

calib <-
  survey_cph %>%
  left_join(census_estimates, by = "year") %>%
  mutate(
    acs_cph          = children / totalhh,
    implied_children = totalhh * survey_cph,
    ratio            = survey_cph / acs_cph
  )

calib_tbl <- calib %>%
  transmute(
    Year        = as.character(year),
    survey_est  = sprintf("%.3f", survey_cph),
    survey_ci   = sprintf("%.3f--%.3f",
                          survey_cph - qnorm(0.975) * survey_cph_se,
                          survey_cph + qnorm(0.975) * survey_cph_se),
    acs_est     = sprintf("%.3f", acs_cph),
    ratio_est   = sprintf("%.2f", ratio),
    implied_pop = paste0(sjmisc::big_mark(round(implied_children / 1e6, 1)), " M"),
    acs_pop     = paste0(sjmisc::big_mark(round(children / 1e6, 1)), " M")
  )

calib_note <- paste0(
  "Table compares the number of children younger than 18 years per household implied ",
  "by the National Survey of Gun Policy with the corresponding American Community ",
  "Survey figure, by survey wave, United States, 2013--2025. The survey quantity is the ",
  "weighted mean of the household child count across all respondent households; the ",
  "American Community Survey quantity is the population younger than 18 years divided ",
  "by total households.",
  "The survey sample contains a higher proportion of households with children than the American Community Survey in all waves, with the largest divergence in 2021 and 2023. This reflects the composition of the recruited panel rather than the survey weights, which adjust child-containing households by a factor between 0.98 and 1.20 across waves. Because the source of the imbalance is the achieved sample rather than any single adjustment, estimates of the share of US children living in a household with a firearm are computed as a ratio of weighted child counts within the survey, so that the imbalance cancels between numerator and denominator.",
  "Confidence ",
  "intervals for the survey quantity are design-based; American Community Survey totals ",
  "are treated as fixed. The 2015 wave is excluded because no household-level firearm ",
  "ownership measure was collected."
)

calib_latex <- calib_tbl %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    linesep   = "",
    caption   = "Calibration of Survey Household Composition Against American Community Survey Totals, 2013-2025",
    label     = "CalibHH",
    col.names = c("Year", "Est.", "95\\% CI", "Est.", "Ratio", "Survey-implied", "ACS"),
    align     = "lcccccc"
  ) %>%
  kable_styling(latex_options = c("hold_position"), font_size = 9) %>%
  add_header_above(
    c(" " = 1,
      "Survey" = 2,
      "ACS" = 1,
      "Survey / ACS" = 1,
      "Population under 18" = 2)
  ) %>%
  add_header_above(
    c(" " = 1, "Children per household" = 4, " " = 2)
  ) %>%
  footnote(number = esc_note(calib_note), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(calib_latex, file = here(path_ol, "tables", "calibration-hh.tex"))
```


## Post stratification

The issue of differential recruitment of households with children in the survey
could be driving the bias we see. As a solution, post stratify using ACS targets for the number of children in the household.

```{r, warning = F, message = F}
acs_years <- c(2013, 2017, 2019, 2021, 2023, 2024)

acs_child_hh <-
  map_df(
    acs_years,
    ~ get_acs(
        geography = "us",
        table     = "B11005",
        year      = .x,
        survey    = "acs1",
        key       = KEY,
        cache_table = TRUE
      ) %>%
      filter(variable %in% c("B11005_001", "B11005_002")) %>%
      select(variable, estimate) %>%
      pivot_wider(names_from = variable, values_from = estimate) %>%
      transmute(
        year        = .x,
        total_hh    = B11005_001,
        hh_child    = B11005_002,
        hh_nochild  = B11005_001 - B11005_002,
        share_child = B11005_002 / B11005_001
      )
  ) %>%
  mutate(year = if_else(year == 2024, 2025, year))

acs_child_hh
```


```{r}
## Build the joint year × child18un target table
ps_targets <-
  acs_child_hh %>%
  pivot_longer(c(hh_child, hh_nochild),
               names_to = "child18un", values_to = "Freq") %>%
  mutate(
    child18un = factor(if_else(child18un == "hh_child", "Yes", "No"),
                       levels = levels(child$child18un)),
    year      = as.integer(year),
    Freq      = as.numeric(Freq)
  ) %>%
  select(year, child18un, Freq) %>%
  arrange(year, child18un)

## verify alignment before post-stratifying
setdiff(unique(child$year), unique(ps_targets$year))   # should be empty
setdiff(unique(ps_targets$year), unique(child$year))   # should be empty
ps_targets

ds_ps <- svydesign(~1, data = child, strata = ~year, weights = ~weight) %>%
  postStratify(strata = ~ year + child18un, population = ps_targets)
```

```{r}
ps_targets %>%
  pivot_wider(names_from = child18un, values_from = Freq) %>%
  mutate(share_child = round(100 * Yes / (Yes + No), 1))
```


### Compare weights


```{r}

## 1. How much did the weights actually move?
w_old <- weights(ds)
w_new <- weights(ds_ps)

tibble(year = child$year, child = child$child18un, w_old, w_new) %>%
  group_by(year, child) %>%
  summarise(factor = round(mean(w_new) / mean(w_old), 2), .groups = "drop") %>%
  pivot_wider(names_from = child, values_from = factor)

## 2. Extreme weights?
tibble(year = child$year, w_new) %>%
  group_by(year) %>%
  summarise(
    max_w  = round(max(w_new), 1),
    p99    = round(quantile(w_new, .99), 1),
    deff   = round(1 + (sd(w_new) / mean(w_new))^2, 2)   # weighting design effect
  )

```


### Compare to prior design

```{r}

ds_ps <- update(ds_ps, child_exposed = ifelse(gunhousehold == "Yes", hh18un, 0))

exposed <- svyby(~child_exposed, ~year, ds_ps, svytotal, vartype = "se") %>%
  clean_names() %>%
  transmute(year, N = child_exposed, N_se = se)

z <- qnorm(0.975); s_hat <- 0.211; s_se <- (0.243 - 0.183) / (2 * z)

exposure_ps <-
  exposed %>%
  left_join(census_estimates %>% select(year, children), by = "year") %>%
  mutate(
    rel_var  = (N_se / N)^2,
    N_lb     = N * exp(-z * sqrt(rel_var)),
    N_ub     = N * exp( z * sqrt(rel_var)),
    pct      = N / children,
    pct_lb   = N_lb / children,
    pct_ub   = N_ub / children,
    U        = N * s_hat,
    rel_varU = rel_var + (s_se / s_hat)^2,
    U_lb     = U * exp(-z * sqrt(rel_varU)),
    U_ub     = U * exp( z * sqrt(rel_varU)),
    pctU     = U / children,
    pctU_lb  = U_lb / children,
    pctU_ub  = U_ub / children
  )

exposure_ps %>%
  transmute(year,
            N   = round(N / 1e6, 1),
            pct = round(100 * pct, 1),
            ci  = paste0(round(100 * pct_lb, 1), "–", round(100 * pct_ub, 1)))

```


```{r}
compare_specs <-
  list(
    orig  = exposure_delta %>% transmute(year, pct = pct_child_est),
    ratio = exposure_ratio %>% transmute(year, pct = pct_children_exposed),
    ps    = exposure_ps    %>% transmute(year, pct = pct)
  ) %>%
  bind_rows(.id = "spec") %>%
  pivot_wider(names_from = spec, values_from = pct)

## adjustment factor from the post-stratification, for context
adj <-
  tibble(year = child$year, ch = child$child18un,
         w_old = weights(ds), w_new = weights(ds_ps)) %>%
  group_by(year) %>%
  summarise(
    adj_child = (sum(w_new[ch == "Yes"]) / sum(w_old[ch == "Yes"])) /
                (sum(w_new) / sum(w_old)),
    .groups = "drop"
  )

compare_tbl <-
  compare_specs %>%
  left_join(adj, by = "year") %>%
  filter(year != 2015) |> 
  transmute(
    Year  = as.character(year),
    orig  = sprintf("%.1f\\%%", 100 * orig),
    ratio = sprintf("%.1f\\%%", 100 * ratio),
    ps    = sprintf("%.1f\\%%", 100 * ps),
    adj   = sprintf("%.2f", adj_child)
  ) 

compare_note <- paste0(
  "Table compares three estimators of the share of US children younger than 18 years ",
  "living in a household with a firearm, by survey wave, United States, 2013--2025. ",
  "The household-scaled estimator multiplies the weighted mean number of ",
  "firearm-exposed children per respondent household by the American Community Survey ",
  "estimate of total US households, and divides by the American Community Survey ",
  "population younger than 18 years. The ratio estimator divides the weighted number ",
  "of children in respondent households with a firearm by the weighted number of ",
  "children in all respondent households. The post-stratified estimator applies the ",
  "household-scaled calculation after reweighting each wave to American Community ",
  "Survey counts of households with and without people younger than 18 years. ",
  "The final column reports the post-stratification adjustment applied to households ",
  "with children in each wave, relative to the wave as a whole; values below one ",
  "indicate that such households were over-represented in the achieved sample. ",
  "The 2015 wave is excluded because no household-level firearm ownership measure ",
  "was collected."
)

compare_latex <-
  compare_tbl %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    linesep   = "",
    caption   = "Comparison of Estimators of Children's Firearm Exposure, 2013-2025",
    label     = "SpecCompare",
    col.names = c("Year", "Household-scaled", "Ratio", "Post-stratified",
                  "Adjustment factor"),
    align     = "lcccc"
  ) %>%
  kable_styling(latex_options = "hold_position", font_size = 9) %>%
  add_header_above(
    c(" " = 1, "Share of US children living with a firearm" = 3, " " = 1)
  ) %>%
  footnote(number = esc_note(compare_note), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(compare_latex, file = here(path_ol, "tables", "estimator-comparison.tex"))
```


### Main table

```{r}
## Design with the derived variable
ds_ps <- update(ds_ps, child_exposed = ifelse(gunhousehold == "Yes", hh18un, 0))

z     <- qnorm(0.975)
s_hat <- 0.211
s_se  <- (0.243 - 0.183) / (2 * z)


## Household-level proportions
prop_ps <-
  svyby(~gunhousehold + child18un + gunchild18un, ~year, ds_ps,
        svymean, vartype = c("se", "ci"), level = 0.95) %>%
  clean_names()

## Exposed children as a population total — no ACS household count needed
exposed_ps <-
  svyby(~child_exposed, ~year, ds_ps, svytotal, vartype = "se") %>%
  clean_names() %>%
  transmute(year, N = child_exposed, N_se = se)

## Unweighted observation counts
obs_counts <- child %>% count(year, name = "obs")

## Combine with ACS child population for the share
exposure_ps <-
  exposed_ps %>%
  left_join(census_estimates %>% select(year, children), by = "year") %>%
  mutate(
    rel_var  = (N_se / N)^2,
    N_lb     = N * exp(-z * sqrt(rel_var)),
    N_ub     = N * exp( z * sqrt(rel_var)),
    pct      = N    / children,
    pct_lb   = N_lb / children,
    pct_ub   = N_ub / children,

    U        = N * s_hat,
    rel_varU = rel_var + (s_se / s_hat)^2,
    U_lb     = U * exp(-z * sqrt(rel_varU)),
    U_ub     = U * exp( z * sqrt(rel_varU)),
    pctU     = U    / children,
    pctU_lb  = U_lb / children,
    pctU_ub  = U_ub / children
  )

fmt_m   <- function(x) paste0(sjmisc::big_mark(round(x / 1e6, 1)), " M")
fmt_pct <- function(x) paste0(round(x * 100, 1), "%")

years_ps <- sort(unique(exposure_ps$year))

main_tbl <- tibble(
  characteristic = c(
    "Observations",
    "Household has a firearm",
    "Household has a child under 18",
    "Household has a firearm and a child under 18",
    "Children living with a firearm (\\% of US children)",
    "Estimated children living with a firearm",
    "Children living with a loaded and unlocked firearm (\\% of US children)",
    "Estimated children living with a loaded and unlocked firearm"
  )
)

for (yr in years_ps) {

  p <- prop_ps    %>% filter(year == yr)
  e <- exposure_ps %>% filter(year == yr)
  o <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")

  vals <- c(
    o,
    fmt_pct(p$gunhousehold_yes),
    fmt_pct(p$child18un_yes),
    fmt_pct(p$gunchild18un_yes),
    fmt_pct(e$pct),
    fmt_m(e$N),
    fmt_pct(e$pctU),
    fmt_m(e$U)
  )

  cis <- c(
    "",
    paste0(fmt_pct(p$ci_l_gunhousehold_yes), "–", fmt_pct(p$ci_u_gunhousehold_yes)),
    paste0(fmt_pct(p$ci_l_child18un_yes),    "–", fmt_pct(p$ci_u_child18un_yes)),
    paste0(fmt_pct(p$ci_l_gunchild18un_yes), "–", fmt_pct(p$ci_u_gunchild18un_yes)),
    paste0(fmt_pct(e$pct_lb),  "–", fmt_pct(e$pct_ub)),
    paste0(fmt_m(e$N_lb),      "–", fmt_m(e$N_ub)),
    paste0(fmt_pct(e$pctU_lb), "–", fmt_pct(e$pctU_ub)),
    paste0(fmt_m(e$U_lb),      "–", fmt_m(e$U_ub))
  )

  main_tbl <- main_tbl %>%
    mutate(
      !!paste0("y", yr, "_est") := vals,
      !!paste0("y", yr, "_ci")  := cis
    )
}

main_tbl <- main_tbl %>%
  mutate(across(starts_with("y"), ~ str_replace_all(.x, "%", "\\\\%")))


note_ps <- paste0(
  "Data represent weighted estimates from the National Survey of Gun Policy, United ",
  "States, 2013--2025. Survey weights were post-stratified within each wave to ",
  "American Community Survey counts of households with and without people younger ",
  "than 18 years, because the achieved sample contained a higher ",
  "proportion of households with children than the US household population in every ",
  "wave. The share of households with a child ",
  "under 18 therefore reproduces the American Community Survey benchmark by ",
  "construction. Estimated numbers of children are weighted population totals from ",
  "the post-stratified design; shares divide these by the American Community Survey ",
  "population younger than 18 years, treated as fixed. Estimates for children living ",
  "with a loaded and unlocked firearm additionally multiply by the 2024 National ",
  "Firearm Survey estimate that 21.1\\% (95\\% CI, 18.3\\%--24.3\\%) of gun-owning ",
  "households with children had at least one firearm stored loaded and unlocked ",
  "(Miller et al.\\ 2026); this proportion is assumed constant across waves and its ",
  "interval reflects sampling error in that estimate only. Because 2025 American ",
  "Community Survey estimates were not yet available, 2024 estimates were used for ",
  "2025. The 2015 wave is excluded because no household-level firearm ownership ",
  "measure was collected. Confidence intervals use the delta method on the log scale ",
  "and incorporate the post-stratification adjustment in the design-based variance. ",
  "Observation counts are unweighted."
)

main_latex <- main_tbl %>%
  kable(
    format    = "latex",
    booktabs  = TRUE,
    escape    = FALSE,
    linesep   = "",
    caption   = "Household Firearm Prevalence and Children's Firearm Exposure, 2013-2025",
    label     = "MainPS",
    col.names = c("Characteristic", rep(c("Est.", "95\\% CI"), length(years_ps))),
    align     = paste0("l", strrep("c", 2 * length(years_ps)))
  ) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  row_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, setNames(rep(2, length(years_ps)), as.character(years_ps)))) %>%
  footnote(number = esc_note(note_ps), threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(main_latex, file = here(path_ol, "tables", "main-poststrat.tex"))

```


## Figure: estimates and 95% CIs by estimator

Two panels: (A) the share of US households with both a firearm and a child under
18, and (B) the share of US children exposed to a loaded and unlocked firearm.
Within each panel the three estimators are distinguished by colour and shape.

Note that the main and ratio estimators use the same (unadjusted) survey design,
so their household-level estimates in Panel A coincide; only the
post-stratified estimator differs there. Points are dodged so overlapping
estimators remain visible.

```{r}
estimator_levels <- c("Main", "Ratio", "Post-stratified")

## Panel A: household has a firearm and a child under 18 -----------------------
panel_a <-
  bind_rows(
    proportion_estimates %>%
      transmute(
        year,
        estimator = "Main",
        est = gunchild18un_yes,
        lb  = ci_l_gunchild18un_yes,
        ub  = ci_u_gunchild18un_yes
      ),
    proportion_estimates %>%
      transmute(
        year,
        estimator = "Ratio",
        est = gunchild18un_yes,
        lb  = ci_l_gunchild18un_yes,
        ub  = ci_u_gunchild18un_yes
      ),
    prop_ps %>%
      transmute(
        year,
        estimator = "Post-stratified",
        est = gunchild18un_yes,
        lb  = ci_l_gunchild18un_yes,
        ub  = ci_u_gunchild18un_yes
      )
  ) %>%
  mutate(panel = "Household has a firearm and a child under 18")

## Panel B: children exposed to loaded and unlocked firearms -------------------
panel_b <-
  bind_rows(
    exposure_delta %>%
      transmute(
        year,
        estimator = "Main",
        est = pct_unsafe_child_est,
        lb  = pct_unsafe_child_lb,
        ub  = pct_unsafe_child_ub
      ),
    exposure_ratio %>%
      transmute(
        year,
        estimator = "Ratio",
        est = pct_children_unsafe,
        lb  = pct_children_unsafe_lb,
        ub  = pct_children_unsafe_ub
      ),
    exposure_ps %>%
      transmute(
        year,
        estimator = "Post-stratified",
        est = pctU,
        lb  = pctU_lb,
        ub  = pctU_ub
      )
  ) %>%
  mutate(panel = "Children exposed to loaded and unlocked firearms")

estimator_df <-
  bind_rows(panel_a, panel_b) %>%
  filter(year != 2015, !is.na(est)) %>%
  mutate(
    across(c(est, lb, ub), ~ .x * 100),
    estimator = factor(estimator, levels = estimator_levels),
    year_f    = factor(year),
    panel     = factor(
      panel,
      levels = c("Household has a firearm and a child under 18",
                 "Children exposed to loaded and unlocked firearms")
    )
  )

estimator_cols <- c(
  "Main"            = "#1f4e5f",
  "Ratio"           = "#e9a13b",
  "Post-stratified" = "#3fa7d6"
)

estimator_shapes <- c(
  "Main"            = 16,   # circle
  "Ratio"           = 17,   # triangle
  "Post-stratified" = 15    # square
)
```


```{r}
## Every layer takes the same dodge so the connecting lines pass through the
## dodged points rather than through the year midpoints. Only the point layer
## contributes to the legend, so the keys show the shape alone.
dodge <- position_dodge(width = 0.5)

estimator_plot <-
  ggplot(
    estimator_df,
    aes(year_f, est,
        color = estimator, shape = estimator, group = estimator)
  ) +
  geom_line(
    position = dodge,
    linewidth = 0.6,
    show.legend = FALSE
  ) +
  geom_errorbar(
    aes(ymin = lb, ymax = ub),
    position = dodge,
    width = 0.25,
    linewidth = 0.6,
    show.legend = FALSE
  ) +
  geom_point(
    position = dodge,
    size = 2.2
  ) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  scale_color_manual(values = estimator_cols) +
  scale_shape_manual(values = estimator_shapes) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0.10, 0.12))
  ) +
  labs(x = NULL, y = "Weighted percentage (95% CI)") +
  theme +
  theme(
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_blank(),
    text             = element_text(size = 14),
    legend.text      = element_text(size = 12)
  )

estimator_plot

ggsave(
  here(path_ol, "figures", "estimator-comparison-over-time.png"),
  estimator_plot, width = 8, height = 7, dpi = 300
)
```