# Comparison table

This chapter compares overall households; households with a firearm (subheader is with or without children in the households) on basic characteristics This pools the characteristics over time in Panel A. Panel B is 2013 and then Panel C is 2025.

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "ggthemes",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    )
```


## Household characteristics by children

```{r}
vars <- c(
  gender       = "Gender",
  age4         = "Age",
  racethnicity = "Race and ethnicity",
  altparty =  "Political affiliation",
  educ5        = "Highest level of education",
  employ       = "Current employment status",
  income4      = "Household income",
  marital      = "Marital status",
  metro        = "Metropolitan area",
  region4      = "Region"
)

child <-
  child |> 
  mutate(
    across(all_of(names(vars)), ~ droplevels(.x))
  )
ds <- svydesign(~1, data = child, strata = ~year, weights = ~weight)
```


```{r}
col_defs <- list(
  "All households"         = quote(!is.na(year)),
  "All firearm households" = quote(gunhousehold == "Yes"),
  "With children"          = quote(gunhousehold == "Yes" & child18un == "Yes"),
  "Without children"       = quote(gunhousehold == "Yes" & child18un == "No")
)


subset_design <- function(design, expr) {
  keep <- eval(expr, envir = design$variables, enclos = parent.frame())
  keep[is.na(keep)] <- FALSE
  design[which(keep), ]
}

wtd_dist <- function(design, vars) {
  map_dfr(names(vars), function(v) {
    est <- svymean(reformulate(v), design, na.rm = TRUE)
    ci  <- confint(est, level = 0.95)
    tibble(
      variable = v,
      level    = str_remove(names(coef(est)), paste0("^", v)),
      pct      = as.numeric(coef(est)) * 100,
      lb       = pmax(ci[, 1] * 100, 0),
      ub       = pmin(ci[, 2] * 100, 100)
    )
  })
}

## Tables are written with escape = FALSE so that the manually escaped percent
## signs survive; that means characteristic labels have to be escaped by hand.
## Household income levels (e.g. "$50,000-74,999") otherwise open LaTeX math mode.
esc_tex <- function(x) {
  x %>%
    str_replace_all(fixed("\\"), "\\textbackslash{}") %>%
    str_replace_all("([&%$#_{}])", "\\\\\\1")
}

fmt_pct <- function(x) ifelse(is.na(x), "--", paste0(sprintf("%.1f", x), "\\%"))
fmt_ci   <- function(lb, ub) {
  ifelse(is.na(lb), "--",
         paste0(sprintf("%.1f", lb), "--", sprintf("%.1f", ub), "\\%"))
}

build_panel <- function(design) {

  pieces <- imap(col_defs, function(expr, nm) {
    d <- subset_design(design, expr)

    dist <- wtd_dist(d, vars) %>%
      transmute(variable, level,
                est = fmt_pct(pct),
                ci  = fmt_ci(lb, ub))

    obs <- tibble(
      variable = "Observations",
      level    = "Observations",
      est      = format(nrow(d$variables), big.mark = ","),
      ci       = ""
    )

    bind_rows(obs, dist) %>%
      rename(!!paste0(nm, "_est") := est,
             !!paste0(nm, "_ci")  := ci)
  })

  reduce(pieces, full_join, by = c("variable", "level"))
}


make_table <- function(
    design, caption, label, file, subtitle_note = "",
    outer_header = c(" " = 3, "Households with a firearm" = 6),
    note_intro = 
      paste0("Table reports weighted demographic characteristics of US households, ",
             "overall and among households with a firearm in the home, by presence of ",
             "a child under 18, United States, 2013--2025. "
                       )) {

  body <- build_panel(design) %>%
    mutate(row   = row_number(),
           label = if_else(variable == "Observations", "Observations",
                           esc_tex(level)))

  var_idx <- body %>%
    filter(variable != "Observations") %>%
    group_by(variable) %>%
    summarise(start = min(row), end = max(row), .groups = "drop") %>%
    mutate(label = vars[variable]) %>%
    arrange(start)

  tbl_print <- body %>%
    select(label, ends_with("_est"), ends_with("_ci")) %>%
    select(label,
           unlist(map(names(col_defs), ~ paste0(.x, c("_est", "_ci")))))

  note <- paste0(
    note_intro,
    subtitle_note,
    "Column percentages are weighted using year-specific general population ",
    "survey weights and may not sum to 100 because of rounding and item ",
    "nonresponse; observation counts are unweighted. Confidence intervals are ",
    "95 percent Wald intervals from the survey design and are truncated at 0 and 100. ",
    "The 2015 wave is excluded because no household-level firearm ownership ",
    "measure was collected."
  )

  out <- tbl_print %>%
    kable(
      format    = "latex",
      booktabs  = TRUE,
      escape    = FALSE,
      caption   = caption,
      label     = label,
      col.names = c("Characteristic", rep(c("Est.", "95\\% CI"), length(col_defs))),
      linesep   = "",
      align     = paste0("l", strrep("c", 2 * length(col_defs)))
    ) %>%
    kable_styling(
      latex_options = c("hold_position", "scale_down"),
      font_size = 8
    ) %>%
    add_header_above(
      c(" " = 1, setNames(rep(2, length(col_defs)), names(col_defs)))
    ) %>%
    add_header_above(outer_header) %>%
    row_spec(1, bold = TRUE)

  for (i in seq_len(nrow(var_idx))) {
    out <- pack_rows(out, var_idx$label[i], var_idx$start[i], var_idx$end[i],
                     bold = TRUE, italic = FALSE, indent = FALSE,
                     latex_gap_space = "0.4em")
  }

  out <- out %>%
    footnote(
      number = note,
      threeparttable = TRUE,
      escape = FALSE,
      fixed_small_size = TRUE
    )

  save_kable(out, file = here(path_ol, "tables", file))

  invisible(out)
}
```

### Run for each sample

```{r}
make_table(
  design  = ds,
  caption = "Characteristics of US Adults Overall and Among Households with a Firearm, by Presence of Children, Pooled 2013--2025",
  label   = "HHCharPooled",
  file    = "household-characteristics-pooled.tex",
  subtitle_note = "Estimates pool all survey waves from 2013 to 2025. "
)

make_table(
  design  = subset_design(ds, quote(year == 2013)),
  caption = "Characteristics of US Adults Overall and Among Households with a Firearm, by Presence of Children, 2013",
  label   = "HHChar2013",
  file    = "household-characteristics-2013.tex",
  subtitle_note = "Estimates are from the 2013 survey wave. "
)

make_table(
  design  = subset_design(ds, quote(year == 2025)),
  caption = "Characteristics of US Adults Overall and Among Households with a Firearm, by Presence of Children, 2025",
  label   = "HHChar2025",
  file    = "household-characteristics-2025.tex",
  subtitle_note = "Estimates are from the 2025 survey wave. "
)
```


## Household characteristics by firearm status


```{r}
col_defs <- list(
  quote(!is.na(year)),
  quote(child18un == "Yes"),
  quote(child18un == "Yes" & gunhousehold == "Yes"),
  quote(child18un == "Yes" & gunhousehold == "No")
)
names(col_defs) <- c(
  "All households",
  "All households with children",
  "With a firearm",
  "Without a firearm"
)

child_note_intro <- paste0(
  "Table reports weighted demographic characteristics of US households, ",
  "overall and among households with a child under 18, by presence of a ",
  "firearm in the home, United States, 2013--2025. "
)

child_outer <- c(" " = 3, "Households with children" = 6)

make_table(
  design  = ds,
  caption = "Characteristics of US Adults Overall and Among Households with Children, by Presence of a Firearm, Pooled 2013--2025",
  label   = "HHChildCharPooled",
  file    = "household-child-characteristics-pooled.tex",
  subtitle_note = "Estimates pool all survey waves from 2013 to 2025. ",
  outer_header  = child_outer,
  note_intro    = child_note_intro
)

make_table(
  design  = subset_design(ds, quote(year == 2013)),
  caption = "Characteristics of US Adults Overall and Among Households with Children, by Presence of a Firearm, 2013",
  label   = "HHChildChar2013",
  file    = "household-child-characteristics-2013.tex",
  subtitle_note = "Estimates are from the 2013 survey wave. ",
  outer_header  = child_outer,
  note_intro    = child_note_intro
)

make_table(
  design  = subset_design(ds, quote(year == 2025)),
  caption = "Characteristics of US Adults Overall and Among Households with Children, by Presence of a Firearm, 2025",
  label   = "HHChildChar2025",
  file    = "household-child-characteristics-2025.tex",
  subtitle_note = "Estimates are from the 2025 survey wave. ",
  outer_header  = child_outer,
  note_intro    = child_note_intro
)
```


## Summary table of the US population

```{r}
demovars <-
  c(
    "gunhousehold",
    "child18un",
    "gunchild18un",
    "gender",
    "age4",
    "racethnicity",
    "altparty",
    "educ",
    "employ",
    "income4",
    "marital",
    "metro",
    "region4"
  )
```

## Yearly Summary

```{r}
ds <- 
  svydesign(
    ~ 1, 
    data = child, 
    strata = ~ year,
    weights = ~ weight
    ) 
```

```{r}
note <-
  paste0(
    "Values are weighted percentages describing the demographic composition of US adults from the National Survey of Gun Policy, a nationally representative survey conducted biennially across the United States from 2013 to 2025. ",
    "``Some college or associate's degree'' includes vocational or technical school. ",
    "``Other'' gender, Asian non-Hispanic, and ``Living with partner'' ",
    "were not collected in all waves (introduced in 2025 and 2019, and discontinued after 2021, respectively)."
  )
```

```{r}
year_tbl <-
  ds %>%
  tbl_svysummary(
    by = year,
    include = demovars,
    missing = "no",
    statistic = all_categorical() ~ "{p}%"
  ) %>%
  # add_ci() %>%
  modify_header(
    label ~ "Characteristic",
    all_stat_cols() ~ "{level}"
  ) %>%
  modify_table_body(
    ~ {

      tbl <- .x

      # compute unweighted yearly totals
      counts <- format(table(child$year), big.mark = ",")

      # create row with obs
      n_row <- tibble(
        variable = NA,
        var_label = NA,
        row_type = "label",
        label = "Observations",
        !!!setNames(
          as.list(counts),
          paste0("stat_", seq_along(counts))
          )
        ) %>%
        mutate(across(starts_with("stat"), trimws)
          )

      bind_rows(n_row, tbl) %>%
        filter(
          ! grepl("Ques", label),
          ! (grepl("Unknown", label)),
          label != "Non-Metro"
          ) %>%
        mutate(
           # escape LaTeX specials ($ in the income levels) before indenting,
           # because the table is written with escape = FALSE
           label = ifelse(
            row_type == "level" & label != "Observations",
            paste0("  ", esc_tex(label)),
            esc_tex(label)),
           across(
             c(stat_5, stat_6),
             ~ ifelse(
               variable == "marital" & grepl("with", label),
               "-", .x
               )
             ),
          across(
            # Asian american not a category until > 2017
            c(stat_1, stat_2),
            ~ ifelse(
              variable == "racethnicity" & grepl("Asian", label),
              "-", .x
              )
            ),
          # other gender category only 2025
          across(
            starts_with("stat"),
            ~ ifelse(
              variable == "gender" & .x == "0%",
              "-", .x
              )
          )
           )
    }
  ) %>%
  modify_footnote(all_stat_cols() ~ NA_character_) %>%
  as_kable_extra(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    label = "GenSummaryYear",
    linesep = "",
    caption =
      "Demographic Composition of US Adults by Survey Year, 2013–2025",
    align = "lccccccc"
  ) %>%
  kable_styling(
    latex_options =
      c("hold_position", "scale_down"),
   font_size = 9
  ) %>%
  column_spec(1, width = "4.7cm") %>%
  column_spec(
    2:8,
    width = "1.3cm"
    ) %>%
  row_spec(1, bold = TRUE) %>%
  footnote(
   number = note,
   threeparttable = TRUE,
   escape = FALSE,
   fixed_small_size = TRUE
  )

save_kable(
  year_tbl,
  file = here(path_ol, "tables/gen-pop-summary-year.tex")
)
```
