# Summary of Number of Exposed Children

This chapter creates a descriptive table that shows the demographics of households with children exposed to guns in the home. 

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    )
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)
```


## Weighted share of exposed households with CIs

```{r}
children_exposure <- 
  svyby(
    ~gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>% 
  transmute(
    year,
    prev_pct = gunchild18un_yes,
    prev_se = se_gunchild18un_yes,
    prev_lb = ci_l_gunchild18un_yes,
    prev_ub = ci_u_gunchild18un_yes
  )
```

## Average number of children in exposed households

```{r}
avg_child <-
  svyby(
    ~hh18un,
    ~year,
    subset(ds, gunchild18un == "Yes"),
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>%
  mutate(across(-year, ~ .x)) %>%
  transmute(
    year,
    avg_child = hh18un,
    avg_child_se = se,
    avg_child_lb = ci_l,
    avg_child_ub = ci_u
  ) 
```

## Get population estimates from the ACS

I applied for an API key using this [website](https://api.census.gov/data/key_signup.html) and my jdada3@jh.edu email.

2025 estimates are not available, so we use the 2024 estimates for that year.

- [B11001_001](https://data.census.gov/table/ACSDT1Y2022.B11001)


```{r, message = F, warning = F}
census_estimates <- 
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = c(total_hh = "B11001_001"),  # Total households
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
    mutate(year = .x) %>%
    select(year, estimate)
  ) %>%
  rename(totalhh = estimate) %>%
  select(year, totalhh) %>% 
  mutate(year = ifelse(year == 2024, 2025, year))
```


## Combine Gun Survey and ACS estimates to compute the 95% CI of the number of children expose

We use the conservative multiplication rule for combining uncertainty from 3 sources
by multiplying the upper/lower bound estimates from each of the parameters included in the final calculation


```{r}
 # 2024 National Firearm Survey estimate of unsafe firearm storage
unsafe_prev <- 0.211
unsafe_lb   <- 0.183
unsafe_ub   <- 0.243
```

```{r}
exposure_estimates <-
  census_estimates %>%
  left_join(children_exposure, by = "year") %>%
  left_join(avg_child, by = "year") %>%
  mutate(

    # households with guns and children
    hh_exposed = totalhh * prev_pct,

    # children exposed to any gun
    child_est = hh_exposed * avg_child,

    # children exposed to loaded & unlocked firearms
    unsafe_child_est = child_est * unsafe_prev,

    # confidence intervals
    child_lb = totalhh * prev_lb * avg_child_lb,
    child_ub = totalhh * prev_ub * avg_child_ub,

    unsafe_child_lb = totalhh * prev_lb * avg_child_lb * unsafe_lb,
    unsafe_child_ub = totalhh * prev_ub * avg_child_ub * unsafe_ub
  ) %>%
  select(
    year,
    child_est,
    child_lb,
    child_ub,
    unsafe_child_est,
    unsafe_child_lb,
    unsafe_child_ub
  )
```

```{r}
exposure_estimates %>% 
  mutate(
    across(-year, ~ sjmisc::big_mark(.x))
  )
```

## Child variables

```{R}
childvars <-
  c(
    "gunhousehold",
    "child18un",
    "gunchild18un"
  )
```

## Percentage of US households with guns and children in the home

```{r}
note <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns), while 2013, 2017, 2019, 2021, 2023, ",
    "and 2025 measured any guns in the home regardless of ownership. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying ",
    "the total number of US households (from the Census Bureau American Community Survey), ",
    "the prevalence of households with guns and children (survey estimate), and ",
    "the average number of children per exposed household (survey estimate). ",
    "Estimated children exposed to loaded and unlocked firearms additionally multiply ",
    "these estimates by the 2024 National Firearm Survey estimate that 21.1% ",
    "(95% CI, 18.3%–24.3%) of gun-owning households with children had at least one ",
    "firearm stored loaded and unlocked, reported by Miller et al. (2026). ",
    "Because 2025 American Community Survey estimates were not yet available, ",
    "2024 ACS household estimates were used for the 2025 calculations. ",
    "The 95 percent confidence intervals for estimated numbers of children use the conservative ",
    "multiplication rule."
  )
```

## Summary table

```{r}
# Calculate all proportion estimates with CIs
proportion_estimates <- 
  svyby(
    ~gunhousehold + child18un + gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names()

# Add observations
obs_counts <- tibble(
  year = as.numeric(names(table(child$year))),
  obs = as.numeric(table(child$year))
)

# Get unique years
years <- sort(unique(proportion_estimates$year))

# Create table structure
share_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Household has gun",
    "Household has child under 18",
    "Household has gun and child under 18",
    "Estimated children exposed",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

# Add columns for each year
for (yr in years) {
  yr_data <- proportion_estimates %>% filter(year == yr)
  yr_obs <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  
  # Exposure estimates for this year
  yr_exposure <- exposure_estimates %>% filter(year == yr)
  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "–", 
                        sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")
  
  # Gun
  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–", 
                   round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")
  
  # Child
  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "–", 
                     round(yr_data$ci_u_child18un_yes * 100, 1), "%")
  
  # Gun and child
  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "–", 
                        round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")
  
  # unsafe estimate
  unsafe_est <-
  paste0(
    sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)),
    " M"
  )

unsafe_ci <-
  paste0(
    sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)),
    "–",
    sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)),
    " M"
  )
  
  # Add columns (Est, CI pairs for this year)
  share_tbl_simple <- share_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has gun" ~ gun_est,
        characteristic == "Household has child under 18" ~ child_est,
        characteristic == "Household has gun and child under 18" ~ gunchild_est,
        characteristic == "Estimated children exposed" ~ exposure_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has gun" ~ gun_ci,
        characteristic == "Household has child under 18" ~ child_ci,
        characteristic == "Household has gun and child under 18" ~ gunchild_ci,
        characteristic == "Estimated children exposed" ~ exposure_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ unsafe_ci
        
      )
    )
}

# Escape % for LaTeX
share_tbl_simple <- share_tbl_simple %>%
  mutate(
    across(
      starts_with("y"),
      ~ str_replace_all(., "%", "\\\\%")
    )
  )

# Create column names for header (without year labels, those go in add_header_above)
col_names <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years)))
align_string <- paste0("l", paste0(rep("c", length(years) * 2), collapse = ""))

share_tbl <- share_tbl_simple %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "Share of US Households with Guns and Children Under 18, 2013-2025",
    label = "ShareYear",
    col.names = col_names,
    linesep = "",
    align = align_string
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(
    row = 1,
    bold = T
  ) %>% 
  add_header_above(
    c(" " = 1, setNames(rep(2, length(years)), paste0(years)))
  ) %>%
  footnote(
    number = note,
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  share_tbl,
  file = here(path_ol, "tables/yearly-share.tex")
)
```

## Appendix table: calculation inputs

Seeing the all of the raw inputs allows us to know what is driving the increase


```{R}
inputs_years <- sort(unique(children_exposure$year))

inputs_tbl_simple <- tibble(
  characteristic = c(
    "Observations",
    "Total US households",
    "Household has gun and child under 18",
    "Mean children per gun-owning household with children",
    "Estimated children exposed",
    "Estimated children exposed to loaded and unlocked firearms"
  )
)

for (yr in inputs_years) {

  yr_obs      <- obs_counts         %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  yr_hh       <- census_estimates   %>% filter(year == yr)
  yr_prev     <- children_exposure  %>% filter(year == yr)
  yr_m        <- avg_child          %>% filter(year == yr)
  yr_exposure <- exposure_estimates %>% filter(year == yr)

  # Total US households (millions) — ACS point estimate, no CI
  hh_est <- paste0(sjmisc::big_mark(round(yr_hh$totalhh / 1e6, 1)), " M")

  # P(gun & child)
  gc_est <- paste0(round(yr_prev$prev_pct * 100, 1), "%")
  gc_ci  <- paste0(round(yr_prev$prev_lb * 100, 1), "–",
                   round(yr_prev$prev_ub * 100, 1), "%")

  # Mean children per exposed household
  m_est <- sprintf("%.2f", yr_m$avg_child)
  m_ci  <- paste0(sprintf("%.2f", yr_m$avg_child_lb), "–",
                  sprintf("%.2f", yr_m$avg_child_ub))

  # Children exposed
  exp_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exp_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "–",
                    sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")

  # Children exposed to loaded & unlocked
  lu_est <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_est / 1e6, 1)), " M")
  lu_ci  <- paste0(sjmisc::big_mark(round(yr_exposure$unsafe_child_lb / 1e6, 1)), "–",
                   sjmisc::big_mark(round(yr_exposure$unsafe_child_ub / 1e6, 1)), " M")

  inputs_tbl_simple <- inputs_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Total US households" ~ hh_est,
        characteristic == "Household has gun and child under 18" ~ gc_est,
        characteristic == "Mean children per exposed household" ~ m_est,
        characteristic == "Estimated children exposed" ~ exp_est,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Total US households" ~ "",
        characteristic == "Household has gun and child under 18" ~ gc_ci,
        characteristic == "Mean children per exposed household" ~ m_ci,
        characteristic == "Estimated children exposed" ~ exp_ci,
        characteristic == "Estimated children exposed to loaded and unlocked firearms" ~ lu_ci
      )
    )
}

# Escape % for LaTeX (same pattern as your share table)
inputs_tbl_simple <- inputs_tbl_simple %>%
  mutate(across(starts_with("y"), ~ str_replace_all(., "%", "\\\\%")))

col_names_inputs <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(inputs_years)))
align_inputs     <- paste0("l", paste0(rep("c", length(inputs_years) * 2), collapse = ""))

inputs_note <-
  paste0(
    "Table reports every quantity used to compute the estimated number of US ",
    "children exposed to firearms in the home, by survey wave. Estimated children ",
    "exposed = total US households (ACS) x weighted prevalence of households ",
    "with a gun and a child under 18 (survey) x weighted mean children per ",
    "exposed household (survey). Estimated children exposed to loaded and unlocked ",
    "firearms additionally multiplies by the 2024 National Firearm Survey estimate ",
    "that 21.1\\% (95\\% CI, 18.3\\%–24.3\\%) of gun-owning households with children ",
    "had at least one firearm stored loaded and unlocked (Miller et al. 2026). ",
    "Total US household counts are ACS 1-year estimates, 2024 ACS ",
    "estimates were used for 2025."
  )

inputs_share_tbl <- inputs_tbl_simple %>%
  kable(
    format   = "latex",
    booktabs = TRUE,
    escape   = FALSE,
    caption  = "Inputs to the Estimated Number of US Children Exposed to Firearms in the Home, 2013-2025",
    label    = "ExposureInputs",
    col.names = col_names_inputs,
    linesep  = "",
    align    = align_inputs
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  ) %>%
  add_header_above(
    c(" " = 1, setNames(rep(2, length(inputs_years)), paste0(inputs_years)))
  ) %>%
  footnote(
    number = inputs_note,
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  inputs_share_tbl,
  file = here(path_ol, "tables/exposure-inputs.tex")
)
```

