# Subgroup analysis

This chapter explores time trends in differential household exposures by race and ethnicity, income and region.

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "ggthemes",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    )
```

```{r}
child %>%
  filter(gunchild18un == "Yes") %>%
  tabyl(racethnicity, year)
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)

race_levels <- c(
  "White, non-Hispanic",
  "Black, non-Hispanic",
  "Hispanic",
  "Asian, non-Hispanic",
  "Other, non-Hispanic",
  "2+ Races, non-Hispanic"
)

ds_race <- update(
  ds,
  exposed_race = factor(
    ifelse(gunchild18un == "Yes", as.character(racethnicity), "Not exposed"),
    levels = c(race_levels, "Not exposed")
  )
)
```


## Racial composition among all households irrespective of ownership


```{r}
comp_long <-
  svyby(
    ~exposed_race,
    ~year,
    ds_race,
    svymean,
    na.rm = TRUE
  ) %>%
  as_tibble() %>%
  select(year, starts_with("exposed_race")) %>%      
  pivot_longer(
    -year,
    names_to     = "race",
    values_to    = "share",
    names_prefix = "exposed_race"
  ) %>%
  filter(race != "Not exposed") %>%
  mutate(
    race  = factor(race, levels = race_levels),
    share = share * 100
  )
```


### Stacked bar over time

```{r}
exposed_race_plot <-
  comp_long |> 
  filter(year != 2015) |> 
  ggplot(aes(x = factor(year), y = share, fill = race)) +
  geom_col(width = 0.7) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.05))
  ) +
  scale_fill_economist() +
  labs(
    x    = NULL,
    y    = "Share of all respondents",
    fill = "Respondent race/ethnicity"
  ) +
  theme +
  theme(
    panel.grid.major.x = element_blank(),
    legend.position    = "bottom"
  )

exposed_race_plot
```


## Composition within gun-and-child households, with 95% CIs

```{R}
ds_exposed <- subset(ds, gunchild18un == "Yes")

raw <- svyby(
  ~racethnicity, ~year, ds_exposed, svymean,
  vartype = "ci", na.rm = TRUE) %>% as_tibble()

grab <- function(df, prefix, value_name) {
  df %>% select(year, starts_with(prefix)) %>%
    pivot_longer(-year, names_to = "race", values_to = value_name) %>%
    mutate(race = str_remove(race, fixed(prefix)))
}

comp_ci <-
  reduce(
    list(
      grab(raw, "racethnicity",      "share"),
      grab(raw, "ci_l.racethnicity", "lb"),
      grab(raw, "ci_u.racethnicity", "ub")
    ),
    left_join, by = c("year", "race")
  ) %>%
  mutate(
    race = factor(race, levels = race_levels),
    across(c(share, lb, ub), ~ .x * 100),
    across(c(lb, ub), ~ ifelse(is.nan(.x) | is.infinite(.x), NA, .x))
  ) %>%
  arrange(year, race)

## Unweighted exposed cell size per year (for reading the CIs)
exposed_n <- child %>% filter(gunchild18un == "Yes") %>% count(year, name = "n")
```


### Export table
```{r}
years <- sort(unique(comp_ci$year))
est_ci_order <- as.vector(rbind(paste0(years, "_est"), paste0(years, "_ci")))

comp_wide <- comp_ci %>%
  mutate(
    est = sprintf("%.1f", share),
    ci  = ifelse(is.na(lb), "—", sprintf("%.1f–%.1f", lb, ub))
  ) %>%
  select(race, year, est, ci) %>%
  pivot_wider(names_from = year, values_from = c(est, ci),
              names_glue = "{year}_{.value}") %>%
  arrange(race) %>%
  mutate(race = as.character(race)) %>%
  select(race, all_of(est_ci_order))

obs_wide <- exposed_n %>%
  mutate(est = format(n, big.mark = ","), ci = "") %>%
  select(year, est, ci) %>%
  pivot_wider(names_from = year, values_from = c(est, ci),
              names_glue = "{year}_{.value}") %>%
  mutate(race = "Observations (exposed households)") %>%
  select(race, all_of(est_ci_order))

comp_tbl <- bind_rows(obs_wide, comp_wide)

col_names <- c("Respondent race/ethnicity",
               rep(c("Est.", "95\\% CI"), length(years)))
align_str <- paste0("l", paste0(rep("c", length(years) * 2), collapse = ""))

comp_note <- paste0(
  "Cells are weighted percentages (Est.\\ column) of gun-and-child households ",
  "in each respondent race/ethnicity group, within survey wave; columns sum to ",
  "100. Denominator is households with both a firearm and a child under 18. ",
  "Race/ethnicity is that of the survey respondent, not the child. Estimates use ",
  "general population survey weights and account for survey design and non-response. ",
  "Racial and ethnic minority groups were not oversampled before 2021 (Black, ",
  "non-Hispanic and Hispanic first oversampled in 2021; Asian, non-Hispanic in 2023), ",
  "so pre-2021 minority segments derive from small unweighted cells and should be ",
  "read as directional. ``—'' denotes a cell with too few observations to estimate. "
)

comp_latex <- comp_tbl %>%
  kable(format = "latex", booktabs = TRUE, escape = FALSE, linesep = "",
        caption = "Racial/Ethnic Composition of US Gun-and-Child Households, 2013-2025",
        label = "ExposedRaceComp",
        col.names = col_names, align = align_str) %>%
  kable_styling(latex_options = c("hold_position", "scale_down"), font_size = 8) %>%
  row_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, setNames(rep(2, length(years)), paste0(years)))) %>%
  footnote(number = comp_note, threeparttable = TRUE,
           escape = FALSE, fixed_small_size = TRUE)

save_kable(comp_latex, file = here(path_ol, "tables", "exp-race-comp.tex"))
```

### Stacked bar chart

```{r}
race_cols <- c(
  "Black, non-Hispanic"    = "#1f4e5f",
  "Hispanic"               = "#3fa7d6",
  "Asian, non-Hispanic"    = "#e9a13b",
  "2+ Races, non-Hispanic" = "#7d5ba6",
  "Other, non-Hispanic"    = "#9bb0ba",
  "White, non-Hispanic"    = "#c9d6dd"
)
plot_order <- names(race_cols)   # bottom -> top

comp_plot_df <- comp_ci %>%
  mutate(race = factor(as.character(race), levels = plot_order))

exposed_comp_plot <-
  ggplot(comp_plot_df, aes(x = factor(year), y = share, fill = race)) +
  geom_col(width = 0.72, position = position_stack(reverse = TRUE)) +
  geom_text(
    aes(label = ifelse(share >= 3, sprintf("%.0f", share), "")),
    position = position_stack(vjust = 0.5, reverse = TRUE),
    size = 3, color = "black", fontface = "bold",
    family = "serif"
  ) +
  scale_fill_manual(values = race_cols, breaks = plot_order) +
  scale_y_continuous(labels = function(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  labs(
    x = NULL, y = "Share of firearm-owning households with children", fill = NULL
  ) +
  theme + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "bottom"
  ) +
  guides(fill = guide_legend(nrow = 1, reverse = FALSE))

exposed_comp_plot

ggsave(here(path_ol, "figures", "exposed-race-composition.png"),
       exposed_comp_plot, width = 10, height = 5.5, dpi = 300)
```


## Income


```{r}
comp_by <- function(design, var, levels_vec) {

  raw <- svyby(
    reformulate(var), ~year, design, svymean,
    vartype = "ci", na.rm = TRUE
  ) %>% as_tibble()

  reduce(
    list(
      grab(raw, var,                  "share"),
      grab(raw, paste0("ci_l.", var), "lb"),
      grab(raw, paste0("ci_u.", var), "ub")
    ),
    left_join, by = c("year", "race")
  ) %>%
    rename(group = race) %>%
    mutate(
      group = factor(group, levels = levels_vec),
      across(c(share, lb, ub), ~ .x * 100),
      across(c(lb, ub), ~ ifelse(is.nan(.x) | is.infinite(.x), NA, .x))
    ) %>%
    arrange(year, group)
}

export_comp_plot <- function(comp_df, cols, y_lab, file,
                             legend_rows = 1, label_cut = 3) {

  plot_order <- names(cols)

  p <- comp_df %>%
    mutate(group = factor(as.character(group), levels = plot_order)) %>%
    ggplot(aes(x = factor(year), y = share, fill = group)) +
    geom_col(width = 0.72, position = position_stack(reverse = TRUE)) +
    geom_text(
      aes(label = ifelse(share >= label_cut, sprintf("%.0f", share), "")),
      position = position_stack(vjust = 0.5, reverse = TRUE),
      size = 3, color = "black", fontface = "bold", family = "serif"
    ) +
    scale_fill_manual(values = cols, breaks = plot_order) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       expand = expansion(mult = c(0, 0.02))) +
    labs(x = NULL, y = y_lab, fill = NULL) +
    theme +
    theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor   = element_blank(),
      legend.position    = "bottom"
    ) +
    guides(fill = guide_legend(nrow = legend_rows, reverse = FALSE))

  ggsave(here(path_ol, "figures", file), p, width = 10, height = 5.5, dpi = 300)

  p
}

```

```{r}
income_levels <- levels(droplevels(as_factor(child$income4)))

comp_income <- comp_by(ds_exposed, "income4", income_levels)

income_cols <- setNames(
  c("#1f4e5f", "#3fa7d6", "#e9a13b", "#c9d6dd")[seq_along(income_levels)],
  income_levels
)

income_comp_plot <- export_comp_plot(
  comp_df = comp_income,
  cols    = income_cols,
  y_lab   = "Share of firearm-owning households with children",
  file    = "exposed-income-composition.png"
)
```


## Region

```{r}
region_levels <- levels(droplevels(as_factor(child$region4)))

comp_region <- comp_by(ds_exposed, "region4", region_levels)


region_cols <- setNames(
  c("#1f4e5f", "#3fa7d6", "#e9a13b", "#7d5ba6")[seq_along(region_levels)],
  region_levels
)

region_comp_plot <- export_comp_plot(
  comp_df = comp_region,
  cols    = region_cols,
  y_lab   = "Share of firearm-owning households with children",
  file    = "exposed-region-composition.png"
)
```

