# Demographic Summary of Households with Children Exposed to Guns.

This chapter creates a descriptive table that shows the demographics of households with 
children exposed to guns in the home. 

**Input**

-   DemographyGunOwners/data/clean/gs-child.rds

**Output**

-   DemographyGunOwners/output

**Last ran**

-   `r format(Sys.Date(), format = "%A, %B %d, %Y")`

```{r, message = FALSE, warning = FALSE}
packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "gtsummary",
    "janitor",
    "dplyr",
    "here",
    "vtable",
    "tidycensus"
  )

pacman::p_load(packages, character.only = TRUE)

#api key for the Census Data
KEY = "5cc38485c4b3852e79d4c13a6f8f5095d2f89db8"
```

## Load data

```{r}
child <- 
  read_rds(
    here(
      path_od,
      "data",
      "clean", 
      "gs-child.rds"
      )
    )
```

## Survey designs

```{r}
ds <- svydesign(
  ~1,
  data    = child,
  strata  = ~year,
  weights = ~weight
)
```


## Weighted share of exposed households with CIs

```{r}
children_exposure <- 
  svyby(
    ~gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>% 
  transmute(
    year,
    prev_pct = gunchild18un_yes,
    prev_se = se_gunchild18un_yes,
    prev_lb = ci_l_gunchild18un_yes,
    prev_ub = ci_u_gunchild18un_yes
  )
```

## Average number of children in exposed households

```{r}
avg_child <-
  svyby(
    ~hh18un,
    ~year,
    subset(ds, gunchild18un == "Yes"),
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names() %>%
  mutate(across(-year, ~ .x)) %>%
  transmute(
    year,
    avg_child = hh18un,
    avg_child_se = se,
    avg_child_lb = ci_l,
    avg_child_ub = ci_u
  ) 
```

## Get population estimates from the ACS

I applied for an API key using this [website](https://api.census.gov/data/key_signup.html) and my jdada3@jh.edu email.

2025 estimates are not available, so we use the 2024 estimates for that year.

- [B11001_001](https://data.census.gov/table/ACSDT1Y2022.B11001)


```{r, message = F, warning = F}
census_estimates <- 
  map_df(
    c(seq(2013, 2023, 2), 2024),
    ~ get_acs(
      geography = "us",
      variables = c(total_hh = "B11001_001"),  # Total households
      year = .x,
      survey = "acs1",
      key = KEY
    ) %>%
    mutate(year = .x) %>%
    select(year, estimate)
  ) %>%
  rename(totalhh = estimate) %>%
  select(year, totalhh) %>% 
  mutate(year = ifelse(year == 2024, 2025, year))
```


## Combine Gun Survey and ACS estimates to compute the 95% CI of the number of children expose

We use the conservative multiplication rule for combining uncertainty from 2 sources
by multiplying the upper/lower bound estimates from each of the parameters included in the final calculation/

```{r}
exposure_estimates <-
  census_estimates %>%
  left_join(children_exposure, by = "year") %>%
  left_join(avg_child, by = "year") %>%
  mutate(
    # Point estimates
    hh_gun = totalhh * prev_pct,
    child_est = hh_gun * avg_child,
    
    # 95% CI
    child_lb = totalhh * prev_lb * avg_child_lb,
    child_ub = totalhh * prev_ub * avg_child_ub
  ) %>%
  select(year, child_est, child_lb, child_ub)
```

```{r}
exposure_estimates %>% 
  mutate(
    across(-year, ~ sjmisc::big_mark(.x))
  )
```

## Child variables

```{R}
childvars <-
  c(
    "gunhousehold",
    "child18un",
    "gunchild18un"
  )
```

## Percentage of US households with guns and children in the home

```{r}
note <-
  paste0(
    "Data represent weighted prevalence estimates of US households. ",
    "All estimates reflect the share of households with: (1) a gun in the home, ",
    "(2) a child under 18 in the household, and (3) both a gun and a child under 18 ",
    "in the home. Weights are based on general population survey weights and account ",
    "for survey design and non-response. The 2015 survey measured ",
    "personal gun ownership (respondent-owned guns), while 2013, 2017, 2019, 2021, 2023, ",
    "and 2025 measured any guns in the home regardless of ownership. ",
    "Estimated US children exposed to guns in the home is calculated by multiplying: ",
    "total US households (from Census Bureau American Community Survey), ",
    "prevalence of households with guns and children (survey estimate), and ",
    "average number of children per exposed household (survey estimate). ",
    "The 95 percent confidence interval for estimated children exposed uses the conservative ",
    "multiplication rule."
  )
```

## Summary table

```{r}
# Calculate all proportion estimates with CIs
proportion_estimates <- 
  svyby(
    ~gunhousehold + child18un + gunchild18un,
    ~year,
    ds,
    svymean,
    vartype = c("se", "ci"),
    level = 0.95
  ) %>% 
  clean_names()

# Add observations
obs_counts <- tibble(
  year = as.numeric(names(table(child$year))),
  obs = as.numeric(table(child$year))
)

# Get unique years
years <- sort(unique(proportion_estimates$year))

# Create table structure
share_tbl_simple <- tibble(
  characteristic = c(
    "Observations", 
    "Household has gun", 
    "Household has child under 18", 
    "Household has gun and child under 18",
    "Estimated children exposed"
  )
)

# Add columns for each year
for (yr in years) {
  yr_data <- proportion_estimates %>% filter(year == yr)
  yr_obs <- obs_counts %>% filter(year == yr) %>% pull(obs) %>% format(big.mark = ",")
  
  # Exposure estimates for this year
  yr_exposure <- exposure_estimates %>% filter(year == yr)
  exposure_est <- paste0(sjmisc::big_mark(round(yr_exposure$child_est / 1e6, 1)), " M")
  exposure_ci <- paste0(sjmisc::big_mark(round(yr_exposure$child_lb / 1e6, 1)), "–", 
                        sjmisc::big_mark(round(yr_exposure$child_ub / 1e6, 1)), " M")
  
  # Gun
  gun_est <- paste0(round(yr_data$gunhousehold_yes * 100, 1), "%")
  gun_ci <- paste0(round(yr_data$ci_l_gunhousehold_yes * 100, 1), "–", 
                   round(yr_data$ci_u_gunhousehold_yes * 100, 1), "%")
  
  # Child
  child_est <- paste0(round(yr_data$child18un_yes * 100, 1), "%")
  child_ci <- paste0(round(yr_data$ci_l_child18un_yes * 100, 1), "–", 
                     round(yr_data$ci_u_child18un_yes * 100, 1), "%")
  
  # Gun and child
  gunchild_est <- paste0(round(yr_data$gunchild18un_yes * 100, 1), "%")
  gunchild_ci <- paste0(round(yr_data$ci_l_gunchild18un_yes * 100, 1), "–", 
                        round(yr_data$ci_u_gunchild18un_yes * 100, 1), "%")
  
  # Add columns (Est, CI pairs for this year)
  share_tbl_simple <- share_tbl_simple %>%
    mutate(
      !!paste0("y", yr, "_est") := case_when(
        characteristic == "Observations" ~ yr_obs,
        characteristic == "Household has gun" ~ gun_est,
        characteristic == "Household has child under 18" ~ child_est,
        characteristic == "Household has gun and child under 18" ~ gunchild_est,
        characteristic == "Estimated children exposed" ~ exposure_est
      ),
      !!paste0("y", yr, "_ci") := case_when(
        characteristic == "Observations" ~ "",
        characteristic == "Household has gun" ~ gun_ci,
        characteristic == "Household has child under 18" ~ child_ci,
        characteristic == "Household has gun and child under 18" ~ gunchild_ci,
        characteristic == "Estimated children exposed" ~ exposure_ci
      )
    )
}

# Escape % for LaTeX
share_tbl_simple <- share_tbl_simple %>%
  mutate(
    across(
      starts_with("y"),
      ~ str_replace_all(., "%", "\\\\%")
    )
  )

# Create column names for header (without year labels, those go in add_header_above)
col_names <- c("Characteristic", rep(c("Est.", "95\\% CI"), length(years)))
align_string <- paste0("l", paste0(rep("c", length(years) * 2), collapse = ""))

share_tbl <- share_tbl_simple %>%
  kable(
    format = "latex",
    booktabs = TRUE,
    escape = FALSE,
    caption = "Share of US Households with Guns and Children Under 18, 2013-2025",
    label = "ShareYear",
    col.names = col_names,
    linesep = "",
    align = align_string
  ) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down"),
    font_size = 8
  ) %>%
  row_spec(
    row = 1,
    bold = T
  ) %>% 
  add_header_above(
    c(" " = 1, setNames(rep(2, length(years)), paste0(years)))
  ) %>%
  footnote(
    number = note,
    threeparttable = TRUE,
    escape = FALSE,
    fixed_small_size = TRUE
  )

save_kable(
  share_tbl,
  file = here(path_ol, "tables/yearly-share.tex")
)
```
