# Summary Statistics

Survey weights were normalized to the sample size; therefore, population counts were obtained by multiplying weighted proportions by external Census population totals.

**Input**

-   DemographyGunOwners/data/clean/gunsurvey/...rds

**Output**

-   DemographyGunOwners/output

**Last ran** - 12/14/2025

```{r,  message = FALSE, warning = FALSE}

packages <-
  c(
    "tidyverse",
    "purrr",
    "haven",
    "survey",
    "tidyr",
    "janitor",
    "dplyr",
    "here"
  )

pacman::p_load(
  packages,
  character.only = TRUE
)
```

## Import data

```{r}
raw <-
  read_rds(
     here(
      path_od, 
      "data",
      "raw",
      "gunsurvey",
      "2017.rds"
      )
  )
```


## Set up survey design

This uses the package `survey`.

```{r}
des <- svydesign(
  ids = ~1,                 # no clustering
  weights = ~ weight1,   # your post-strat weight
  data = raw
) 
```

## NRA membership

```{r}
raw %>% tabyl(q29)
```

```{r}
svymean(~ q29, des)
```


```{r}
p <- svymean(~ q29, des)[1]   # weighted proportion
N <- 258e6                         # US adult population
owners <- p * N
```

### By gender

```{r}
svyby(
  ~q29,
  ~gender,
  des,
  svymean
)
```
